---
title: "STAT 544 - SWITRS Crash Severity Notebook (Rebuilt for Correctly Merged Data)"
output:
  html_notebook: default
  pdf_document: default
  html_document: default
---

# Package Chunk
```{r setup, message=FALSE, warning=FALSE}
# What: Global setup (packages, options, theme).
# Why: Ensures reproducibility and consistent formatting across the notebook.

options(stringsAsFactors = FALSE)
set.seed(544)

required_pkgs <- c(
  "dplyr", "tidyr", "stringr", "ggplot2", "scales", "purrr",
  "readr", "forcats", "corrplot", "broom", "rlang"
)

installed <- rownames(installed.packages())
to_install <- setdiff(required_pkgs, installed)
if (length(to_install) > 0) install.packages(to_install)

library(rlang)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(scales)
library(purrr)
library(readr)
library(forcats)
library(corrplot)
library(broom)

theme_set(theme_minimal(base_size = 12))
```

# Load data (correctly merged dataset)
```{r}
# What: Load crash-grain merged SWITRS dataset (1 row per CASE_ID).
# Why: Parties/Victims have already been aggregated to crash-level in the merge notebook.

na_tokens <- c("NA", "N/A", "NULL", "null")

switrs_raw <- read.csv("merged_fixed_switrs.csv", na.strings = na_tokens, stringsAsFactors = FALSE)

switrs <- switrs_raw
rm(switrs_raw)

dim(switrs)
glimpse(switrs)
```

# Basic sanity checks
```{r}
# What: Confirm crash-grain (1 row per crash).
# Why: The merge notebook output should already be 1 row per CASE_ID.

stopifnot("CASE_ID" %in% names(switrs))
stopifnot(n_distinct(switrs$CASE_ID) == nrow(switrs))

# Optional: show a few IDs to verify
head(switrs$CASE_ID, 10)
```

# Standardize "not stated" / placeholder missingness (do not lump into OTHER)
```{r}
# What: Normalize placeholder encodings with SWITRS-aware rules.
# Why: Standardizes "-" and blank encodings across the crash-grain dataset.
#      Preserves SWITRS semantics for "Y or blank" flags (blank/NA means No).

trim_chr <- function(x) stringr::str_trim(as.character(x))

# Global helper used later (EDA labels, hypothesis tests, plots)
# NOTE: we default blank_to_notstated=TRUE, but in the cleaning step below
#       we set blank_to_notstated=FALSE so Y-or-blank flags can still become "N".
normalize_not_stated <- function(x, not_stated_label = "Not Stated", blank_to_notstated = TRUE) {
  x <- trim_chr(x)

  # dashes and explicit "not stated"
  x[x %in% c("-", " -", "- ", " - ")] <- not_stated_label
  x[toupper(x) %in% c("NOT STATED", "NOTSTATED")] <- not_stated_label

  # optionally treat blank as Not Stated
  if (isTRUE(blank_to_notstated)) {
    x[x == ""] <- not_stated_label
  }
  x
}

# SWITRS "Y or blank" flags: blank (and often NA after read.csv) means NO
y_or_blank_flags <- intersect(
  c("PEDESTRIAN_ACCIDENT","BICYCLE_ACCIDENT","MOTORCYCLE_ACCIDENT","TRUCK_ACCIDENT",
    "NOT_PRIVATE_PROPERTY","ALCOHOL_INVOLVED"),
  names(switrs)
)

# 1) Trim + normalize dash-style placeholders on all character fields
#    IMPORTANT: do NOT convert blank->Not Stated here; we handle blanks in steps 2/3
switrs_clean <- switrs %>%
  mutate(across(where(is.character), ~ normalize_not_stated(.x, blank_to_notstated = FALSE)))

# 2) For most character fields: "" => "Not Stated" (do NOT touch y-or-blank flags here)
char_cols <- names(switrs_clean)[sapply(switrs_clean, is.character)]
other_char_cols <- setdiff(char_cols, y_or_blank_flags)

switrs_clean <- switrs_clean %>%
  mutate(across(all_of(other_char_cols), ~ {
    x <- .x
    x[x == ""] <- "Not Stated"
    x
  }))

# 3) For y-or-blank flags: NA/"" => "N" (blank means No); keep "Not Stated" if it exists
switrs_clean <- switrs_clean %>%
  mutate(across(all_of(y_or_blank_flags), ~ {
    x <- trim_chr(.x)
    x[is.na(x) | x == ""] <- "N"
    x[x == "Y"] <- "Y"
    x
  }))
```

# Convert "UNKNOWN" road names to NA (special case)
```{r}
# What: Convert literal UNKNOWN / Not Stated for roadway name fields to NA.
# Why: For roadway names, these behave like missingness (not meaningful categories).

road_name_vars <- intersect(c("PRIMARY_RD", "SECONDARY_RD"), names(switrs_clean))

switrs_clean <- switrs_clean %>%
  mutate(across(all_of(road_name_vars), ~ {
    x <- as.character(.x)
    x <- str_trim(x)
    x[toupper(x) == "UNKNOWN"] <- NA_character_
    x[x == "Not Stated"] <- NA_character_
    x
  }))
```

# Convert Y/N and Y-only flags carefully
```{r}
# What: Standardize common flag formats into consistent factors.
# Why: SWITRS includes Y/N/blank-not-stated fields AND Y-or-blank fields (blank = No).

# Convert the "Y or blank" crash flags into factor(N, Y, Not Stated)
y_or_blank_flags <- intersect(
  c("PEDESTRIAN_ACCIDENT","BICYCLE_ACCIDENT","MOTORCYCLE_ACCIDENT","TRUCK_ACCIDENT",
    "NOT_PRIVATE_PROPERTY","ALCOHOL_INVOLVED"),
  names(switrs_clean)
)

switrs_clean <- switrs_clean %>%
  mutate(across(all_of(y_or_blank_flags), ~ {
    x <- str_trim(as.character(.x))
    # SWITRS: blank means No
    x[is.na(x) | x == ""] <- "N"
    # Dash was already normalized to "Not Stated" earlier
    x[x == "Y"] <- "Y"
    factor(x, levels = c("N","Y","Not Stated"))
  }))

# Standard Y/N/Not Stated fields (e.g., INTERSECTION, STATE_HWY_IND, TOW_AWAY, AT_FAULT, etc.)
crash_yn_whitelist <- intersect(
  c("INTERSECTION","STATE_HWY_IND","TOW_AWAY","AT_FAULT"),
  names(switrs_clean)
)

switrs_clean[crash_yn_whitelist] <- lapply(switrs_clean[crash_yn_whitelist], function(x) {
  x <- str_trim(as.character(x))
  x[x %in% c("", NA)] <- "Not Stated"
  factor(x, levels = c("N","Y","Not Stated"))
})

# HIT_AND_RUN is F/M/N (+ Not Stated), not Y/N.
if ("HIT_AND_RUN" %in% names(switrs_clean)) {
  switrs_clean <- switrs_clean %>%
    mutate(HIT_AND_RUN = factor(
      ifelse(is.na(HIT_AND_RUN), "Not Stated", HIT_AND_RUN),
      levels = c("N","M","F","Not Stated")
    ))
}

# PCF_VIOL_CATEGORY is Char(2): keep leading zeros (01, 02, ..., 24, 00).
if ("PCF_VIOL_CATEGORY" %in% names(switrs_clean)) {
  switrs_clean <- switrs_clean %>%
    mutate(
      PCF_VIOL_CATEGORY = as.character(PCF_VIOL_CATEGORY),
      PCF_VIOL_CATEGORY = str_trim(PCF_VIOL_CATEGORY),
      PCF_VIOL_CATEGORY = ifelse(PCF_VIOL_CATEGORY == "Not Stated", "Not Stated",
                                 ifelse(PCF_VIOL_CATEGORY %in% c("", NA), NA_character_, PCF_VIOL_CATEGORY)),
      PCF_VIOL_CATEGORY = ifelse(
        !is.na(PCF_VIOL_CATEGORY) & PCF_VIOL_CATEGORY != "Not Stated" & str_detect(PCF_VIOL_CATEGORY, "^[0-9]+$"),
        str_pad(PCF_VIOL_CATEGORY, width = 2, pad = "0"),
        PCF_VIOL_CATEGORY
      )
    )
}
```

# Force known codebook-coded categoricals to be categorical
```{r}
# What: Coerce known SWITRS coded fields to factors before EDA, plots, and hypothesis tests.
# Why: Prevents code-like variables from being misread as numeric and sent to t-tests.

# DAY_OF_WEEK is coded categorical, not numeric
if ("DAY_OF_WEEK" %in% names(switrs_clean)) {
  switrs_clean <- switrs_clean %>%
    dplyr::mutate(
      DAY_OF_WEEK = dplyr::case_when(
        as.character(DAY_OF_WEEK) %in% c("1","2","3","4","5","6","7") ~ as.character(DAY_OF_WEEK),
        TRUE ~ "Not Stated"
      ),
      DAY_OF_WEEK = factor(
        DAY_OF_WEEK,
        levels = c("1","2","3","4","5","6","7","Not Stated"),
        labels = c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday","Not Stated")
      )
    )
}

# These are coded categorical fields; keep as factors, not numeric
coded_factor_vars <- intersect(
  c(
    "WEATHER_1",
    "WEATHER_2",
    "LIGHTING",
    "ROAD_SURFACE",
    "TYPE_OF_COLLISION",
    "PRIMARY_COLL_FACTOR",
    "PCF_VIOL_CATEGORY",
    "PED_ACTION",
    "CONTROL_DEVICE",
    "CHP_SHIFT",
    "BEAT_TYPE",
    "CHP_BEAT_TYPE",
    "CHP_BEAT_CLASS",
    "SPECIAL_COND"
  ),
  names(switrs_clean)
)

switrs_clean <- switrs_clean %>%
  dplyr::mutate(
    dplyr::across(all_of(coded_factor_vars), ~ {
      x <- trimws(as.character(.x))
      x[x %in% c("", NA)] <- "Not Stated"
      factor(x)
    })
  )

# Binary engineered / aggregate variables that should behave like categorical in plots/tests
binary_factor_vars <- intersect(
  c(
    "INTERSECTION",
    "STATE_HWY_IND",
    "TOW_AWAY",
    "AT_FAULT",
    "PEDESTRIAN_ACCIDENT",
    "BICYCLE_ACCIDENT",
    "MOTORCYCLE_ACCIDENT",
    "TRUCK_ACCIDENT",
    "NOT_PRIVATE_PROPERTY",
    "ALCOHOL_INVOLVED"
  ),
  names(switrs_clean)
)

switrs_clean <- switrs_clean %>%
  dplyr::mutate(
    dplyr::across(all_of(binary_factor_vars), ~ {
      x <- trimws(as.character(.x))
      x[x %in% c("", NA)] <- "Not Stated"
      factor(x)
    })
  )

cat("Early categorical typing complete.\n")
```

# quick audit
```{r}
check_vars <- intersect(c(
  "DAY_OF_WEEK", "WEATHER_1", "LIGHTING", "ROAD_SURFACE",
  "PRIMARY_COLL_FACTOR", "TYPE_OF_COLLISION", "PCF_VIOL_CATEGORY"
), names(switrs_clean))

sapply(switrs_clean[check_vars], class)
```

# Build the crash-level analysis table
```{r}
# What: Finalize crash-level analysis table.
# Why: Data are already crash-grain; after cleaning we proceed with switrs_reduced.
#      Keep the original switrs object in memory for report-only structure/independence checks.

switrs_reduced <- switrs_clean
rm(switrs_clean)
```

# validate Road_Surface
```{r}
# What: Validate ROAD_SURFACE against codebook-valid codes.
# Why: Codebook defines ROAD_SURFACE as A-D (plus Not Stated). Anything else is out-of-spec.

if ("ROAD_SURFACE" %in% names(switrs_reduced)) {

  valid_road_surface <- c("A","B","C","D","Not Stated")

  switrs_reduced <- switrs_reduced %>%
    mutate(
      ROAD_SURFACE = normalize_not_stated(ROAD_SURFACE, blank_to_notstated = TRUE),
      ROAD_SURFACE = ifelse(is.na(ROAD_SURFACE), "Not Stated", ROAD_SURFACE),

      ROAD_SURFACE_invalid = ifelse(ROAD_SURFACE %in% valid_road_surface, 0L, 1L),
      ROAD_SURFACE_raw = ifelse(ROAD_SURFACE_invalid == 1L, ROAD_SURFACE, NA_character_),

      ROAD_SURFACE = ifelse(ROAD_SURFACE_invalid == 1L, "Not Stated", ROAD_SURFACE)
    )

  cat("Out-of-spec ROAD_SURFACE codes (flagged + recoded to Not Stated):\n")
  print(table(switrs_reduced$ROAD_SURFACE_raw, useNA = "ifany"))
}
```

# Cleanup: remove ROAD_SURFACE validation helper columns (data-quality flags)
```{r}
switrs_reduced <- switrs_reduced %>%
  dplyr::select(-dplyr::any_of(c("ROAD_SURFACE_invalid","ROAD_SURFACE_raw")))
```

# Missingness Diagnostics (before dropping columns)
```{r}
# What: Missingness diagnostics at the correct crash-level grain.
# Why: NA% on merged rows is misleading (structural NA from join).

missing_na_pct <- colMeans(is.na(switrs_reduced)) * 100

notstated_pct <- sapply(switrs_reduced, function(x) {
  if (!(is.character(x) || is.factor(x))) return(NA_real_)
  mean(as.character(x) == "Not Stated", na.rm = TRUE) * 100
})

missing_tbl <- tibble::tibble(
  variable = names(switrs_reduced),
  pct_na = as.numeric(missing_na_pct),
  pct_notstated = as.numeric(notstated_pct)
) %>%
  arrange(desc(pct_na), desc(pct_notstated))

head(missing_tbl, 30)
```

# drop columns with too much effective missingness
```{r}
# What: Drop columns exceeding missingness threshold using NA for numeric variables
#       and NA/"Not Stated" for categorical variables.
# Why: Many SWITRS categoricals were standardized to "Not Stated", so using only true NA
#      understates how incomplete those variables really are.

high_miss_threshold <- 30

effective_missing_pct <- sapply(switrs_reduced, function(x) {
  if (is.character(x) || is.factor(x)) {
    mean(is.na(x) | as.character(x) == "Not Stated") * 100
  } else {
    mean(is.na(x)) * 100
  }
})

high_miss_vars <- names(effective_missing_pct)[effective_missing_pct > high_miss_threshold]

protect_vars <- intersect(
  c("COLLISION_SEVERITY", "NUMBER_KILLED", "NUMBER_INJURED"),
  names(switrs_reduced)
)

high_miss_vars <- setdiff(high_miss_vars, protect_vars)

cat("Variables above effective missingness threshold:\n")
print(sort(effective_missing_pct[high_miss_vars], decreasing = TRUE))

switrs_reduced <- switrs_reduced %>%
  dplyr::select(-dplyr::all_of(high_miss_vars))

cat("\nRemaining dimensions after dropping high-missingness variables:\n")
print(dim(switrs_reduced))
```

# drop deprecated / non-analytic variables
```{r}
# What: Remove variables deprecated in the codebook or not analytically useful.

drop_vars <- intersect(
  c("CHP_SHIFT","BEAT_TYPE","CHP_BEAT_CLASS","CALTRANS_COUNTY","CALTRANS_DISTRICT","STATE_ROUTE",
    "ROUTE_SUFFIX","POSTMILE_PREFIX","POSTMILE","LOCATION_TYPE","RAMP_INTERSECTION","CHP_ROAD_TYPE"),
  names(switrs_reduced)
)

switrs_reduced <- switrs_reduced %>% select(-all_of(drop_vars))
dim(switrs_reduced)
```

# Preserve raw-coded severity and create SevereBinary outcome
```{r}
# Preserve raw-coded severity and create outcomes (KSI + fatal-only + deaths)
# What: Create SevereBinary (KSI proxy) from COLLISION_SEVERITY, plus FatalBinary/Deaths for impact reporting.
# Why: SevereBinary (codes 1/2) is the main actionable “prevent worst outcomes” target.
#      FatalBinary/Deaths keep the story explicitly “save lives” without leaking into predictors.

if ("COLLISION_SEVERITY" %in% names(switrs_reduced)) {
  switrs_reduced <- switrs_reduced %>%
    mutate(
      COLLISION_SEVERITY_CODE = str_trim(as.character(COLLISION_SEVERITY)),
      COLLISION_SEVERITY_CODE = ifelse(COLLISION_SEVERITY_CODE %in% c("", "Not Stated"), NA_character_, COLLISION_SEVERITY_CODE),

      SevereBinary = case_when(
        COLLISION_SEVERITY_CODE %in% c("1","2") ~ "Severe/Fatal",
        COLLISION_SEVERITY_CODE %in% c("0","3","4") ~ "Other",
        TRUE ~ NA_character_
      ),
      SevereBinary = factor(SevereBinary, levels = c("Other","Severe/Fatal"))
    )
} else {
  stop("COLLISION_SEVERITY not found; cannot create SevereBinary/KSI.")
}

if ("NUMBER_KILLED" %in% names(switrs_reduced)) {
  switrs_reduced <- switrs_reduced %>%
    mutate(
      NUMBER_KILLED_NUM = suppressWarnings(as.integer(NUMBER_KILLED)),
      FatalBinary = case_when(
        is.na(NUMBER_KILLED_NUM) ~ NA_integer_,
        NUMBER_KILLED_NUM > 0 ~ 1L,
        TRUE ~ 0L
      ),
      Deaths = NUMBER_KILLED_NUM
    )
} else {
  switrs_reduced$FatalBinary <- NA_integer_
  switrs_reduced$Deaths <- NA_integer_
}

cat("\nOutcome counts:\n")
print(table(switrs_reduced$SevereBinary, useNA = "ifany"))

if (all(c("FatalBinary","COLLISION_SEVERITY_CODE") %in% names(switrs_reduced))) {
  sev_fatal_code <- switrs_reduced$COLLISION_SEVERITY_CODE == "1"
  cat("\nAudit table: COLLISION_SEVERITY==1 (fatal code) vs FatalBinary(NUMBER_KILLED>0)\n")
  print(table(sev_fatal_code, switrs_reduced$FatalBinary, useNA = "ifany"))

  mismatch <- sum(sev_fatal_code & switrs_reduced$FatalBinary == 0, na.rm = TRUE) +
              sum(!sev_fatal_code & switrs_reduced$FatalBinary == 1, na.rm = TRUE)
  cat("Fatal-code vs killed-count mismatches (non-NA):", mismatch, "\n")
}
```

# Leak-proofing: ban outcome/harm variables from ever being used as predictors
```{r}
leakage_vars <- c(
  "COLLISION_SEVERITY", "COLLISION_SEVERITY_CODE", "COLLISION_SEVERITY_LAB",
  "SevereBinary",
  "NUMBER_KILLED", "NUMBER_KILLED_NUM", "FatalBinary", "Deaths",
  "NUMBER_INJURED",
  "COUNT_SEVERE_INJ", "COUNT_VISIBLE_INJ",
  "COUNT_PED_KILLED", "COUNT_PED_INJURED",
  "COUNT_BICYCLIST_KILLED", "COUNT_BICYCLIST_INJURED",
  "COUNT_MC_KILLED", "COUNT_MC_INJURED",
  "n_victims_killed", "any_victim_ksi"
)

leakage_vars <- intersect(leakage_vars, names(switrs_reduced))
cat("Leakage vars detected in data (will be banned from predictors):\n")
print(leakage_vars)
```

# Location hotspots
```{r}
# WHERE: Hotspot ranking (severe/fatal rate) using available location fields

has_latlon <- all(c("LATITUDE", "LONGITUDE") %in% names(switrs_reduced))
has_roads  <- all(c("PRIMARY_RD", "SECONDARY_RD") %in% names(switrs_reduced))
has_city   <- "CITY" %in% names(switrs_reduced)

switrs_where <- switrs_reduced %>%
  dplyr::filter(!is.na(SevereBinary)) %>%
  dplyr::mutate(
    lat_bin = if (has_latlon) round(as.numeric(LATITUDE), 2) else NA_real_,
    lon_bin = if (has_latlon) round(as.numeric(LONGITUDE), 2) else NA_real_,
    latlon_key = if (has_latlon) paste0(lat_bin, ", ", lon_bin) else NA_character_,

    rd1 = if (has_roads) stringr::str_squish(toupper(as.character(PRIMARY_RD))) else NA_character_,
    rd2 = if (has_roads) stringr::str_squish(toupper(as.character(SECONDARY_RD))) else NA_character_,

    rd1 = dplyr::if_else(is.na(rd1) | rd1 %in% c("", "NOT STATED", "UNKNOWN"), NA_character_, rd1),
    rd2 = dplyr::if_else(is.na(rd2) | rd2 %in% c("", "NOT STATED", "UNKNOWN"), NA_character_, rd2),

    rd_lo = dplyr::if_else(!is.na(rd1) & !is.na(rd2), pmin(rd1, rd2), NA_character_),
    rd_hi = dplyr::if_else(!is.na(rd1) & !is.na(rd2), pmax(rd1, rd2), NA_character_),
    inter_key = dplyr::if_else(!is.na(rd_lo) & !is.na(rd_hi), paste0(rd_lo, " & ", rd_hi), NA_character_),

    city_key = if (has_city) {
      x <- stringr::str_squish(toupper(as.character(CITY)))
      dplyr::if_else(is.na(x) | x %in% c("", "NOT STATED", "UNKNOWN"), NA_character_, x)
    } else NA_character_,

    location_key = dplyr::case_when(
      !is.na(inter_key) ~ inter_key,
      is.na(inter_key) & !is.na(latlon_key) ~ latlon_key,
      is.na(inter_key) & is.na(latlon_key) & !is.na(city_key) ~ city_key,
      TRUE ~ NA_character_
    ),
    location_type = dplyr::case_when(
      !is.na(inter_key) ~ "Intersection (PRIMARY_RD & SECONDARY_RD)",
      is.na(inter_key) & !is.na(latlon_key) ~ "Lat/Lon bin (~0.01°)",
      is.na(inter_key) & is.na(latlon_key) & !is.na(city_key) ~ "City",
      TRUE ~ "Missing"
    )
  ) %>%
  dplyr::filter(!is.na(location_key))

min_crashes <- 50

hotspot_tbl <- switrs_where %>%
  dplyr::group_by(location_type, location_key) %>%
  dplyr::summarise(
    n_crashes = dplyr::n(),
    severe_n = sum(SevereBinary == "Severe/Fatal", na.rm = TRUE),
    severe_per_100 = 100 * severe_n / n_crashes,
    .groups = "drop"
  ) %>%
  dplyr::filter(n_crashes >= min_crashes) %>%
  dplyr::arrange(dplyr::desc(severe_per_100), dplyr::desc(n_crashes))

min_crashes <- 10
min_severe  <- 1

hotspot_tbl2 <- switrs_where %>%
  group_by(location_type, location_key) %>%
  summarise(
    n_crashes = n(),
    severe_n = sum(SevereBinary == "Severe/Fatal", na.rm = TRUE),
    severe_per_100 = 100 * severe_n / n_crashes,
    fatal_crash_n = sum(FatalBinary == 1, na.rm = TRUE),
    deaths = sum(Deaths, na.rm = TRUE),
    deaths_per_100 = 100 * deaths / n_crashes,
    .groups = "drop"
  )

hotspots_by_rate <- hotspot_tbl2 %>%
  filter(n_crashes >= min_crashes, severe_n >= min_severe) %>%
  arrange(desc(severe_per_100), desc(n_crashes))

cat("\nTop 20 hotspots by Severe/Fatal RATE (min_crashes =", min_crashes,
    ", min_severe =", min_severe, "): \n")
print(head(hotspots_by_rate, 20))

hotspots_by_count <- hotspot_tbl2 %>%
  filter(n_crashes >= min_crashes) %>%
  arrange(desc(severe_n), desc(n_crashes))

hotspots_by_deaths <- hotspot_tbl2 %>%
  filter(n_crashes >= min_crashes) %>%
  arrange(desc(deaths), desc(n_crashes))

cat("\nTop 20 hotspots by DEATHS (min_crashes =", min_crashes, "): \n")
print(head(hotspots_by_deaths, 20))

cat("\nTop 20 hotspots by Severe/Fatal COUNT (min_crashes =", min_crashes, "): \n")
print(head(hotspots_by_count, 20))
```

# Sanity Check
```{r}
dplyr::count(switrs_where, location_type)
summary(nchar(switrs_where$location_key))
```

# Derive core time features(+ engineered risk flags & interactions)
```{r}
# What: Derive standardized time features and engineer proposal-aligned risk flags + interaction terms.

safe_parse_yyyymmdd <- function(x) {
  x <- as.character(x)
  x <- str_trim(x)
  x[x %in% c("", "Not Stated")] <- NA_character_
  suppressWarnings(as.Date(x, format = "%Y%m%d"))
}

safe_parse_hhmm <- function(x) {
  x <- as.character(x)
  x <- str_trim(x)
  x[x %in% c("", "Not Stated")] <- NA_character_
  x_num <- suppressWarnings(as.integer(x))
  x_num[x_num >= 2500] <- NA_integer_
  hr <- x_num %/% 100
  hr[hr < 0 | hr > 23] <- NA_integer_
  hr
}

if ("ACCIDENT_YEAR" %in% names(switrs_reduced)) {
  switrs_reduced <- switrs_reduced %>% mutate(year = as.integer(ACCIDENT_YEAR))
}

if ("COLLISION_TIME" %in% names(switrs_reduced)) {
  switrs_reduced <- switrs_reduced %>% mutate(hour = safe_parse_hhmm(COLLISION_TIME))
}

if ("DAY_OF_WEEK" %in% names(switrs_reduced)) {
  switrs_reduced <- switrs_reduced %>%
    mutate(
      dow_chr = as.character(DAY_OF_WEEK),
      weekend = case_when(
        dow_chr %in% c("6","7","Saturday","Sunday") ~ 1L,
        dow_chr %in% c("1","2","3","4","5","Monday","Tuesday","Wednesday","Thursday","Friday") ~ 0L,
        TRUE ~ NA_integer_
      ),
      weekend = factor(weekend, levels = c(0,1), labels = c("Weekday","Weekend")),
      day_group = case_when(
        dow_chr %in% c("1","2","3","4","Monday","Tuesday","Wednesday","Thursday") ~ "Mon-Thu",
        dow_chr %in% c("5","Friday") ~ "Friday",
        dow_chr %in% c("6","Saturday") ~ "Saturday",
        dow_chr %in% c("7","Sunday") ~ "Sunday",
        TRUE ~ NA_character_
      ),
      day_group = factor(day_group, levels = c("Mon-Thu","Friday","Saturday","Sunday"))
    ) %>%
    select(-dow_chr)
}

if ("hour" %in% names(switrs_reduced)) {
  switrs_reduced <- switrs_reduced %>%
    mutate(
      time_bucket = case_when(
        is.na(hour) ~ NA_character_,
        hour >= 0 & hour <= 5 ~ "Late Night (12–5)",
        hour >= 6 & hour <= 9 ~ "AM Commute (6–9)",
        hour >= 10 & hour <= 15 ~ "Midday (10–3)",
        hour >= 16 & hour <= 19 ~ "PM Commute (4–7)",
        TRUE ~ "Evening (8–11)"
      ),
      time_bucket = factor(time_bucket, levels = c(
        "Late Night (12–5)","AM Commute (6–9)","Midday (10–3)",
        "PM Commute (4–7)","Evening (8–11)"
      ))
    )
}

has_pcf      <- "PCF_VIOL_CATEGORY" %in% names(switrs_reduced)
has_weather  <- "WEATHER_1" %in% names(switrs_reduced)
has_light    <- "LIGHTING" %in% names(switrs_reduced)
has_ped      <- "PEDESTRIAN_ACCIDENT" %in% names(switrs_reduced)
has_hour     <- "hour" %in% names(switrs_reduced)

switrs_reduced <- switrs_reduced %>%
  mutate(
    night = if (has_hour) {
      case_when(
        is.na(hour) ~ NA_integer_,
        hour %in% c(20:23, 0:5) ~ 1L,
        TRUE ~ 0L
      )
    } else NA_integer_,

    friday_night = case_when(
      is.na(night) ~ NA_integer_,
      as.character(DAY_OF_WEEK) %in% c("5", "Friday") & night == 1L ~ 1L,
      as.character(DAY_OF_WEEK) %in% c("1","2","3","4","6","7",
                                       "Monday","Tuesday","Wednesday","Thursday",
                                       "Saturday","Sunday") |
        night == 0L ~ 0L,
      TRUE ~ NA_integer_
    ),

    pcf_dui = if (has_pcf) {
      x <- as.character(PCF_VIOL_CATEGORY)
      x <- str_trim(x)
      case_when(
        x %in% c("01","23") ~ 1L,
        x %in% c("Not Stated", "", NA) ~ NA_integer_,
        TRUE ~ 0L
      )
    } else NA_integer_,

    pcf_speed = if (has_pcf) {
      x <- as.character(PCF_VIOL_CATEGORY)
      x <- str_trim(x)
      case_when(
        x == "03" ~ 1L,
        x %in% c("Not Stated", "", NA) ~ NA_integer_,
        TRUE ~ 0L
      )
    } else NA_integer_,

    is_rain = if (has_weather) {
      x <- as.character(WEATHER_1)
      x <- str_trim(x)
      case_when(
        x == "C" ~ 1L,
        x %in% c("Not Stated", "", NA) ~ NA_integer_,
        TRUE ~ 0L
      )
    } else NA_integer_,

    is_dark = if (has_light) {
      x <- as.character(LIGHTING)
      x <- str_trim(x)
      case_when(
        x %in% c("C","D","E") ~ 1L,
        x %in% c("Not Stated", "", NA) ~ NA_integer_,
        TRUE ~ 0L
      )
    } else NA_integer_,

    is_ped = if (has_ped) {
      x <- as.character(PEDESTRIAN_ACCIDENT)
      x <- str_trim(x)
      case_when(
        x == "Y" ~ 1L,
        x == "N" ~ 0L,
        x %in% c("Not Stated", "", NA) ~ NA_integer_,
        TRUE ~ NA_integer_
      )
    } else NA_integer_,

    night_dui  = ifelse(!is.na(night) & !is.na(pcf_dui),     as.integer(night == 1L & pcf_dui == 1L), NA_integer_),
    rain_speed = ifelse(!is.na(is_rain) & !is.na(pcf_speed), as.integer(is_rain == 1L & pcf_speed == 1L), NA_integer_),
    ped_dark   = ifelse(!is.na(is_ped) & !is.na(is_dark),    as.integer(is_ped == 1L & is_dark == 1L), NA_integer_)
  ) %>%
  mutate(across(
    c(night, friday_night, pcf_dui, pcf_speed, is_rain, is_dark, is_ped, night_dui, rain_speed, ped_dark),
    ~ factor(.x, levels = c(0,1), labels = c("No","Yes"))
  ))
```

# Grouped lighting/weather variables for more interpretable modeling
```{r}
switrs_reduced <- switrs_reduced %>%
  mutate(
    lighting_group = case_when(
      as.character(LIGHTING) == "A" ~ "Daylight",
      as.character(LIGHTING) == "B" ~ "Dusk/Dawn",
      as.character(LIGHTING) %in% c("C", "D", "E") ~ "Dark",
      TRUE ~ NA_character_
    ),
    lighting_group = factor(
      lighting_group,
      levels = c("Daylight", "Dusk/Dawn", "Dark")
    ),

    weather_group = case_when(
      as.character(WEATHER_1) == "A" ~ "Clear",
      as.character(WEATHER_1) == "C" ~ "Rain",
      as.character(WEATHER_1) %in% c("B", "D", "E", "F", "G") ~ "Other Weather",
      TRUE ~ NA_character_
    ),
    weather_group = factor(
      weather_group,
      levels = c("Clear", "Rain", "Other Weather")
    )
  )

cat("Grouped variables added: lighting_group, weather_group\n")
```

# School-zone / school-bus proxy
```{r}
switrs_reduced <- switrs_reduced %>%
  mutate(
    school_bus_proxy = case_when(
      as.character(PED_ACTION) == "G" ~ "Yes",
      as.character(SPECIAL_COND) %in% c("1", "3") ~ "Yes",
      !is.na(PED_ACTION) | !is.na(SPECIAL_COND) ~ "No",
      TRUE ~ NA_character_
    ),
    school_bus_proxy = factor(school_bus_proxy, levels = c("No", "Yes"))
  )

cat("Added school_bus_proxy\n")
```

# Additional subgroup interaction variables
```{r}
switrs_reduced <- switrs_reduced %>%
  mutate(
    intersection_ped = case_when(
      as.character(INTERSECTION) %in% c("Y", "Yes") & as.character(is_ped) == "Yes" ~ "Yes",
      as.character(INTERSECTION) %in% c("N", "No", "Not Stated") | as.character(is_ped) %in% c("No", "Not Stated") ~ "No",
      TRUE ~ NA_character_
    ),
    intersection_ped = factor(intersection_ped, levels = c("No", "Yes")),

    highway_night = case_when(
      as.character(STATE_HWY_IND) %in% c("Y", "Yes") & as.character(night) == "Yes" ~ "Yes",
      as.character(STATE_HWY_IND) %in% c("N", "No", "Not Stated") | as.character(night) == "No" ~ "No",
      TRUE ~ NA_character_
    ),
    highway_night = factor(highway_night, levels = c("No", "Yes"))
  )

cat("Added subgroup interaction variables: intersection_ped, highway_night\n")
```

# HYPOTHESIS TEST (ONE-SIDED): Weekend has higher Severe/Fatal rate than Weekday
```{r}
if (!all(c("weekend", "SevereBinary") %in% names(switrs_reduced))) {
  stop("Need 'weekend' and 'SevereBinary' in switrs_reduced. Run the feature engineering + outcome chunks first.")
}

df_wk <- switrs_reduced %>%
  dplyr::filter(!is.na(weekend), !is.na(SevereBinary)) %>%
  dplyr::mutate(
    weekend = as.character(weekend),
    is_severe = (SevereBinary == "Severe/Fatal")
  ) %>%
  dplyr::filter(weekend %in% c("Weekend", "Weekday"))

n_weekend <- sum(df_wk$weekend == "Weekend")
n_weekday <- sum(df_wk$weekend == "Weekday")

x_weekend <- sum(df_wk$is_severe[df_wk$weekend == "Weekend"], na.rm = TRUE)
x_weekday <- sum(df_wk$is_severe[df_wk$weekend == "Weekday"], na.rm = TRUE)

p_weekend <- x_weekend / n_weekend
p_weekday <- x_weekday / n_weekday
diff_pp   <- 100 * (p_weekend - p_weekday)

cat("\n============================================================\n")
cat("ONE-SIDED TWO-PROPORTION TEST: Weekend > Weekday\n")
cat("============================================================\n")

cat("\nH0: p_weekend <= p_weekday\n")
cat("HA: p_weekend  > p_weekday\n")

cat("\nGroup summaries:\n")
cat("  Weekend: n =", n_weekend, "| severe =", x_weekend, "| p̂ =", round(p_weekend, 4), "\n")
cat("  Weekday: n =", n_weekday, "| severe =", x_weekday, "| p̂ =", round(p_weekday, 4), "\n")
cat("\nDifference (Weekend - Weekday) =", round(diff_pp, 3), "percentage points\n")

test_wk <- suppressWarnings(
  prop.test(
    x = c(x_weekend, x_weekday),
    n = c(n_weekend, n_weekday),
    alternative = "greater",
    correct = FALSE
  )
)

cat("\nTest statistic (X-squared):", round(unname(test_wk$statistic), 4), "\n")
cat("One-sided p-value:", signif(test_wk$p.value, 5), "\n")

if (!is.na(test_wk$p.value) && test_wk$p.value < 0.05) {
  cat("\nConclusion (alpha=0.05): Reject H0. Weekends have a HIGHER Severe/Fatal rate than weekdays.\n")
} else {
  cat("\nConclusion (alpha=0.05): Fail to reject H0. Not enough evidence that weekends are higher.\n")
}
```

# Hypothesis testing continued: Severe/Fatal vs Other for all predictors (or as many as possible)
```{r}
max_levels <- 60
chisq_B    <- 2000
min_n_per_group <- 10
verbose_per_var <- TRUE

if (!exists("switrs_reduced")) stop("switrs_reduced not found.")
if (!("SevereBinary" %in% names(switrs_reduced))) stop("SevereBinary not found in switrs_reduced.")
if (!exists("leakage_vars")) leakage_vars <- character(0)

exclude_always <- c(
  "SevereBinary",
  "CASE_ID",
  "COLLISION_SEVERITY",
  "COLLISION_SEVERITY_CODE",
  "FatalBinary","Deaths","NUMBER_KILLED","NUMBER_KILLED_NUM","NUMBER_INJURED"
)

vars_all <- setdiff(names(switrs_reduced), unique(c(exclude_always, leakage_vars)))
cat("Total candidate variables to test:", length(vars_all), "\n")

label_rows_if_available <- function(var_name, tab) {
  if (exists("legend_maps") && exists("lookup_meaning") && var_name %in% names(legend_maps)) {
    rn <- rownames(tab)
    rownames(tab) <- lookup_meaning(rn, legend_maps[[var_name]])
  }
  tab
}

test_one_var_all <- function(var_name, data = switrs_reduced, outcome = "SevereBinary") {

  df <- data %>%
    dplyr::select(all_of(c(outcome, var_name))) %>%
    dplyr::filter(!is.na(.data[[outcome]]))

  df[[outcome]] <- factor(df[[outcome]], levels = c("Other","Severe/Fatal"))
  x <- df[[var_name]]

  if (all(is.na(x))) {
    if (verbose_per_var) {
      cat("\n============================================================\n")
      cat("VARIABLE:", var_name, "\n")
      cat("============================================================\n")
      cat("SKIP: All missing\n")
    }
    return(tibble::tibble(variable = var_name, type = "NA", test = "SKIP",
                          p_value = NA_real_, p_value_alt = NA_real_,
                          n_used = 0L, note = "All missing"))
  }

  if (is.numeric(x) || is.integer(x)) {
    ux <- unique(x[is.finite(x)])
    if (length(ux) <= 2) x <- as.factor(x)
  }

  if (is.numeric(x) || is.integer(x)) {
    df2 <- df %>%
      dplyr::mutate(x_num = as.numeric(.data[[var_name]])) %>%
      dplyr::filter(is.finite(x_num))

    n_other <- sum(df2[[outcome]] == "Other")
    n_sev   <- sum(df2[[outcome]] == "Severe/Fatal")

    if (verbose_per_var) {
      cat("\n============================================================\n")
      cat("VARIABLE:", var_name, "\n")
      cat("============================================================\n")
      cat("Type: numeric\n")
      cat("H0: mean(x|Severe/Fatal) = mean(x|Other)\n")
      cat("HA: means differ\n\n")
      cat("Group means:\n")
      cat("  Other:       n =", n_other, " mean =", round(mean(df2$x_num[df2[[outcome]]=="Other"]), 4), "\n")
      cat("  Severe/Fatal:n =", n_sev,   " mean =", round(mean(df2$x_num[df2[[outcome]]=="Severe/Fatal"]), 4), "\n\n")
    }

    if (n_other < min_n_per_group || n_sev < min_n_per_group) {
      if (verbose_per_var) cat("SKIP: Too few observations in one group\n")
      return(tibble::tibble(variable = var_name, type = "numeric", test = "SKIP",
                            p_value = NA_real_, p_value_alt = NA_real_,
                            n_used = nrow(df2), note = "Too few observations in one group"))
    }

    df2$g <- df2[[outcome]]

    tt <- tryCatch(t.test(x_num ~ g, data = df2, var.equal = FALSE), error = function(e) NULL)
    wx <- tryCatch(wilcox.test(x_num ~ g, data = df2, exact = FALSE), error = function(e) NULL)

    if (verbose_per_var) {
      cat("Welch t-test p-value:", if (!is.null(tt)) signif(tt$p.value, 6) else "NA", "\n")
      cat("Wilcoxon p-value:    ", if (!is.null(wx)) signif(wx$p.value, 6) else "NA", "\n")
    }

    return(tibble::tibble(
      variable = var_name,
      type = "numeric",
      test = "Welch t-test (+ Wilcoxon)",
      p_value = if (!is.null(tt)) tt$p.value else NA_real_,
      p_value_alt = if (!is.null(wx)) wx$p.value else NA_real_,
      n_used = nrow(df2),
      note = NA_character_
    ))
  }

  x_chr <- normalize_not_stated(x, blank_to_notstated = TRUE)
  df2 <- df %>% dplyr::mutate(x = x_chr) %>% dplyr::filter(!is.na(x))

  k <- length(unique(df2$x))
  if (k <= 1) {
    if (verbose_per_var) {
      cat("\n============================================================\n")
      cat("VARIABLE:", var_name, "\n")
      cat("============================================================\n")
      cat("SKIP: Constant after filtering\n")
    }
    return(tibble::tibble(variable = var_name, type = "categorical", test = "SKIP",
                          p_value = NA_real_, p_value_alt = NA_real_,
                          n_used = nrow(df2), note = "Constant after filtering"))
  }
  if (k > max_levels) {
    if (verbose_per_var) {
      cat("\n============================================================\n")
      cat("VARIABLE:", var_name, "\n")
      cat("============================================================\n")
      cat("SKIP: High cardinality (", k, " levels > ", max_levels, ")\n", sep = "")
    }
    return(tibble::tibble(variable = var_name, type = "categorical", test = "SKIP",
                          p_value = NA_real_, p_value_alt = NA_real_,
                          n_used = nrow(df2), note = paste0("High cardinality (", k, " levels > ", max_levels, ")")))
  }

  tab <- table(df2$x, df2[[outcome]], useNA = "no")

  if (verbose_per_var) {
    cat("\n============================================================\n")
    cat("VARIABLE:", var_name, "\n")
    cat("============================================================\n")
    cat("Type: categorical\n\n")
    tab_print <- label_rows_if_available(var_name, tab)
    cat("Contingency table (rows = predictor, cols = outcome):\n")
    print(addmargins(tab_print))
    cat("\n")
  }

  if (nrow(tab) == 2) {
    x_succ <- tab[, "Severe/Fatal"]
    n_tot  <- rowSums(tab)
    pt <- suppressWarnings(prop.test(x = x_succ, n = n_tot, correct = FALSE))

    if (verbose_per_var) cat("Two-proportion test p-value:", signif(pt$p.value, 6), "\n")

    return(tibble::tibble(
      variable = var_name,
      type = "categorical (2 levels)",
      test = "Two-proportion test (prop.test)",
      p_value = pt$p.value,
      p_value_alt = NA_real_,
      n_used = sum(tab),
      note = NA_character_
    ))
  }

  cs0 <- suppressWarnings(chisq.test(tab))
  use_sim <- any(cs0$expected < 5)
  cs <- if (use_sim) suppressWarnings(chisq.test(tab, simulate.p.value = TRUE, B = chisq_B)) else cs0

  if (verbose_per_var) cat("Chi-square p-value:", signif(cs$p.value, 6), if (use_sim) " (sim)\n" else "\n")

  tibble::tibble(
    variable = var_name,
    type = paste0("categorical (", nrow(tab), " levels)"),
    test = if (use_sim) paste0("Chi-square (sim p, B=", chisq_B, ")") else "Chi-square",
    p_value = cs$p.value,
    p_value_alt = NA_real_,
    n_used = sum(tab),
    note = NA_character_
  )
}

results_all <- purrr::map_dfr(vars_all, ~ test_one_var_all(.x))

results_all <- results_all %>%
  dplyr::mutate(
    p_adj_fdr = p.adjust(p_value, method = "BH")
  ) %>%
  dplyr::arrange(p_adj_fdr, p_value)

cat("\n\n==================== SUMMARY (sorted by FDR) ====================\n")
print(head(results_all, 50))

cat("\nHow many were skipped?\n")
print(table(results_all$test, useNA = "ifany"))

cat("\nTop skip reasons:\n")
print(results_all %>% dplyr::filter(test == "SKIP") %>% dplyr::count(note, sort = TRUE) %>% head(20))
```

# Month and seasonality panel (feature engineering and storyboard-ready rate plot)
```{r}
if (!("COLLISION_DATE" %in% names(switrs_reduced))) {
  stop("COLLISION_DATE not found in switrs_reduced.")
}

parse_date_multi <- function(x) {
  x <- as.character(x)
  x <- trimws(x)
  x[x %in% c("", "Not Stated", "UNKNOWN", "NA", "N/A", "NULL", "null")] <- NA_character_

  d1 <- suppressWarnings(as.Date(x, format = "%Y%m%d"))
  d2 <- suppressWarnings(as.Date(x, format = "%Y-%m-%d"))
  d3 <- suppressWarnings(as.Date(x, format = "%m/%d/%Y"))

  out <- d1
  out[is.na(out)] <- d2[is.na(out)]
  out[is.na(out)] <- d3[is.na(out)]
  out
}

switrs_reduced$collision_date <- parse_date_multi(switrs_reduced$COLLISION_DATE)

cat("collision_date parsed % = ",
    round(mean(!is.na(switrs_reduced$collision_date)) * 100, 2), "%\n", sep = "")

switrs_reduced <- switrs_reduced %>%
  dplyr::mutate(
    month = as.integer(format(collision_date, "%m")),
    month_name = factor(month.abb[month], levels = month.abb, ordered = TRUE)
  )

month_tbl <- switrs_reduced %>%
  dplyr::filter(!is.na(SevereBinary), !is.na(month_name)) %>%
  dplyr::group_by(month_name) %>%
  dplyr::summarise(
    n_crashes = dplyr::n(),
    severe_n = sum(SevereBinary == "Severe/Fatal", na.rm = TRUE),
    severe_per_100 = 100 * severe_n / n_crashes,
    .groups = "drop"
  ) %>%
  dplyr::arrange(month_name)

cat("Severe/Fatal rate per 100 crashes by month:\n")
print(month_tbl)

ggplot(month_tbl, aes(x = month_name, y = severe_per_100)) +
  geom_col() +
  geom_text(aes(label = round(severe_per_100, 2)), vjust = -0.3, size = 3.5) +
  labs(
    title = "Severe/Fatal Rate by Month",
    x = "Month",
    y = "Severe/Fatal per 100 crashes"
  ) +
  theme(panel.grid.minor = element_blank())
```

# Statistics report chunk: Hypothesis tests on Severe/Fatal proportions (week 8 deliverable)
```{r}
two_prop_report <- function(data, group_var, group_a, group_b,
                            outcome_var = "SevereBinary",
                            outcome_success = "Severe/Fatal",
                            title = NULL) {

  if (!all(c(group_var, outcome_var) %in% names(data))) {
    stop("Missing required columns for test: ", group_var, " and/or ", outcome_var)
  }

  df <- data %>%
    dplyr::select(all_of(c(group_var, outcome_var))) %>%
    dplyr::filter(!is.na(.data[[group_var]]), !is.na(.data[[outcome_var]])) %>%
    dplyr::mutate(
      g = as.character(.data[[group_var]]),
      y = (.data[[outcome_var]] == outcome_success)
    ) %>%
    dplyr::filter(g %in% c(group_a, group_b))

  n_a <- sum(df$g == group_a)
  n_b <- sum(df$g == group_b)
  x_a <- sum(df$y[df$g == group_a], na.rm = TRUE)
  x_b <- sum(df$y[df$g == group_b], na.rm = TRUE)

  p_a <- x_a / n_a
  p_b <- x_b / n_b
  diff <- p_a - p_b

  test <- suppressWarnings(stats::prop.test(x = c(x_a, x_b), n = c(n_a, n_b), correct = FALSE))
  ci <- test$conf.int
  pval <- test$p.value

  if (is.null(title)) title <- paste0(outcome_success, " proportion: ", group_a, " vs ", group_b)

  cat("\n============================================================\n")
  cat("STATISTICS REPORT —", title, "\n")
  cat("============================================================\n")

  cat("\nH0: p(", group_a, ") = p(", group_b, ")\n", sep = "")
  cat("HA: p(", group_a, ") ≠ p(", group_b, ")\n", sep = "")

  cat("\nGroup summaries:\n")
  cat("  ", group_a, ": n = ", n_a, ", severe = ", x_a,
      ", p̂ = ", round(p_a, 4), "\n", sep = "")
  cat("  ", group_b, ": n = ", n_b, ", severe = ", x_b,
      ", p̂ = ", round(p_b, 4), "\n", sep = "")

  cat("\nEstimated difference (p̂A − p̂B) = ", round(diff, 4), "\n", sep = "")
  cat("95% CI for (pA − pB): [", round(ci[1], 4), ", ", round(ci[2], 4), "]\n", sep = "")
  cat("Two-proportion test p-value: ", signif(pval, 4), "\n", sep = "")

  if (!is.na(pval) && pval < 0.05) {
    cat("\nInterpretation: Reject H0 at α = 0.05. The Severe/Fatal rate differs between groups.\n")
  } else {
    cat("\nInterpretation: Fail to reject H0 at α = 0.05. No statistically significant difference detected.\n")
  }

  invisible(list(
    group_a = group_a, group_b = group_b,
    n = c(n_a, n_b), x = c(x_a, x_b),
    p = c(p_a, p_b), diff = diff,
    ci = ci, p_value = pval
  ))
}

if ("weekend" %in% names(switrs_reduced)) {
  two_prop_report(
    data = switrs_reduced,
    group_var = "weekend",
    group_a = "Weekend",
    group_b = "Weekday",
    title = "Weekend vs Weekday"
  )
} else {
  cat("\nSkipping Test 1: 'weekend' not found.\n")
}

if ("WEATHER_1" %in% names(switrs_reduced)) {
  df_weather <- switrs_reduced %>%
    dplyr::mutate(
      WEATHER_1 = trimws(as.character(WEATHER_1)),
      WEATHER_1 = dplyr::case_when(
        WEATHER_1 %in% c("A","C") ~ WEATHER_1,
        TRUE ~ NA_character_
      )
    )

  two_prop_report(
    data = df_weather,
    group_var = "WEATHER_1",
    group_a = "C",
    group_b = "A",
    title = "Raining (WEATHER_1=C) vs Clear (WEATHER_1=A)"
  )
} else {
  cat("\nSkipping Test 2: 'WEATHER_1' not found.\n")
}

if ("night_dui" %in% names(switrs_reduced)) {
  df_nd <- switrs_reduced %>%
    dplyr::mutate(night_dui = as.character(night_dui)) %>%
    dplyr::filter(night_dui %in% c("Yes","No"))

  two_prop_report(
    data = df_nd,
    group_var = "night_dui",
    group_a = "Yes",
    group_b = "No",
    title = "Night × DUI proxy (night_dui=Yes) vs Not (night_dui=No)"
  )
} else {
  cat("\nSkipping Test 3: 'night_dui' not found (re-run the time-feature chunk if needed).\n")
}
```

# Statistics report (add-on):Dark vs Daylight, Alcohol vs No Alcohol, Rain×Speeding
```{r}
if (!exists("two_prop_report")) stop("two_prop_report() not found. Run the main Statistics Report chunk first.")

if ("LIGHTING" %in% names(switrs_reduced)) {
  df_light <- switrs_reduced %>%
    dplyr::filter(!is.na(SevereBinary)) %>%
    dplyr::mutate(
      LIGHTING = trimws(as.character(LIGHTING)),
      light_group = dplyr::case_when(
        LIGHTING == "A" ~ "Daylight",
        LIGHTING %in% c("C","D","E") ~ "Dark",
        TRUE ~ NA_character_
      )
    )

  two_prop_report(
    data = df_light,
    group_var = "light_group",
    group_a = "Dark",
    group_b = "Daylight",
    title = "Dark (LIGHTING=C/D/E) vs Daylight (LIGHTING=A)"
  )
} else {
  cat("\nSkipping Test 4: 'LIGHTING' not found.\n")
}

if ("ALCOHOL_INVOLVED" %in% names(switrs_reduced)) {
  df_alc <- switrs_reduced %>%
    dplyr::filter(!is.na(SevereBinary)) %>%
    dplyr::mutate(
      ALCOHOL_INVOLVED = trimws(as.character(ALCOHOL_INVOLVED)),
      alc_group = dplyr::case_when(
        ALCOHOL_INVOLVED == "Y" ~ "Alcohol",
        ALCOHOL_INVOLVED == "N" ~ "No Alcohol",
        TRUE ~ NA_character_
      )
    )

  two_prop_report(
    data = df_alc,
    group_var = "alc_group",
    group_a = "Alcohol",
    group_b = "No Alcohol",
    title = "Alcohol involved vs Not alcohol involved"
  )
} else {
  cat("\nSkipping Test 5: 'ALCOHOL_INVOLVED' not found.\n")
}

if ("rain_speed" %in% names(switrs_reduced)) {
  df_rs <- switrs_reduced %>%
    dplyr::mutate(rain_speed = as.character(rain_speed)) %>%
    dplyr::filter(rain_speed %in% c("Yes","No"))

  two_prop_report(
    data = df_rs,
    group_var = "rain_speed",
    group_a = "Yes",
    group_b = "No",
    title = "Rain × Unsafe speed proxy (rain_speed=Yes) vs Not (rain_speed=No)"
  )
} else {
  cat("\nSkipping Test 6: 'rain_speed' not found (re-run the time-feature chunk if needed).\n")
}
```

# Statistics report (subgroups): newly added subgroup variables
```{r}
if (!exists("two_prop_report")) stop("two_prop_report() not found. Run the main Statistics Report chunk first.")

run_subgroup_prop_test <- function(data, var_name, title_text) {

  if (!var_name %in% names(data)) {
    cat("\nSkipping", title_text, ":", var_name, "not found.\n")
    return(invisible(NULL))
  }

  df_sub <- data %>%
    dplyr::filter(!is.na(SevereBinary)) %>%
    dplyr::mutate(
      subgroup_raw = as.character(.data[[var_name]]),
      subgroup_raw = stringr::str_trim(subgroup_raw),
      subgroup = dplyr::case_when(
        subgroup_raw %in% c("Yes", "Y", "1") ~ "Yes",
        subgroup_raw %in% c("No", "N", "0") ~ "No",
        TRUE ~ NA_character_
      )
    ) %>%
    dplyr::filter(!is.na(subgroup))

  if (nrow(df_sub) == 0) {
    cat("\nSkipping", title_text, ":", var_name, "has no usable Yes/No rows.\n")
    return(invisible(NULL))
  }

  two_prop_report(
    data = df_sub,
    group_var = "subgroup",
    group_a = "Yes",
    group_b = "No",
    title = title_text
  )
}

subgroup_test_results <- list()

subgroup_test_results$friday_night <- run_subgroup_prop_test(
  switrs_reduced,
  "friday_night",
  "Friday night (friday_night=Yes) vs Not (friday_night=No)"
)

subgroup_test_results$school_bus_proxy <- run_subgroup_prop_test(
  switrs_reduced,
  "school_bus_proxy",
  "School bus proxy (school_bus_proxy=Yes) vs Not (school_bus_proxy=No)"
)

subgroup_test_results$intersection_ped <- run_subgroup_prop_test(
  switrs_reduced,
  "intersection_ped",
  "Intersection × pedestrian (intersection_ped=Yes) vs Not (intersection_ped=No)"
)

subgroup_test_results$highway_night <- run_subgroup_prop_test(
  switrs_reduced,
  "highway_night",
  "Highway × night (highway_night=Yes) vs Not (highway_night=No)"
)
```

# Scenario Ranking (top combinations & rates & counts)
```{r}
df_scen <- switrs_reduced %>%
  dplyr::filter(!is.na(SevereBinary)) %>%
  dplyr::mutate(
    time_bucket = if ("time_bucket" %in% names(.)) as.character(time_bucket) else NA_character_,
    weekend     = if ("weekend" %in% names(.)) as.character(weekend) else NA_character_,
    light = if ("LIGHTING" %in% names(.)) {
      x <- trimws(as.character(LIGHTING))
      dplyr::case_when(
        x == "A" ~ "Daylight",
        x %in% c("C","D","E") ~ "Dark",
        TRUE ~ NA_character_
      )
    } else NA_character_,
    ped = if ("is_ped" %in% names(.)) {
      x <- as.character(is_ped)
      dplyr::case_when(
        x %in% c("Yes","No") ~ x,
        TRUE ~ NA_character_
      )
    } else if ("PEDESTRIAN_ACCIDENT" %in% names(.)) {
      x <- trimws(as.character(PEDESTRIAN_ACCIDENT))
      dplyr::case_when(
        x == "Y" ~ "Yes",
        x == "N" ~ "No",
        TRUE ~ NA_character_
      )
    } else NA_character_,
    alc_yesno = if ("ALCOHOL_INVOLVED" %in% names(.)) {
      x <- trimws(as.character(ALCOHOL_INVOLVED))
      dplyr::case_when(
        x == "Y" ~ "Yes",
        x == "N" ~ "No",
        TRUE ~ NA_character_
      )
    } else NA_character_,
    dui_yesno = if ("pcf_dui" %in% names(.)) {
      x <- as.character(pcf_dui)
      dplyr::case_when(
        x %in% c("Yes","No") ~ x,
        TRUE ~ NA_character_
      )
    } else NA_character_,
    impairment = dplyr::case_when(
      alc_yesno == "Yes" | dui_yesno == "Yes" ~ "Yes",
      alc_yesno == "No"  & dui_yesno == "No"  ~ "No",
      TRUE ~ NA_character_
    ),
    rain = if ("is_rain" %in% names(.)) {
      x <- as.character(is_rain)
      dplyr::case_when(x %in% c("Yes","No") ~ x, TRUE ~ NA_character_)
    } else NA_character_,
    speed = if ("pcf_speed" %in% names(.)) {
      x <- as.character(pcf_speed)
      dplyr::case_when(x %in% c("Yes","No") ~ x, TRUE ~ NA_character_)
    } else NA_character_
  )

scenario_vars <- c("time_bucket","weekend","light","impairment","ped")
scenario_vars <- scenario_vars[scenario_vars %in% names(df_scen)]

df_scen2 <- df_scen %>%
  dplyr::filter(dplyr::if_all(dplyr::all_of(scenario_vars), ~ !is.na(.x)))

cat("Rows with complete CORE scenario fields:", nrow(df_scen2), "\n")
cat("Scenario dimensions used:", paste(scenario_vars, collapse = ", "), "\n")

df_rank <- df_scen2 %>%
  dplyr::mutate(
    scenario = paste(
      paste0("Time=", time_bucket),
      paste0("DayType=", weekend),
      paste0("Light=", light),
      paste0("Impairment=", impairment),
      paste0("Ped=", ped),
      sep = " | "
    ),
    is_severe = (SevereBinary == "Severe/Fatal")
  ) %>%
  dplyr::group_by(scenario) %>%
  dplyr::summarise(
    n_crashes = dplyr::n(),
    severe_n = sum(is_severe, na.rm = TRUE),
    severe_per_100 = 100 * severe_n / n_crashes,
    .groups = "drop"
  )

min_crashes_count <- 30

top_by_count <- df_rank %>%
  dplyr::filter(n_crashes >= min_crashes_count) %>%
  dplyr::arrange(dplyr::desc(severe_n), dplyr::desc(n_crashes))

cat("\nTop 15 scenarios by Severe/Fatal COUNT",
    "(min_crashes=", min_crashes_count, "): \n", sep = "")
print(head(top_by_count, 15))
cat("\nScenarios meeting COUNT filter:", nrow(top_by_count), "\n")

min_crashes_rate <- 15
min_severe_rate  <- 1

top_by_rate <- df_rank %>%
  dplyr::filter(n_crashes >= min_crashes_rate, severe_n >= min_severe_rate) %>%
  dplyr::arrange(dplyr::desc(severe_per_100), dplyr::desc(n_crashes))

cat("\nTop 15 scenarios by Severe/Fatal RATE",
    "(min_crashes=", min_crashes_rate, ", min_severe=", min_severe_rate, "): \n", sep = "")
print(head(top_by_rate, 15))
cat("\nScenarios meeting RATE filter:", nrow(top_by_rate), "\n")

interaction_vars <- intersect(c("night_dui","rain_speed","ped_dark"), names(switrs_reduced))

if (length(interaction_vars) > 0) {
  cat("\nInteraction-term summaries (Severe/Fatal per 100 crashes):\n")

  interaction_tbl <- purrr::map_dfr(interaction_vars, function(v) {
    switrs_reduced %>%
      dplyr::filter(!is.na(SevereBinary)) %>%
      dplyr::mutate(g = as.character(.data[[v]])) %>%
      dplyr::filter(g %in% c("Yes","No")) %>%
      dplyr::group_by(term = v, group = g) %>%
      dplyr::summarise(
        n_crashes = dplyr::n(),
        severe_n = sum(SevereBinary == "Severe/Fatal"),
        severe_per_100 = 100 * severe_n / n_crashes,
        .groups = "drop"
      )
  })

  print(interaction_tbl %>% arrange(term, desc(group)))
} else {
  cat("\nNo interaction variables found (night_dui/rain_speed/ped_dark). If we added the time-feature chunk, re-run it.\n")
}
```

# report_data_structure_and_descriptives
```{r, message=FALSE, warning=FALSE}
raw_df_for_report <- if (exists("switrs")) switrs else switrs_reduced

count_storage_types <- function(df) {
  tibble::tibble(
    storage_type = c("numeric", "integer", "character", "factor", "logical", "other"),
    n_columns = c(
      sum(vapply(df, function(x) is.numeric(x) && !is.integer(x), logical(1))),
      sum(vapply(df, is.integer, logical(1))),
      sum(vapply(df, is.character, logical(1))),
      sum(vapply(df, is.factor, logical(1))),
      sum(vapply(df, is.logical, logical(1))),
      sum(!vapply(df, function(x) {
        is.numeric(x) || is.integer(x) || is.character(x) || is.factor(x) || is.logical(x)
      }, logical(1)))
    )
  ) %>%
    dplyr::mutate(pct_columns = round(100 * n_columns / sum(n_columns), 1))
}

numeric_desc_tbl <- function(data, vars, group_var = NULL) {
  vars <- intersect(vars, names(data))
  if (length(vars) == 0) return(tibble::tibble())

  keep_cols <- if (is.null(group_var)) vars else c(group_var, vars)

  long_df <- data %>%
    dplyr::select(dplyr::all_of(keep_cols)) %>%
    tidyr::pivot_longer(cols = dplyr::all_of(vars), names_to = "variable", values_to = "value")

  if (!is.null(group_var) && group_var %in% names(long_df)) {
    out <- long_df %>%
      dplyr::group_by(.data[[group_var]], variable) %>%
      dplyr::summarise(
        n_nonmissing = sum(is.finite(value)),
        missing_pct = round(100 * mean(!is.finite(value)), 2),
        mean = mean(value, na.rm = TRUE),
        sd = sd(value, na.rm = TRUE),
        median = median(value, na.rm = TRUE),
        IQR = IQR(value, na.rm = TRUE),
        min = min(value, na.rm = TRUE),
        max = max(value, na.rm = TRUE),
        .groups = "drop"
      )
  } else {
    out <- long_df %>%
      dplyr::group_by(variable) %>%
      dplyr::summarise(
        n_nonmissing = sum(is.finite(value)),
        missing_pct = round(100 * mean(!is.finite(value)), 2),
        mean = mean(value, na.rm = TRUE),
        sd = sd(value, na.rm = TRUE),
        median = median(value, na.rm = TRUE),
        IQR = IQR(value, na.rm = TRUE),
        min = min(value, na.rm = TRUE),
        max = max(value, na.rm = TRUE),
        .groups = "drop"
      )
  }

  out %>%
    dplyr::mutate(
      dplyr::across(c(mean, sd, median, IQR, min, max), ~ round(.x, 2))
    )
}

cat_desc_tbl <- function(data, vars, outcome = "SevereBinary") {
  vars <- intersect(vars, names(data))
  if (length(vars) == 0 || !(outcome %in% names(data))) return(tibble::tibble())

  purrr::map_dfr(vars, function(v) {
    tmp <- data %>%
      dplyr::filter(!is.na(.data[[v]]), !is.na(.data[[outcome]])) %>%
      dplyr::mutate(
        level = as.character(.data[[v]]),
        outcome_chr = as.character(.data[[outcome]])
      ) %>%
      dplyr::count(level, outcome_chr, name = "n") %>%
      tidyr::pivot_wider(names_from = outcome_chr, values_from = n, values_fill = 0)

    if (!("Other" %in% names(tmp))) tmp$Other <- 0
    if (!("Severe/Fatal" %in% names(tmp))) tmp$`Severe/Fatal` <- 0

    tmp %>%
      dplyr::mutate(
        variable = v,
        total = Other + `Severe/Fatal`,
        severe_rate = round(`Severe/Fatal` / total, 4)
      ) %>%
      dplyr::select(variable, level, total, Other, `Severe/Fatal`, severe_rate)
  })
}

case_id_nunique <- if ("CASE_ID" %in% names(raw_df_for_report)) {
  dplyr::n_distinct(raw_df_for_report$CASE_ID)
} else {
  NA_integer_
}

dup_case_id <- if ("CASE_ID" %in% names(raw_df_for_report)) {
  sum(duplicated(raw_df_for_report$CASE_ID))
} else {
  NA_integer_
}

raw_structure_tbl <- tibble::tibble(
  dataset = "Merged crash-level SWITRS file",
  rows = nrow(raw_df_for_report),
  columns = ncol(raw_df_for_report),
  unique_case_id = case_id_nunique,
  duplicated_case_id = dup_case_id
)

analysis_structure_tbl <- tibble::tibble(
  dataset = "Analysis frame after cleaning / feature engineering",
  rows = nrow(switrs_reduced),
  columns = ncol(switrs_reduced),
  severebinary_missing = if ("SevereBinary" %in% names(switrs_reduced)) sum(is.na(switrs_reduced$SevereBinary)) else NA_integer_
)

raw_missing_tbl <- tibble::tibble(
  variable = names(raw_df_for_report),
  missing_pct = round(100 * colMeans(is.na(raw_df_for_report)), 2)
) %>%
  dplyr::arrange(dplyr::desc(missing_pct))

cat("Merged file structure:\n")
print(raw_structure_tbl)
cat("\nMerged file column types:\n")
print(count_storage_types(raw_df_for_report))
cat("\nAnalysis-frame structure:\n")
print(analysis_structure_tbl)
cat("\nAnalysis-frame column types:\n")
print(count_storage_types(switrs_reduced))
cat("\nTop 15 variables by missingness in merged file:\n")
print(head(raw_missing_tbl, 15))

report_num_vars <- intersect(
  c("year", "n_parties", "n_victims", "mean_party_age", "mean_victim_age", "min_victim_age", "max_victim_age"),
  names(switrs_reduced)
)

report_cat_vars <- intersect(
  c("weekend", "night", "is_rain", "is_dark", "pcf_dui", "pcf_speed", "INTERSECTION", "TYPE_OF_COLLISION"),
  names(switrs_reduced)
)

cat("\nNumeric descriptive statistics (overall):\n")
print(numeric_desc_tbl(switrs_reduced, report_num_vars))

if ("SevereBinary" %in% names(switrs_reduced)) {
  cat("\nNumeric descriptive statistics by severity group:\n")
  print(
    numeric_desc_tbl(
      switrs_reduced %>% dplyr::filter(!is.na(SevereBinary)),
      vars = report_num_vars,
      group_var = "SevereBinary"
    )
  )
}

cat("\nCategorical descriptive summaries (counts + severe rate):\n")
print(cat_desc_tbl(switrs_reduced, report_cat_vars))
```

# Storyboard panels (FIXED)
```{r storyboard_panels, fig.width=14, fig.height=10}
library(dplyr)
library(ggplot2)
library(stringr)

base_theme_fix <- theme(
  plot.title = element_text(size = 16, face = "bold"),
  plot.subtitle = element_text(size = 11),
  plot.margin = margin(t = 12, r = 12, b = 12, l = 12),
  panel.grid.minor = element_blank()
)

if (exists("hotspots_by_count") && nrow(hotspots_by_count) > 0) {

  p1 <- hotspots_by_count %>%
    slice_head(n = 15) %>%
    mutate(location_key = factor(location_key, levels = rev(location_key))) %>%
    ggplot(aes(x = location_key, y = severe_n)) +
    geom_col() +
    coord_flip(clip = "off") +
    labs(
      title = "Top 15 Hotspots by Severe/Fatal Count",
      subtitle = "Higher count = larger severe/fatal burden (deployment prioritization)",
      x = NULL,
      y = "Severe/Fatal crashes"
    ) +
    base_theme_fix

  print(p1)

} else {
  cat("hotspots_by_count not found or empty — run the hotspots chunk first.\n")
}

if (exists("top_by_rate") && nrow(top_by_rate) > 0) {

  p2 <- top_by_rate %>%
    slice_head(n = 15) %>%
    mutate(scenario = factor(scenario, levels = rev(scenario))) %>%
    ggplot(aes(x = scenario, y = severe_per_100)) +
    geom_col() +
    coord_flip(clip = "off") +
    labs(
      title = "Top 15 Scenarios by Severe/Fatal Rate",
      subtitle = "Higher rate = higher risk intensity (targeted interventions)",
      x = NULL,
      y = "Severe/Fatal per 100 crashes"
    ) +
    base_theme_fix +
    theme(axis.text.y = element_text(size = 9))

  print(p2)

} else {
  cat("top_by_rate not found or empty — run the scenario ranking chunk first.\n")
}

interaction_vars <- intersect(c("night_dui","rain_speed","ped_dark"), names(switrs_reduced))

if (length(interaction_vars) > 0) {

  interaction_tbl <- purrr::map_dfr(interaction_vars, function(v) {
    switrs_reduced %>%
      filter(!is.na(SevereBinary)) %>%
      mutate(group = as.character(.data[[v]])) %>%
      filter(group %in% c("Yes","No")) %>%
      group_by(term = v, group) %>%
      summarise(
        n_crashes = n(),
        severe_n = sum(SevereBinary == "Severe/Fatal"),
        severe_per_100 = 100 * severe_n / n_crashes,
        .groups = "drop"
      )
  }) %>%
    mutate(
      term_label = recode(
        term,
        night_dui  = "Night × DUI proxy",
        rain_speed = "Rain × Unsafe speed proxy",
        ped_dark   = "Pedestrian × Dark"
      ),
      term_label = factor(term_label, levels = rev(unique(term_label))),
      group = factor(group, levels = c("No","Yes"))
    )

  p3 <- interaction_tbl %>%
    ggplot(aes(x = term_label, y = severe_per_100, fill = group)) +
    geom_col(position = position_dodge(width = 0.8)) +
    coord_flip(clip = "off") +
    labs(
      title = "Compound Risk Conditions (Interaction Terms)",
      subtitle = "Severe/Fatal rate per 100 crashes: compare “Yes” vs “No” for each condition",
      x = NULL,
      y = "Severe/Fatal per 100 crashes",
      fill = "Condition present?"
    ) +
    base_theme_fix

  print(p3)

} else {
  cat("No interaction variables found (night_dui/rain_speed/ped_dark). Re-run the time-feature chunk if needed.\n")
}
```

# codebook label dictionaries
```{r}
lookup_meaning <- function(x, legend_vec) {
  x0 <- normalize_not_stated(x, blank_to_notstated = TRUE)

  out <- unname(legend_vec[x0])

  needs_pad <- is.na(out) & grepl("^[0-9]+$", x0)
  x2 <- x0
  x2[needs_pad] <- sprintf("%02d", as.integer(x0[needs_pad]))
  out2 <- unname(legend_vec[x2])

  out[is.na(out)] <- out2[is.na(out)]
  out[is.na(out) | out == ""] <- x0[is.na(out) | out == ""]
  out
}

legend_maps <- list(
  DAY_OF_WEEK = c(
    "1"="Monday","2"="Tuesday","3"="Wednesday","4"="Thursday",
    "5"="Friday","6"="Saturday","7"="Sunday",
    "Not Stated"="Not Stated"
  ),
  LIGHTING = c(
    "A"="Daylight",
    "B"="Dusk - Dawn",
    "C"="Dark - Street Lights",
    "D"="Dark - No Street Lights",
    "E"="Dark - Street Lights Not Functioning",
    "Not Stated"="Not Stated"
  ),
  WEATHER_1 = c(
    "A"="Clear","B"="Cloudy","C"="Raining","D"="Snowing","E"="Fog","F"="Other","G"="Wind",
    "Not Stated"="Not Stated"
  ),
  ROAD_SURFACE = c(
    "A"="Dry","B"="Wet","C"="Snowy or Icy","D"="Slippery (Muddy/Oily/etc.)",
    "Not Stated"="Not Stated"
  ),
  INTERSECTION = c(
    "Y"="Yes","N"="No",
    "1"="Yes","0"="No",
    "Not Stated"="Not Stated"
  ),
  PRIMARY_COLL_FACTOR = c(
    "A"="Code Violation",
    "B"="Other Improper Driving",
    "C"="Other Than Driver",
    "D"="Unknown",
    "E"="Fell Asleep",
    "Not Stated"="Not Stated"
  ),
  TYPE_OF_COLLISION = c(
    "A"="Head-On",
    "B"="Sideswipe",
    "C"="Rear End",
    "D"="Broadside",
    "E"="Hit Object",
    "F"="Overturned",
    "G"="Vehicle/Pedestrian",
    "H"="Other",
    "Not Stated"="Not Stated"
  ),
  PCF_VIOL_CATEGORY = c(
    "01"="DUI (Alcohol/Drug)",
    "02"="Impeding Traffic",
    "03"="Unsafe Speed",
    "04"="Following Too Closely",
    "05"="Wrong Side of Road",
    "06"="Improper Passing",
    "07"="Unsafe Lane Change",
    "08"="Improper Turning",
    "09"="Automobile Right-of-Way",
    "10"="Pedestrian Right-of-Way",
    "11"="Pedestrian Violation",
    "12"="Traffic Signals & Signs",
    "13"="Hazardous Parking",
    "14"="Lights",
    "15"="Brakes",
    "16"="Other Equipment",
    "17"="Other Hazardous Violation",
    "18"="Other Than Driver (or Pedestrian)",
    "21"="Unsafe Starting/Backing",
    "22"="Other Improper Driving",
    "23"="Ped/Other DUI (Alcohol/Drug)",
    "24"="Fell Asleep",
    "00"="Unknown",
    "Not Stated"="Not Stated"
  ),
  HIT_AND_RUN = c(
    "F"="Felony",
    "M"="Misdemeanor",
    "N"="Not Hit and Run",
    "Not Stated"="Not Stated"
  ),
  PEDESTRIAN_ACCIDENT = c(
    "Y"="Yes","N"="No",
    "1"="Yes","0"="No",
    "Not Stated"="Not Stated"
  ),
  MOTORCYCLE_ACCIDENT = c(
    "Y"="Yes","N"="No",
    "1"="Yes","0"="No",
    "Not Stated"="Not Stated"
  ),
  ANY_PARTY_DRUG = c(
    "Y"="Yes","N"="No",
    "1"="Yes","0"="No",
    "Not Stated"="Not Stated"
  )
)

legend_maps$TYPE_OF_COLLISION <- c(
  "A"="Head-On",
  "B"="Sideswipe",
  "C"="Rear End",
  "D"="Broadside",
  "E"="Hit Object",
  "F"="Overturned",
  "G"="Vehicle/Pedestrian",
  "H"="Other",
  "Not Stated"="Not Stated"
)

legend_maps$LIGHTING <- c(
  "A"="Daylight",
  "B"="Dusk - Dawn",
  "C"="Dark - Street Lights",
  "D"="Dark - No Street Lights",
  "E"="Dark - Street Lights Not Functioning",
  "Not Stated"="Not Stated"
)

legend_maps$PEDESTRIAN_ACCIDENT <- c(
  "Y"="Yes","N"="No",
  "Not Stated"="Not Stated"
)

legend_maps$MOTORCYCLE_ACCIDENT <- c(
  "Y"="Yes","N"="No",
  "Not Stated"="Not Stated"
)

legend_maps$ANY_PARTY_DRUG <- c(
  "1"="Yes","0"="No","Y"="Yes","N"="No",
  "Not Stated"="Not Stated"
)
legend_maps$any_party_drug <- legend_maps$ANY_PARTY_DRUG

print_legend <- function(var_name, dedupe_meaning = TRUE) {
  if (!var_name %in% names(legend_maps)) return(invisible(NULL))
  m <- legend_maps[[var_name]]

  df <- data.frame(
    code = names(m),
    meaning = unname(m),
    stringsAsFactors = FALSE
  )

  if (dedupe_meaning) df <- df[!duplicated(df$meaning), ]

  cat("\nLEGEND for", var_name, ":\n")
  print(df, row.names = FALSE)
  invisible(df)
}
```

# EDA: Severity rate table for key categorical variables
```{r}
severity_rate_table <- function(data, var) {
  if (!var %in% names(data)) return(NULL)

  df <- data %>%
    dplyr::filter(!is.na(SevereBinary)) %>%
    dplyr::mutate(v = as.character(.data[[var]])) %>%
    dplyr::mutate(v = ifelse(is.na(v) | v == "", "Not Stated", v))

  if (var %in% names(legend_maps)) {
    map <- legend_maps[[var]]
    df <- df %>% mutate(v_label = dplyr::recode(v, !!!map, .default = v))
    grp <- "v_label"
  } else {
    grp <- "v"
  }

  df %>%
    group_by(.data[[grp]]) %>%
    summarise(
      n = n(),
      severe_n = sum(SevereBinary == "Severe/Fatal"),
      severe_per_100 = 100 * severe_n / n,
      .groups = "drop"
    ) %>%
    rename(category = .data[[grp]]) %>%
    arrange(desc(severe_per_100), desc(n))
}

eda_vars <- intersect(
  c(
    "DAY_OF_WEEK","time_bucket","weekend","night",
    "WEATHER_1","is_rain",
    "LIGHTING","is_dark",
    "ROAD_SURFACE",
    "INTERSECTION","STATE_HWY_IND","TYPE_OF_COLLISION","PRIMARY_COLL_FACTOR",
    "PCF_VIOL_CATEGORY","pcf_dui","pcf_speed",
    "HIT_AND_RUN",
    "ALCOHOL_INVOLVED",
    "PEDESTRIAN_ACCIDENT","BICYCLE_ACCIDENT","MOTORCYCLE_ACCIDENT","TRUCK_ACCIDENT",
    "night_dui","rain_speed","ped_dark"
  ),
  names(switrs_reduced)
)

rate_tables <- purrr::map(eda_vars, ~ severity_rate_table(switrs_reduced, .x))
names(rate_tables) <- eda_vars

rate_tables[1:min(6, length(rate_tables))]
```

# EDA: Decision-oriented severity rate plots (ordered, horizontal)
```{r}
plot_severity_by_variable <- function(var_name,
                                      data = switrs_reduced,
                                      fill_var = "SevereBinary") {

  if (!var_name %in% names(data)) return(NULL)

  data[[fill_var]] <- factor(data[[fill_var]], levels = c("Other", "Severe/Fatal"))

  df <- data %>%
    dplyr::filter(!is.na(.data[[fill_var]])) %>%
    dplyr::transmute(
      x_code = normalize_not_stated(.data[[var_name]]),
      Outcome = .data[[fill_var]]
    ) %>%
    dplyr::count(x_code, Outcome, name = "n") %>%
    dplyr::group_by(x_code) %>%
    dplyr::mutate(
      prop = n / sum(n),
      pct_label = scales::percent(prop, accuracy = 0.1)
    ) %>%
    dplyr::ungroup()

  ordering <- df %>%
    dplyr::group_by(x_code) %>%
    dplyr::summarise(
      severe_prop = sum(prop[Outcome == "Severe/Fatal"], na.rm = TRUE),
      total_n = sum(n),
      .groups = "drop"
    ) %>%
    dplyr::arrange(dplyr::desc(severe_prop), dplyr::desc(total_n)) %>%
    dplyr::pull(x_code)

  df$x_code <- factor(df$x_code, levels = ordering)

  label_fun <- function(x) {
    if (exists("legend_maps") && var_name %in% names(legend_maps)) {
      out <- lookup_meaning(x, legend_maps[[var_name]])
      out[is.na(out) | out == ""] <- x[is.na(out) | out == ""]
      out
    } else {
      x
    }
  }

  p <- ggplot(df, aes(x = x_code, y = prop, fill = Outcome)) +
    geom_bar(stat = "identity", position = "stack") +
    geom_text(
      aes(label = pct_label),
      position = position_stack(vjust = 0.5),
      color = "white",
      size = 3.6,
      fontface = "bold"
    ) +
    coord_flip() +
    scale_y_continuous(expand = c(0, 0), breaks = NULL, labels = NULL) +
    scale_x_discrete(labels = label_fun) +
    scale_fill_manual(
      values = c(
        "Other" = "#A8D5BA",
        "Severe/Fatal" = "#1B5E20"
      ),
      drop = FALSE
    ) +
    labs(
      title = paste("Severity Distribution by", gsub("_", " ", var_name)),
      x = gsub("_", " ", var_name),
      y = NULL,
      fill = "Outcome"
    ) +
    theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank(),
      legend.position = "right"
    )

  print(p)

  if (exists("print_legend") && exists("legend_maps") && var_name %in% names(legend_maps)) {
    print_legend(var_name)
  }
}

vars_to_plot <- intersect(
  c(
    "time_bucket","weekend","night",
    "WEATHER_1","is_rain",
    "LIGHTING","is_dark",
    "ROAD_SURFACE",
    "INTERSECTION","STATE_HWY_IND","TYPE_OF_COLLISION","PRIMARY_COLL_FACTOR",
    "PCF_VIOL_CATEGORY","pcf_dui","pcf_speed",
    "night_dui","rain_speed","ped_dark"
  ),
  names(switrs_reduced)
)

purrr::walk(vars_to_plot, ~ plot_severity_by_variable(.x))
```

# Correlation Check (numeric predictors)
```{r}
num_vars <- names(switrs_reduced)[sapply(switrs_reduced, is.numeric)]
num_vars <- setdiff(num_vars, c("ACCIDENT_YEAR"))

if (length(num_vars) >= 2) {
  cor_matrix <- cor(switrs_reduced[num_vars], use = "pairwise.complete.obs")
  corrplot(cor_matrix, method = "color", type = "lower", tl.cex = 0.6)

  high_corr <- c()
  moderate_corr <- c()

  for (i in 1:(ncol(cor_matrix) - 1)) {
    for (j in (i + 1):ncol(cor_matrix)) {
      r <- cor_matrix[i, j]
      if (!is.na(r)) {
        pair <- paste(colnames(cor_matrix)[i], colnames(cor_matrix)[j], "r =", round(r, 3))
        if (abs(r) >= 0.8) high_corr <- c(high_corr, pair)
        else if (abs(r) >= 0.5) moderate_corr <- c(moderate_corr, pair)
      }
    }
  }

  cat("High correlations (|r| >= 0.8):\n")
  high_corr
  cat("\nModerate correlations (0.5 <= |r| < 0.8):\n")
  moderate_corr
} else {
  cat("Not enough numeric columns for correlation analysis.\n")
}
```

# Shared imputation
```{r shared_imputation, message=FALSE, warning=FALSE}
if (!exists("switrs_reduced")) stop("switrs_reduced not found.")

switrs_model_ready <- switrs_reduced

if ("SevereBinary" %in% names(switrs_model_ready)) {
  switrs_model_ready <- switrs_model_ready %>%
    dplyr::mutate(
      SevereBinary = factor(SevereBinary, levels = c("Other", "Severe/Fatal"))
    )
}

switrs_model_ready <- switrs_model_ready %>%
  dplyr::mutate(across(where(is.character), as.factor))

factor_cols <- names(switrs_model_ready)[sapply(switrs_model_ready, is.factor)]
factor_cols <- setdiff(factor_cols, "SevereBinary")

if (length(factor_cols) > 0) {
  switrs_model_ready[factor_cols] <- lapply(
    switrs_model_ready[factor_cols],
    function(x) forcats::fct_explicit_na(x, na_level = "Not Stated")
  )
}

numeric_cols <- names(switrs_model_ready)[sapply(switrs_model_ready, function(x) is.numeric(x) || is.integer(x))]
numeric_cols <- setdiff(numeric_cols, "CASE_ID")

shared_numeric_medians <- setNames(vector("list", length(numeric_cols)), numeric_cols)

for (v in numeric_cols) {
  miss_name <- paste0(v, "_miss")

  switrs_model_ready[[miss_name]] <- as.integer(is.na(switrs_model_ready[[v]]))

  med <- suppressWarnings(median(switrs_model_ready[[v]], na.rm = TRUE))
  if (!is.finite(med)) med <- 0

  shared_numeric_medians[[v]] <- med
  switrs_model_ready[[v]][is.na(switrs_model_ready[[v]])] <- med
}

expand_predictors_with_miss <- function(df, predictors) {
  predictors <- intersect(predictors, names(df))

  numeric_base <- predictors[
    sapply(df[predictors], function(x) is.numeric(x) || is.integer(x))
  ]

  unique(c(
    predictors,
    intersect(paste0(numeric_base, "_miss"), names(df))
  ))
}

cat("Shared model-ready dataset created.\n")
cat("Rows =", nrow(switrs_model_ready), "| Columns =", ncol(switrs_model_ready), "\n")
cat("Remaining NA count =", sum(is.na(switrs_model_ready)), "\n")
```

# Shared train/test split
```{r shared_train_test_split, message=FALSE, warning=FALSE}
set.seed(544)

if (!exists("switrs_model_ready")) {
  stop("switrs_model_ready not found. Run the shared_imputation chunk first.")
}

if (!all(c("CASE_ID", "SevereBinary") %in% names(switrs_model_ready))) {
  stop("Need CASE_ID and SevereBinary in switrs_model_ready before creating the shared split.")
}

split_frame <- switrs_model_ready %>%
  dplyr::select(CASE_ID, SevereBinary) %>%
  dplyr::filter(!is.na(CASE_ID), !is.na(SevereBinary)) %>%
  dplyr::distinct(CASE_ID, .keep_all = TRUE) %>%
  dplyr::mutate(
    SevereBinary = factor(SevereBinary, levels = c("Other", "Severe/Fatal"))
  )

test_frac <- 0.30

idx_sev <- which(split_frame$SevereBinary == "Severe/Fatal")
idx_oth <- which(split_frame$SevereBinary == "Other")

shared_test_idx <- c(
  sample(idx_sev, size = floor(length(idx_sev) * test_frac)),
  sample(idx_oth, size = floor(length(idx_oth) * test_frac))
)

test_case_ids  <- split_frame$CASE_ID[shared_test_idx]
train_case_ids <- setdiff(split_frame$CASE_ID, test_case_ids)

train_data <- switrs_model_ready %>%
  dplyr::filter(CASE_ID %in% train_case_ids)

test_data <- switrs_model_ready %>%
  dplyr::filter(CASE_ID %in% test_case_ids)

cat("Shared split created.\n")
cat("Train crashes =", length(train_case_ids), "| Test crashes =", length(test_case_ids), "\n")
cat("Train rows =", nrow(train_data), "| Test rows =", nrow(test_data), "\n")
cat("Shared test fraction =", round(length(test_case_ids) / nrow(split_frame), 4), "\n")
```

# Baseline GLM modeling dataset
```{r}
set.seed(544)

baseline_glm_name <- "Baseline GLM"

if (!exists("leakage_vars")) leakage_vars <- character(0)

candidate_predictors <- intersect(
  c(
    "year",
    "day_group",
    "night",
    "friday_night",
    "weather_group",
    "lighting_group",
    "ROAD_SURFACE",
    "INTERSECTION",
    "STATE_HWY_IND",
    "TYPE_OF_COLLISION",
    "PRIMARY_COLL_FACTOR",
    "school_bus_proxy",
    "pcf_dui",
    "pcf_speed",
    "ALCOHOL_INVOLVED",
    "HIT_AND_RUN",
    "is_ped",
    "BICYCLE_ACCIDENT",
    "MOTORCYCLE_ACCIDENT",
    "TRUCK_ACCIDENT",
    "night_dui",
    "rain_speed",
    "ped_dark",
    "n_parties",
    "any_party_alcohol",
    "any_party_drug",
    "any_party_at_fault",
    "mean_party_age"
  ),
  names(switrs_model_ready)
)

candidate_predictors <- setdiff(candidate_predictors, leakage_vars)

if (length(intersect(candidate_predictors, leakage_vars)) > 0) {
  stop("Leakage detected: a harm/outcome variable is in candidate_predictors.")
}

forbidden_redundant <- c(
  "DAY_OF_WEEK", "weekend", "time_bucket",
  "WEATHER_1", "is_rain",
  "LIGHTING", "is_dark",
  "PCF_VIOL_CATEGORY",
  "PEDESTRIAN_ACCIDENT"
)

if (length(intersect(candidate_predictors, forbidden_redundant)) > 0) {
  stop("Redundant predictors detected in candidate_predictors. Remove parent/raw duplicates.")
}

cat(baseline_glm_name, "predictors:\n")
print(candidate_predictors)

if (!exists("train_data") || !exists("test_data")) {
  stop("train_data/test_data not found. Run shared_imputation and shared_train_test_split first.")
}
if (!exists("expand_predictors_with_miss")) {
  stop("expand_predictors_with_miss() not found. Run the shared_imputation chunk first.")
}

baseline_predictors_final <- expand_predictors_with_miss(train_data, candidate_predictors)

train_set_raw <- train_data %>%
  dplyr::select(all_of(c("CASE_ID", "SevereBinary", baseline_predictors_final)))

test_set_raw <- test_data %>%
  dplyr::select(all_of(c("CASE_ID", "SevereBinary", baseline_predictors_final)))

cat(baseline_glm_name, "raw split sizes: Train =", nrow(train_set_raw), "| Test =", nrow(test_set_raw), "\n")

is_constant_train <- sapply(train_set_raw %>% dplyr::select(-CASE_ID, -SevereBinary), function(x) {
  ux <- unique(x)
  ux <- ux[!is.na(ux)]
  length(ux) <= 1
})

if (any(is_constant_train)) {
  cat("Dropping", baseline_glm_name, "train-constant predictors:\n")
  print(names(is_constant_train)[is_constant_train])

  keep_cols <- c("CASE_ID", "SevereBinary", names(is_constant_train)[!is_constant_train])
  train_set_raw <- train_set_raw %>% dplyr::select(all_of(keep_cols))
  test_set_raw  <- test_set_raw  %>% dplyr::select(all_of(keep_cols))
}

is_perfect_separator <- function(x, y, min_level_n = 30) {
  if (is.character(x) || is.factor(x)) {
    tab <- table(x, y, useNA = "no")
    lvl_n <- rowSums(tab)
    tab2 <- tab[lvl_n >= min_level_n, , drop = FALSE]
    if (nrow(tab2) == 0) return(FALSE)
    any(tab2[, "Other"] == 0 | tab2[, "Severe/Fatal"] == 0)
  } else if (is.numeric(x) || is.integer(x)) {
    ux <- sort(unique(x))
    ux <- ux[!is.na(ux)]
    if (length(ux) <= 1) return(FALSE)
    qs <- unique(stats::quantile(x, probs = seq(0.1, 0.9, by = 0.1), na.rm = TRUE))
    qs <- qs[!is.na(qs)]
    for (cut in qs) {
      left  <- y[x <= cut]
      right <- y[x >  cut]
      if (length(left) >= min_level_n && length(right) >= min_level_n) {
        if (all(left == "Other") || all(left == "Severe/Fatal") ||
            all(right == "Other") || all(right == "Severe/Fatal")) {
          return(TRUE)
        }
      }
    }
    FALSE
  } else {
    FALSE
  }
}

y_train <- train_set_raw$SevereBinary
pred_names_train <- setdiff(names(train_set_raw), c("CASE_ID", "SevereBinary"))

sep_flags <- sapply(pred_names_train, function(v) is_perfect_separator(train_set_raw[[v]], y_train))
bad_vars <- names(sep_flags)[sep_flags]

cat(baseline_glm_name, "predictors dropped for practical separation:", length(bad_vars), "\n")
if (length(bad_vars) > 0) print(bad_vars)

if (length(bad_vars) > 0) {
  train_set_raw <- train_set_raw %>% dplyr::select(-dplyr::all_of(bad_vars))
  test_set_raw  <- test_set_raw  %>% dplyr::select(-dplyr::all_of(bad_vars))
}

train_set <- train_set_raw %>% dplyr::select(-CASE_ID)
test_set  <- test_set_raw  %>% dplyr::select(-CASE_ID)

cat("\nFinal", baseline_glm_name, "columns:\n")
print(names(train_set))

model_df_clean2 <- dplyr::bind_rows(train_set, test_set)

cat(baseline_glm_name, "final split sizes: Train =", nrow(train_set), "| Test =", nrow(test_set), "\n")

baseline_glm_fit <- glm(
  SevereBinary ~ .,
  data = train_set,
  family = binomial(),
  control = glm.control(maxit = 50)
)

glm_fit_final <- baseline_glm_fit

cat(baseline_glm_name, "converged? ", baseline_glm_fit$converged, "\n", sep = "")

baseline_glm_test_phat <- as.numeric(predict(baseline_glm_fit, newdata = test_set, type = "response"))
test_phat <- baseline_glm_test_phat
test_y <- test_set$SevereBinary

if (!requireNamespace("pROC", quietly = TRUE)) install.packages("pROC")
if (requireNamespace("pROC", quietly = TRUE)) {
  roc_obj <- pROC::roc(test_y, baseline_glm_test_phat, levels = c("Other","Severe/Fatal"), quiet = TRUE)
  cat(baseline_glm_name, "TEST AUC:", round(as.numeric(pROC::auc(roc_obj)), 4), "\n")
}

pred_05 <- factor(ifelse(baseline_glm_test_phat >= 0.5, "Severe/Fatal", "Other"), levels = levels(test_y))
cat("\n", baseline_glm_name, "confusion matrix (cutoff=0.50):\n", sep = "")
print(table(Pred = pred_05, Truth = test_y))

prev <- mean(train_set$SevereBinary == "Severe/Fatal")
pred_prev <- factor(ifelse(baseline_glm_test_phat >= prev, "Severe/Fatal", "Other"), levels = levels(test_y))
cat("\n", baseline_glm_name, "confusion matrix (cutoff=prevalence=", round(prev, 4), "): \n", sep = "")
print(table(Pred = pred_prev, Truth = test_y))

coefs <- summary(baseline_glm_fit)$coefficients
or_tbl <- data.frame(
  term = rownames(coefs),
  estimate = coefs[, "Estimate"],
  std.error = coefs[, "Std. Error"],
  p.value = coefs[, "Pr(>|z|)"],
  stringsAsFactors = FALSE
) %>%
  dplyr::filter(term != "(Intercept)") %>%
  dplyr::filter(is.finite(estimate), is.finite(std.error)) %>%
  dplyr::mutate(
    OR = exp(estimate),
    conf.low = exp(estimate - 1.96 * std.error),
    conf.high = exp(estimate + 1.96 * std.error),
    log_or = abs(estimate)
  ) %>%
  dplyr::arrange(dplyr::desc(log_or)) %>%
  dplyr::select(term, OR, conf.low, conf.high, p.value)

cat("\nTop 20", baseline_glm_name, "odds ratios by |log(OR)| (FAST Wald CI, train fit):\n")
print(head(or_tbl, 20))
```

# Coefficient analysis
```{r report_glm_coefficient_analysis, message=FALSE, warning=FALSE}
if (!exists("or_tbl")) stop("or_tbl not found. Run the baseline GLM chunk first.")

coef_report_tbl <- or_tbl %>%
  dplyr::filter(is.finite(OR), is.finite(conf.low), is.finite(conf.high), is.finite(p.value)) %>%
  dplyr::mutate(
    effect_direction = dplyr::if_else(
      OR > 1,
      "Higher odds of severe/fatal crash",
      "Lower odds of severe/fatal crash"
    ),
    distance_from_null = abs(log(OR))
  ) %>%
  dplyr::filter(!grepl("_miss$", term)) %>%
  dplyr::filter(!grepl("Not Stated", term, fixed = TRUE)) %>%
  dplyr::arrange(p.value, dplyr::desc(distance_from_null))

coef_increase_tbl <- coef_report_tbl %>%
  dplyr::filter(OR > 1) %>%
  dplyr::slice_head(n = 10) %>%
  dplyr::mutate(
    OR = round(OR, 3),
    conf.low = round(conf.low, 3),
    conf.high = round(conf.high, 3),
    p.value = signif(p.value, 3)
  ) %>%
  dplyr::select(term, OR, conf.low, conf.high, p.value, effect_direction)

coef_decrease_tbl <- coef_report_tbl %>%
  dplyr::filter(OR < 1) %>%
  dplyr::slice_head(n = 10) %>%
  dplyr::mutate(
    OR = round(OR, 3),
    conf.low = round(conf.low, 3),
    conf.high = round(conf.high, 3),
    p.value = signif(p.value, 3)
  ) %>%
  dplyr::select(term, OR, conf.low, conf.high, p.value, effect_direction)

cat("Top GLM terms associated with HIGHER severe/fatal odds:\n")
print(coef_increase_tbl)
cat("\nTop GLM terms associated with LOWER severe/fatal odds:\n")
print(coef_decrease_tbl)
```

# What-if / one-at-a-time probability changes (GLM)
```{r}
if (!exists("glm_fit_final")) stop("glm_fit_final not found. Run the baseline GLM chunk first.")
if (!exists("train_set")) stop("train_set not found. Run the baseline GLM chunk first.")

model_vars <- setdiff(names(train_set), "SevereBinary")
X_ref <- train_set[, model_vars, drop = FALSE]
model_xlev <- glm_fit_final$xlevels

mode_level <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_character_)
  names(which.max(table(x)))[1]
}

safe_predict_prob <- function(model, newdata) {
  as.numeric(suppressWarnings(predict(model, newdata = newdata, type = "response")))
}

baseline_row <- as.data.frame(lapply(names(X_ref), function(v) {
  col <- X_ref[[v]]

  if (v %in% names(model_xlev)) {
    allowed <- model_xlev[[v]]
    obs_mode <- mode_level(as.character(col))

    if (is.na(obs_mode) || !(obs_mode %in% allowed)) {
      freq <- sort(table(as.character(col)), decreasing = TRUE)
      cand <- intersect(names(freq), allowed)
      obs_mode <- if (length(cand) > 0) cand[1] else allowed[1]
    }

    return(factor(obs_mode, levels = allowed))
  }

  if (is.integer(col)) {
    return(as.integer(round(stats::median(col, na.rm = TRUE))))
  }

  if (is.numeric(col)) {
    return(as.numeric(stats::median(col, na.rm = TRUE)))
  }

  mode_level(as.character(col))
}), stringsAsFactors = FALSE)

names(baseline_row) <- names(X_ref)

for (v in names(X_ref)) {
  if (is.integer(X_ref[[v]]) && !(v %in% names(model_xlev))) {
    baseline_row[[v]] <- as.integer(round(as.numeric(baseline_row[[v]])))
  }
}

p0 <- safe_predict_prob(glm_fit_final, baseline_row)

cat("Baseline predicted probability of Severe/Fatal crash:\n")
print(round(p0, 4))

what_if <- function(var_name, values = NULL,
                    model = glm_fit_final, base = baseline_row, ref = X_ref, xlev = model_xlev) {

  if (!var_name %in% names(base)) stop("Variable not in model predictors: ", var_name)

  if (var_name %in% names(xlev)) {
    allowed <- xlev[[var_name]]

    if (is.null(values)) values <- allowed
    values <- as.character(values)
    values <- values[values %in% allowed]

    if (length(values) == 0) stop("No provided values are valid levels for ", var_name)

    out <- lapply(values, function(val) {
      nd <- base
      nd[[var_name]] <- factor(val, levels = allowed)
      p <- safe_predict_prob(model, nd)

      data.frame(
        variable = var_name,
        value = val,
        p_severe = p,
        stringsAsFactors = FALSE
      )
    })

    out <- dplyr::bind_rows(out) %>%
      dplyr::mutate(
        p_baseline = p0,
        delta_pp = 100 * (p_severe - p_baseline),
        baseline_value = as.character(base[[var_name]])
      ) %>%
      dplyr::arrange(dplyr::desc(p_severe))

    return(out)
  }

  xref <- ref[[var_name]]

  if (is.null(values)) {
    qs <- stats::quantile(xref, probs = c(0.10, 0.50, 0.90), na.rm = TRUE, names = FALSE)
    values <- as.numeric(qs)
  }

  out <- lapply(values, function(val) {
    nd <- base

    if (is.integer(xref)) {
      nd[[var_name]] <- as.integer(val)
    } else {
      nd[[var_name]] <- as.numeric(val)
    }

    p <- safe_predict_prob(model, nd)

    data.frame(
      variable = var_name,
      value = as.character(val),
      p_severe = p,
      stringsAsFactors = FALSE
    )
  })

  dplyr::bind_rows(out) %>%
    dplyr::mutate(
      p_baseline = p0,
      delta_pp = 100 * (p_severe - p_baseline),
      baseline_value = as.character(base[[var_name]])
    ) %>%
    dplyr::arrange(dplyr::desc(p_severe))
}

whatif_results <- list()

if ("day_group" %in% names(baseline_row)) {
  vals <- intersect(c("Mon-Thu", "Friday", "Saturday", "Sunday"), model_xlev$day_group)
  whatif_results$day_group <- what_if("day_group", vals)
}

if ("friday_night" %in% names(baseline_row)) {
  vals <- intersect(c("No", "Yes"), model_xlev$friday_night)
  whatif_results$friday_night <- what_if("friday_night", vals)
}

if ("night" %in% names(baseline_row)) {
  vals <- intersect(c("No", "Yes"), model_xlev$night)
  whatif_results$night <- what_if("night", vals)
}

if ("lighting_group" %in% names(baseline_row)) {
  vals <- intersect(c("Daylight", "Dusk/Dawn", "Dark"), model_xlev$lighting_group)
  whatif_results$lighting_group <- what_if("lighting_group", vals)
}

if ("weather_group" %in% names(baseline_row)) {
  vals <- intersect(c("Clear", "Rain", "Other Weather"), model_xlev$weather_group)
  whatif_results$weather_group <- what_if("weather_group", vals)
}

if ("school_bus_proxy" %in% names(baseline_row)) {
  vals <- intersect(c("No", "Yes"), model_xlev$school_bus_proxy)
  whatif_results$school_bus_proxy <- what_if("school_bus_proxy", vals)
}

if ("INTERSECTION" %in% names(baseline_row)) {
  vals <- intersect(c("N", "Y", "Not Stated"), model_xlev$INTERSECTION)
  whatif_results$INTERSECTION <- what_if("INTERSECTION", vals)
}

if ("STATE_HWY_IND" %in% names(baseline_row)) {
  vals <- intersect(c("N", "Y", "Not Stated"), model_xlev$STATE_HWY_IND)
  whatif_results$STATE_HWY_IND <- what_if("STATE_HWY_IND", vals)
}

if ("PRIMARY_COLL_FACTOR" %in% names(baseline_row)) {
  vals <- model_xlev$PRIMARY_COLL_FACTOR
  whatif_results$PRIMARY_COLL_FACTOR <- what_if("PRIMARY_COLL_FACTOR", vals)
}

if ("pcf_dui" %in% names(baseline_row)) {
  vals <- intersect(c("No", "Yes"), model_xlev$pcf_dui)
  whatif_results$pcf_dui <- what_if("pcf_dui", vals)
}

if ("pcf_speed" %in% names(baseline_row)) {
  vals <- intersect(c("No", "Yes"), model_xlev$pcf_speed)
  whatif_results$pcf_speed <- what_if("pcf_speed", vals)
}

if ("ALCOHOL_INVOLVED" %in% names(baseline_row)) {
  vals <- intersect(c("N", "Y", "Not Stated"), model_xlev$ALCOHOL_INVOLVED)
  whatif_results$ALCOHOL_INVOLVED <- what_if("ALCOHOL_INVOLVED", vals)
}

if ("is_ped" %in% names(baseline_row)) {
  vals <- intersect(c("No", "Yes"), model_xlev$is_ped)
  whatif_results$is_ped <- what_if("is_ped", vals)
}

if ("HIT_AND_RUN" %in% names(baseline_row)) {
  vals <- intersect(c("N", "M", "F", "Not Stated"), model_xlev$HIT_AND_RUN)
  whatif_results$HIT_AND_RUN <- what_if("HIT_AND_RUN", vals)
}

if ("ROAD_SURFACE" %in% names(baseline_row)) {
  vals <- model_xlev$ROAD_SURFACE
  whatif_results$ROAD_SURFACE <- what_if("ROAD_SURFACE", vals)
}

if ("TYPE_OF_COLLISION" %in% names(baseline_row)) {
  vals <- model_xlev$TYPE_OF_COLLISION
  whatif_results$TYPE_OF_COLLISION <- what_if("TYPE_OF_COLLISION", vals)
}

cat("\n================ WHAT-IF RESULTS (GLM) ================\n")
for (nm in names(whatif_results)) {
  cat("\n------------------------------------------------------\n")
  cat("Variable:", nm, "\n")
  cat("------------------------------------------------------\n")
  print(whatif_results[[nm]])
}
```

# Penalized logistic regression: GLMNET model
```{r glmnet_model, message=FALSE, warning=FALSE}
lasso_glmnet_name <- "LASSO GLMNET"

if (!requireNamespace("glmnet", quietly = TRUE)) install.packages("glmnet")
library(glmnet)

if (!exists("train_set") || !exists("test_set")) {
  stop("train_set/test_set not found. Run the Baseline GLM chunk first.")
}
if (!("SevereBinary" %in% names(train_set))) stop("SevereBinary not found in train_set.")

train_set$.split <- "train"
test_set$.split  <- "test"
combo <- dplyr::bind_rows(train_set, test_set)

train_idx <- combo$.split == "train"
test_idx  <- combo$.split == "test"

combo_mm <- combo %>% dplyr::select(-.split)

X_all <- model.matrix(SevereBinary ~ ., data = combo_mm)[, -1, drop = FALSE]

x_train <- X_all[train_idx, , drop = FALSE]
x_test  <- X_all[test_idx,  , drop = FALSE]

y_train <- as.integer(combo$SevereBinary[train_idx] == "Severe/Fatal")
y_test  <- as.integer(combo$SevereBinary[test_idx]  == "Severe/Fatal")

train_set$.split <- NULL
test_set$.split  <- NULL
combo$.split     <- NULL

set.seed(544)
lasso_glmnet_cv_fit <- cv.glmnet(
  x = x_train,
  y = y_train,
  family = "binomial",
  alpha = 1,
  nfolds = 5,
  type.measure = "auc"
)

cv_fit <- lasso_glmnet_cv_fit

cat(lasso_glmnet_name, "best lambda (min):", lasso_glmnet_cv_fit$lambda.min, "\n")
cat(lasso_glmnet_name, "1-SE lambda:", lasso_glmnet_cv_fit$lambda.1se, "\n")

plot(lasso_glmnet_cv_fit)

lasso_glmnet_test_phat <- as.numeric(
  predict(lasso_glmnet_cv_fit, newx = x_test, s = "lambda.1se", type = "response")
)
p_test <- lasso_glmnet_test_phat

if (!requireNamespace("pROC", quietly = TRUE)) install.packages("pROC")
auc_glmnet <- NA_real_
if (requireNamespace("pROC", quietly = TRUE)) {
  roc_obj <- pROC::roc(response = y_test, predictor = lasso_glmnet_test_phat, quiet = TRUE)
  auc_glmnet <- as.numeric(pROC::auc(roc_obj))
  cat(lasso_glmnet_name, "TEST AUC:", round(auc_glmnet, 4), "\n")
}

pred_05 <- ifelse(lasso_glmnet_test_phat >= 0.5, 1, 0)
cat("\n", lasso_glmnet_name, "confusion matrix (cutoff=0.50):\n", sep = "")
print(table(Pred = pred_05, Truth = y_test))

prev <- mean(y_train == 1)
pred_prev <- ifelse(lasso_glmnet_test_phat >= prev, 1, 0)
cat("\n", lasso_glmnet_name, "confusion matrix (cutoff=prevalence=", round(prev, 4), "): \n", sep = "")
print(table(Pred = pred_prev, Truth = y_test))

coef_mat <- as.matrix(coef(lasso_glmnet_cv_fit, s = "lambda.1se"))
coef_df <- data.frame(
  term = rownames(coef_mat),
  coef = as.numeric(coef_mat[, 1]),
  row.names = NULL
) %>%
  dplyr::filter(term != "(Intercept)", coef != 0) %>%
  dplyr::arrange(dplyr::desc(abs(coef)))

cat("\nTop 20 non-zero coefficients by |coef| (", lasso_glmnet_name, ", lambda.1se):\n", sep = "")
print(head(coef_df, 20))
```

# Feature selection(glmnet LASSO)
```{r feature_selection_summary}
if (!exists("cv_fit")) stop("cv_fit not found. Run glmnet_model chunk first.")

coef_mat <- as.matrix(coef(cv_fit, s = "lambda.1se"))
coef_df <- data.frame(
  term = rownames(coef_mat),
  coef = as.numeric(coef_mat[, 1]),
  stringsAsFactors = FALSE
) %>%
  dplyr::filter(term != "(Intercept)", coef != 0) %>%
  dplyr::arrange(dplyr::desc(abs(coef)))

cat("Number of selected (non-zero) one-hot terms:", nrow(coef_df), "\n\n")
cat("Top 30 selected terms by |coef|:\n")
print(head(coef_df, 30))

candidate_base_vars <- setdiff(names(train_set), c("SevereBinary"))
candidate_base_vars <- candidate_base_vars[order(nchar(candidate_base_vars), decreasing = TRUE)]

base_var_from_term <- function(term, vars) {
  hits <- vars[startsWith(term, vars)]
  if (length(hits) == 0) return(term)
  hits[1]
}

coef_df$base_var <- vapply(coef_df$term, base_var_from_term, character(1), vars = candidate_base_vars)

selected_vars <- coef_df %>%
  dplyr::count(base_var, sort = TRUE)

cat("\nSelected base variables (ranked by how many active dummy terms they contribute):\n")
print(selected_vars)
```

# GLM what-if analysis for LASSO-selected features
```{r}
if (!exists("what_if")) stop("what_if() not found. Run the earlier 'What-if / one-at-a-time probability changes (GLM)' chunk first.")
if (!exists("baseline_row")) stop("baseline_row not found. Run the earlier GLM what-if chunk first.")
if (!exists("selected_vars")) stop("selected_vars not found. Run feature_selection_summary first.")
if (!exists("train_set")) stop("train_set not found. Run the baseline modeling chunk first.")
if (!exists("glm_fit_final")) stop("glm_fit_final not found. Run the baseline modeling chunk first.")

top_k <- 10

important_vars <- selected_vars %>%
  dplyr::arrange(dplyr::desc(n)) %>%
  dplyr::pull(base_var) %>%
  unique()

important_vars <- important_vars[!grepl("_miss$", important_vars)]
important_vars <- important_vars[important_vars %in% names(baseline_row)]
important_vars <- head(important_vars, top_k)

cat("Most important variables (top ", top_k, "): \n", sep = "")
print(important_vars)

pick_levels <- function(var_name, ref, xlev, base_row, max_show = 8) {
  allowed <- xlev[[var_name]]
  if (length(allowed) <= max_show) return(allowed)

  base_val <- as.character(base_row[[var_name]])
  freq <- sort(table(as.character(ref[[var_name]])), decreasing = TRUE)
  top_levs <- intersect(names(freq), allowed)

  vals <- unique(c(base_val, head(top_levs, max_show)))
  vals[vals %in% allowed]
}

whatif_tables <- list()

for (v in important_vars) {
  if (v %in% names(glm_fit_final$xlevels)) {
    vals <- pick_levels(v, train_set, glm_fit_final$xlevels, baseline_row, max_show = 8)
    whatif_tables[[v]] <- what_if(v, values = vals)
  } else {
    whatif_tables[[v]] <- what_if(v)
  }
}

whatif_summary <- purrr::imap_dfr(whatif_tables, function(tbl, varname) {
  best <- tbl[which.max(tbl$p_severe), , drop = FALSE]
  worst <- tbl[which.min(tbl$p_severe), , drop = FALSE]

  data.frame(
    variable = varname,
    baseline_value = best$baseline_value[1],
    baseline_p = best$p_baseline[1],
    best_value = best$value[1],
    best_p = best$p_severe[1],
    best_delta_pp = best$delta_pp[1],
    worst_value = worst$value[1],
    worst_p = worst$p_severe[1],
    worst_delta_pp = worst$delta_pp[1],
    swing_pp = best$delta_pp[1] - worst$delta_pp[1],
    stringsAsFactors = FALSE
  )
}) %>%
  dplyr::arrange(dplyr::desc(abs(swing_pp)))

cat("\n================ WHAT-IF SUMMARY (top features) ================\n")
print(whatif_summary)
```

# report assumption checks
```{r, message=FALSE, warning=FALSE}
raw_df_for_report <- if (exists("switrs")) switrs else switrs_reduced

sample_for_shapiro <- function(x, max_n = 5000) {
  x <- x[is.finite(x)]
  if (length(x) > max_n) {
    set.seed(544)
    x <- sample(x, max_n)
  }
  x
}

sample_skewness <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) < 3) return(NA_real_)
  s <- stats::sd(x)
  if (!is.finite(s) || s == 0) return(0)
  mean(((x - mean(x)) / s)^3)
}

assumption_num_vars <- intersect(
  c("n_parties", "n_victims", "mean_party_age", "mean_victim_age", "min_victim_age", "max_victim_age"),
  names(switrs_reduced)
)

independence_tbl <- tibble::tibble(
  rows = nrow(raw_df_for_report),
  unique_case_id = if ("CASE_ID" %in% names(raw_df_for_report)) dplyr::n_distinct(raw_df_for_report$CASE_ID) else NA_integer_,
  duplicated_case_id = if ("CASE_ID" %in% names(raw_df_for_report)) sum(duplicated(raw_df_for_report$CASE_ID)) else NA_integer_,
  comment = if ("CASE_ID" %in% names(raw_df_for_report)) {
    ifelse(
      sum(duplicated(raw_df_for_report$CASE_ID)) == 0,
      "No duplicated CASE_ID values in the crash-level merged file.",
      "Duplicate CASE_ID values found; row-level independence is weaker than intended."
    )
  } else {
    "CASE_ID not available to check row-level independence directly."
  }
)

cat("Independence check:\n")
print(independence_tbl)

if (length(assumption_num_vars) == 0) {
  cat("\nNo numeric variables from the report assumption list were found in switrs_reduced.\n")
} else {
  assumption_tbl <- purrr::map_dfr(assumption_num_vars, function(v) {
    df <- switrs_reduced %>%
      dplyr::select(SevereBinary, value = dplyr::all_of(v)) %>%
      dplyr::filter(!is.na(SevereBinary), is.finite(value))

    x_other <- sample_for_shapiro(df$value[df$SevereBinary == "Other"])
    x_sev   <- sample_for_shapiro(df$value[df$SevereBinary == "Severe/Fatal"])

    shapiro_other <- tryCatch(
      if (length(x_other) >= 3) shapiro.test(x_other)$p.value else NA_real_,
      error = function(e) NA_real_
    )

    shapiro_sev <- tryCatch(
      if (length(x_sev) >= 3) shapiro.test(x_sev)$p.value else NA_real_,
      error = function(e) NA_real_
    )

    fligner_p <- tryCatch(
      fligner.test(value ~ SevereBinary, data = df)$p.value,
      error = function(e) NA_real_
    )

    tibble::tibble(
      variable = v,
      n_other = sum(df$SevereBinary == "Other"),
      n_severe = sum(df$SevereBinary == "Severe/Fatal"),
      skew_other = round(sample_skewness(x_other), 3),
      skew_severe = round(sample_skewness(x_sev), 3),
      shapiro_p_other = signif(shapiro_other, 3),
      shapiro_p_severe = signif(shapiro_sev, 3),
      fligner_p = signif(fligner_p, 3),
      recommended_test = "Welch t-test primary; Wilcoxon sensitivity check"
    )
  })

  cat("\nQuick numeric-assumption screen:\n")
  print(assumption_tbl)
}

cat("\nCategorical-test assumption handling already built into the notebook:\n")
cat("- Chi-square expected counts are checked automatically.\n")
cat("- When expected counts are small, the code already switches to simulated chi-square p-values.\n")
cat("- For numeric predictors, Welch t-tests are paired with Wilcoxon tests for robustness.\n")
```

# UNIVARIATE HYPOTHESIS TESTING HELPERS (t-test / Wilcoxon / prop-test / chi-square)
```{r}
library(dplyr)
library(purrr)
library(broom)
library(tibble)

cohen_d <- function(x, g) {
  x0 <- x[g == "Other"]; x1 <- x[g == "Severe/Fatal"]
  x0 <- x0[is.finite(x0)]; x1 <- x1[is.finite(x1)]
  if (length(x0) < 2 || length(x1) < 2) return(NA_real_)
  s0 <- sd(x0); s1 <- sd(x1)
  sp <- sqrt(((length(x0)-1)*s0^2 + (length(x1)-1)*s1^2) / (length(x0)+length(x1)-2))
  if (!is.finite(sp) || sp == 0) return(NA_real_)
  (mean(x1) - mean(x0)) / sp
}

cramers_v <- function(tab, chi2_stat) {
  n <- sum(tab)
  r <- nrow(tab); c <- ncol(tab)
  if (n == 0 || min(r-1, c-1) <= 0) return(NA_real_)
  sqrt(as.numeric(chi2_stat) / (n * min(r-1, c-1)))
}

run_univariate_test <- function(data, var,
                                outcome = "SevereBinary",
                                alpha = 0.05,
                                max_levels = 60,
                                chisq_B = 2000) {
  if (!all(c(var, outcome) %in% names(data))) return(NULL)

  df <- data %>%
    dplyr::select(all_of(c(var, outcome))) %>%
    dplyr::filter(!is.na(.data[[outcome]]))

  df[[outcome]] <- factor(df[[outcome]], levels = c("Other", "Severe/Fatal"))
  df <- df %>% dplyr::filter(!is.na(.data[[outcome]]))

  x <- df[[var]]

  if (is.numeric(x) || is.integer(x)) {
    ux <- unique(x[is.finite(x)])
    if (length(ux) <= 2) x <- as.factor(x)
  }

  if (is.numeric(x) || is.integer(x)) {
    df2 <- df %>% mutate(x = as.numeric(.data[[var]])) %>% filter(is.finite(x))
    if (nrow(df2) < 10) return(tibble(
      variable = var, type = "numeric", n_used = nrow(df2),
      test = "SKIP (too few non-missing)", statistic = NA_real_,
      p_value = NA_real_, effect = NA_real_, effect_label = NA_character_
    ))

    df2$g <- df2[[outcome]]

    tt_welch <- tryCatch(t.test(x ~ g, data = df2, var.equal = FALSE), error = function(e) NULL)
    tt_student <- tryCatch(t.test(x ~ g, data = df2, var.equal = TRUE), error = function(e) NULL)
    wt <- tryCatch(wilcox.test(x ~ g, data = df2, exact = FALSE), error = function(e) NULL)

    m_other <- mean(df2$x[df2[[outcome]] == "Other"], na.rm = TRUE)
    m_sev   <- mean(df2$x[df2[[outcome]] == "Severe/Fatal"], na.rm = TRUE)
    d <- cohen_d(df2$x, df2[[outcome]])

    tibble(
      variable = var,
      type = "numeric",
      n_used = nrow(df2),
      test = "t-test (Welch) + t-test (Student) + Wilcoxon",
      statistic = if (!is.null(tt_welch)) unname(tt_welch$statistic) else NA_real_,
      p_value = if (!is.null(tt_welch)) tt_welch$p.value else NA_real_,
      p_value_student = if (!is.null(tt_student)) tt_student$p.value else NA_real_,
      p_value_wilcox  = if (!is.null(wt)) wt$p.value else NA_real_,
      mean_other = m_other,
      mean_severe = m_sev,
      mean_diff = m_sev - m_other,
      effect = d,
      effect_label = "Cohen_d (Severe - Other)"
    )
  }

  x <- as.character(x)
  x[x == ""] <- NA_character_
  df2 <- df %>% mutate(x = x) %>% filter(!is.na(x))

  if (nrow(df2) < 10) return(tibble(
    variable = var, type = "categorical", n_used = nrow(df2),
    test = "SKIP (too few non-missing)", statistic = NA_real_,
    p_value = NA_real_, effect = NA_real_, effect_label = NA_character_
  ))

  k <- length(unique(df2$x))
  if (k > max_levels) {
    return(tibble(
      variable = var, type = "categorical", n_used = nrow(df2),
      test = paste0("SKIP (high cardinality: ", k, " levels > ", max_levels, ")"),
      statistic = NA_real_, p_value = NA_real_, effect = NA_real_, effect_label = NA_character_
    ))
  }

  tab <- table(df2$x, df2[[outcome]])

  if (nrow(tab) == 2) {
    x_succ <- tab[, "Severe/Fatal"]
    n_tot  <- rowSums(tab)
    pt <- suppressWarnings(prop.test(x = x_succ, n = n_tot, correct = FALSE))

    p1 <- x_succ[1] / n_tot[1]
    p2 <- x_succ[2] / n_tot[2]
    diff <- p1 - p2
    v <- cramers_v(tab, pt$statistic)

    tibble(
      variable = var,
      type = "categorical (2 levels)",
      n_used = sum(tab),
      test = "Two-proportion test (prop.test) (≈ Z-test) + 2x2 chi-square equivalent",
      statistic = unname(pt$statistic),
      p_value = pt$p.value,
      level1 = rownames(tab)[1],
      level2 = rownames(tab)[2],
      p_severe_level1 = p1,
      p_severe_level2 = p2,
      risk_diff = diff,
      effect = v,
      effect_label = "CramersV (2x2)"
    )
  } else {
    cs <- suppressWarnings(chisq.test(tab))
    use_sim <- any(cs$expected < 5)
    cs2 <- if (use_sim) suppressWarnings(chisq.test(tab, simulate.p.value = TRUE, B = chisq_B)) else cs
    v <- cramers_v(tab, cs2$statistic)

    tibble(
      variable = var,
      type = paste0("categorical (", nrow(tab), " levels)"),
      n_used = sum(tab),
      test = if (use_sim) paste0("Chi-square independence (sim p, B=", chisq_B, ")") else "Chi-square independence",
      statistic = unname(cs2$statistic),
      p_value = cs2$p.value,
      df = unname(cs2$parameter),
      effect = v,
      effect_label = "CramersV"
    )
  }
}

run_univariate_suite <- function(data, vars, outcome = "SevereBinary") {
  res <- purrr::map_dfr(vars, ~ run_univariate_test(data, .x, outcome = outcome))
  if (nrow(res) == 0) return(res)
  res %>%
    mutate(p_adj_fdr = p.adjust(p_value, method = "BH")) %>%
    arrange(p_adj_fdr, p_value)
}
```

# UNIVARIATE TEST BATTERY: Selected features + (optional) all candidate predictors
```{r}
if (!exists("switrs_reduced")) stop("switrs_reduced not found. Run our cleaning/feature engineering first.")
if (!("SevereBinary" %in% names(switrs_reduced))) stop("SevereBinary not found in switrs_reduced.")

if (exists("selected_vars") && "base_var" %in% names(selected_vars)) {
  selected_base_vars <- unique(selected_vars$base_var)
} else if (exists("train_set")) {
  selected_base_vars <- setdiff(names(train_set), c("SevereBinary"))
} else {
  stop("Neither selected_vars nor train_set exists. Run glmnet feature selection (or baseline model) first.")
}

if (exists("leakage_vars")) selected_base_vars <- setdiff(selected_base_vars, leakage_vars)
selected_base_vars <- intersect(selected_base_vars, names(switrs_reduced))

cat("Selected base variables to test:", length(selected_base_vars), "\n")

tests_selected <- run_univariate_suite(switrs_reduced, selected_base_vars, outcome = "SevereBinary")

cat("\nTop 30 (FDR-adjusted) significant selected variables:\n")
print(head(tests_selected, 30))

if (exists("candidate_predictors")) {
  all_vars <- candidate_predictors
  if (exists("leakage_vars")) all_vars <- setdiff(all_vars, leakage_vars)
  all_vars <- intersect(all_vars, names(switrs_reduced))

  cat("\nAll candidate variables to test:", length(all_vars), "\n")
  tests_all <- run_univariate_suite(switrs_reduced, all_vars, outcome = "SevereBinary")

  cat("\nTop 30 (FDR-adjusted) significant variables among ALL candidates:\n")
  print(head(tests_all, 30))
} else {
  cat("\nSkipping ALL-variable tests: candidate_predictors not found (run baseline modeling chunk first).\n")
}
```

# Re-run tests after removing "Not Stated" categories (sensitivity check)
```{r}
switrs_noNS <- switrs_reduced %>%
  dplyr::mutate(across(where(is.character), ~ dplyr::na_if(.x, "Not Stated"))) %>%
  dplyr::mutate(across(where(is.factor), ~ {
    x <- as.character(.x)
    x[x == "Not Stated"] <- NA_character_
    factor(x)
  }))

tests_selected_noNS <- run_univariate_suite(switrs_noNS, selected_base_vars, outcome = "SevereBinary")

cat("\nTop 30 selected vars (NO Not Stated, FDR-adjusted):\n")
print(head(tests_selected_noNS, 30))
```

# Print contingency tables for top categorical predictors (by FDR)
```{r}
print_top_tables <- function(test_tbl, data, top_k = 8) {
  top_vars <- test_tbl %>%
    dplyr::filter(grepl("^categorical", type)) %>%
    dplyr::arrange(p_adj_fdr) %>%
    dplyr::pull(variable) %>%
    unique() %>%
    head(top_k)

  for (v in top_vars) {
    cat("\n=============================\n")
    cat("Contingency table for:", v, "\n")
    cat("=============================\n")

    tab <- table(
      normalize_not_stated(data[[v]], blank_to_notstated = TRUE),
      data$SevereBinary,
      useNA = "no"
    )

    if (exists("legend_maps") && v %in% names(legend_maps) && exists("lookup_meaning")) {
      rn <- rownames(tab)
      rownames(tab) <- lookup_meaning(rn, legend_maps[[v]])
    }

    print(addmargins(tab))
  }
}

if (exists("tests_selected") && nrow(tests_selected) > 0) {
  print_top_tables(tests_selected, switrs_reduced, top_k = 8)
}
```

# TARGETED SCENARIO MODEL: interaction-focused policy model
```{r}
set.seed(544)

targeted_scenario_glm_name <- "Targeted Scenario GLM"

scenario_predictors <- intersect(
  c(
    "day_group",
    "night",
    "is_ped",
    "INTERSECTION",
    "STATE_HWY_IND",
    "lighting_group",
    "pcf_dui",
    "friday_night",
    "ped_dark",
    "intersection_ped",
    "highway_night",
    "night_dui"
  ),
  names(switrs_model_ready)
)

cat(targeted_scenario_glm_name, "predictors:\n")
print(scenario_predictors)

if (!exists("train_data") || !exists("test_data")) {
  stop("train_data/test_data not found. Run shared_imputation and shared_train_test_split first.")
}
if (!exists("expand_predictors_with_miss")) {
  stop("expand_predictors_with_miss() not found. Run the shared_imputation chunk first.")
}

scenario_predictors_final <- expand_predictors_with_miss(train_data, scenario_predictors)

train_scen_raw <- train_data %>%
  dplyr::select(all_of(c("CASE_ID", "SevereBinary", scenario_predictors_final)))

test_scen_raw <- test_data %>%
  dplyr::select(all_of(c("CASE_ID", "SevereBinary", scenario_predictors_final)))

cat(targeted_scenario_glm_name, "raw split sizes: Train =", nrow(train_scen_raw), "| Test =", nrow(test_scen_raw), "\n")

is_constant_scen <- sapply(train_scen_raw %>% dplyr::select(-CASE_ID, -SevereBinary), function(x) {
  ux <- unique(x)
  ux <- ux[!is.na(ux)]
  length(ux) <= 1
})

if (any(is_constant_scen)) {
  cat("Dropping", targeted_scenario_glm_name, "train-constant predictors:\n")
  print(names(is_constant_scen)[is_constant_scen])

  keep_cols_scen <- c("CASE_ID", "SevereBinary", names(is_constant_scen)[!is_constant_scen])
  train_scen_raw <- train_scen_raw %>% dplyr::select(all_of(keep_cols_scen))
  test_scen_raw  <- test_scen_raw  %>% dplyr::select(all_of(keep_cols_scen))
}

if (!exists("is_perfect_separator")) {
  is_perfect_separator <- function(x, y, min_level_n = 30) {
    if (is.character(x) || is.factor(x)) {
      tab <- table(x, y, useNA = "no")
      lvl_n <- rowSums(tab)
      tab2 <- tab[lvl_n >= min_level_n, , drop = FALSE]
      if (nrow(tab2) == 0) return(FALSE)
      any(tab2[, "Other"] == 0 | tab2[, "Severe/Fatal"] == 0)
    } else if (is.numeric(x) || is.integer(x)) {
      ux <- sort(unique(x))
      ux <- ux[!is.na(ux)]
      if (length(ux) <= 1) return(FALSE)
      qs <- unique(stats::quantile(x, probs = seq(0.1, 0.9, by = 0.1), na.rm = TRUE))
      qs <- qs[!is.na(qs)]
      for (cut in qs) {
        left  <- y[x <= cut]
        right <- y[x >  cut]
        if (length(left) >= min_level_n && length(right) >= min_level_n) {
          if (all(left == "Other") || all(left == "Severe/Fatal") ||
              all(right == "Other") || all(right == "Severe/Fatal")) {
            return(TRUE)
          }
        }
      }
      FALSE
    } else {
      FALSE
    }
  }
}

y_train_scen <- train_scen_raw$SevereBinary
pred_names_scen <- setdiff(names(train_scen_raw), c("CASE_ID", "SevereBinary"))

sep_flags_scen <- sapply(pred_names_scen, function(v) {
  is_perfect_separator(train_scen_raw[[v]], y_train_scen)
})
bad_vars_scen <- names(sep_flags_scen)[sep_flags_scen]

cat(targeted_scenario_glm_name, "predictors dropped for practical separation:", length(bad_vars_scen), "\n")
if (length(bad_vars_scen) > 0) print(bad_vars_scen)

if (length(bad_vars_scen) > 0) {
  train_scen_raw <- train_scen_raw %>% dplyr::select(-dplyr::all_of(bad_vars_scen))
  test_scen_raw  <- test_scen_raw  %>% dplyr::select(-dplyr::all_of(bad_vars_scen))
}

train_scen <- train_scen_raw %>% dplyr::select(-CASE_ID)
test_scen  <- test_scen_raw  %>% dplyr::select(-CASE_ID)

cat("\nFinal", targeted_scenario_glm_name, "columns:\n")
print(names(train_scen))
cat(targeted_scenario_glm_name, "final split sizes: Train =", nrow(train_scen), "| Test =", nrow(test_scen), "\n")

targeted_scenario_glm_fit <- glm(
  SevereBinary ~ .,
  data = train_scen,
  family = binomial(),
  control = glm.control(maxit = 50)
)

glm_scenario <- targeted_scenario_glm_fit

cat(targeted_scenario_glm_name, "converged? ", targeted_scenario_glm_fit$converged, "\n", sep = "")

targeted_scenario_test_phat <- predict(targeted_scenario_glm_fit, newdata = test_scen, type = "response")
test_phat_scen <- targeted_scenario_test_phat
test_y_scen <- test_scen$SevereBinary

if (!requireNamespace("pROC", quietly = TRUE)) install.packages("pROC")
roc_obj <- pROC::roc(test_y_scen, targeted_scenario_test_phat, levels = c("Other","Severe/Fatal"), quiet = TRUE)

cat(targeted_scenario_glm_name, "TEST AUC:", round(pROC::auc(roc_obj), 4), "\n")

pred_05_scen <- factor(ifelse(targeted_scenario_test_phat >= 0.5, "Severe/Fatal", "Other"), levels = levels(test_y_scen))
cat("\n", targeted_scenario_glm_name, "confusion matrix (0.50 cutoff):\n", sep = "")
print(table(Pred = pred_05_scen, Truth = test_y_scen))

prev_scen <- mean(train_scen$SevereBinary == "Severe/Fatal")
pred_prev_scen <- factor(ifelse(targeted_scenario_test_phat >= prev_scen, "Severe/Fatal", "Other"), levels = levels(test_y_scen))
cat("\n", targeted_scenario_glm_name, "confusion matrix (prevalence cutoff):\n", sep = "")
print(table(Pred = pred_prev_scen, Truth = test_y_scen))

coefs <- summary(targeted_scenario_glm_fit)$coefficients

or_tbl_scen <- data.frame(
  term = rownames(coefs),
  estimate = coefs[, "Estimate"],
  std.error = coefs[, "Std. Error"],
  p.value = coefs[, "Pr(>|z|)"],
  stringsAsFactors = FALSE
) %>%
  dplyr::filter(term != "(Intercept)") %>%
  dplyr::mutate(
    OR = exp(estimate),
    conf.low = exp(estimate - 1.96 * std.error),
    conf.high = exp(estimate + 1.96 * std.error)
  ) %>%
  dplyr::arrange(dplyr::desc(abs(estimate)))

cat("\nTop", targeted_scenario_glm_name, "odds ratios:\n")
print(head(or_tbl_scen, 15))
```

# Targeted subgroup extension model
```{r subgroup_extension_model, message=FALSE, warning=FALSE}
baseline_glm_name <- "Baseline GLM"
subgroup_extension_glm_name <- "Baseline GLM + Subgroup Extension"

if (!exists("switrs_model_ready")) stop("switrs_model_ready not found.")
if (!exists("candidate_predictors")) stop("candidate_predictors not found. Run the Baseline GLM chunk first.")
if (!exists("train_data") || !exists("test_data")) stop("train_data/test_data not found. Run shared_imputation and shared_train_test_split first.")
if (!exists("expand_predictors_with_miss")) stop("expand_predictors_with_miss() not found. Run the shared_imputation chunk first.")
if (!exists("is_perfect_separator")) stop("is_perfect_separator() not found. Run the Baseline GLM chunk first.")

subgroup_addons <- intersect(c("intersection_ped", "highway_night"), names(switrs_model_ready))
if (length(subgroup_addons) == 0) stop("Neither intersection_ped nor highway_night is available in switrs_model_ready.")

candidate_predictors_subgroup <- unique(c(candidate_predictors, subgroup_addons))
candidate_predictors_subgroup <- setdiff(candidate_predictors_subgroup, leakage_vars)

cat(subgroup_extension_glm_name, "predictors added:\n")
print(subgroup_addons)

subgroup_predictors_final <- expand_predictors_with_miss(train_data, candidate_predictors_subgroup)

train_set_sub_raw <- train_data %>%
  dplyr::select(all_of(c("CASE_ID", "SevereBinary", subgroup_predictors_final)))

test_set_sub_raw <- test_data %>%
  dplyr::select(all_of(c("CASE_ID", "SevereBinary", subgroup_predictors_final)))

is_constant_sub <- sapply(train_set_sub_raw %>% dplyr::select(-CASE_ID, -SevereBinary), function(x) {
  ux <- unique(x)
  ux <- ux[!is.na(ux)]
  length(ux) <= 1
})

if (any(is_constant_sub)) {
  cat("Dropping", subgroup_extension_glm_name, "train-constant predictors:\n")
  print(names(is_constant_sub)[is_constant_sub])

  keep_cols_sub <- c("CASE_ID", "SevereBinary", names(is_constant_sub)[!is_constant_sub])
  train_set_sub_raw <- train_set_sub_raw %>% dplyr::select(all_of(keep_cols_sub))
  test_set_sub_raw  <- test_set_sub_raw  %>% dplyr::select(all_of(keep_cols_sub))
}

y_train_sub <- train_set_sub_raw$SevereBinary
pred_names_sub <- setdiff(names(train_set_sub_raw), c("CASE_ID", "SevereBinary"))

sep_flags_sub <- sapply(pred_names_sub, function(v) {
  is_perfect_separator(train_set_sub_raw[[v]], y_train_sub)
})
bad_vars_sub <- names(sep_flags_sub)[sep_flags_sub]

cat(subgroup_extension_glm_name, "predictors dropped for practical separation:", length(bad_vars_sub), "\n")
if (length(bad_vars_sub) > 0) print(bad_vars_sub)

if (length(bad_vars_sub) > 0) {
  train_set_sub_raw <- train_set_sub_raw %>% dplyr::select(-dplyr::all_of(bad_vars_sub))
  test_set_sub_raw  <- test_set_sub_raw  %>% dplyr::select(-dplyr::all_of(bad_vars_sub))
}

train_set_sub <- train_set_sub_raw %>% dplyr::select(-CASE_ID)
test_set_sub  <- test_set_sub_raw  %>% dplyr::select(-CASE_ID)

subgroup_extension_glm_fit <- glm(
  SevereBinary ~ .,
  data = train_set_sub,
  family = binomial(),
  control = glm.control(maxit = 50)
)

glm_fit_subgroup <- subgroup_extension_glm_fit

cat(subgroup_extension_glm_name, "converged? ", subgroup_extension_glm_fit$converged, "\n", sep = "")

subgroup_extension_test_phat <- as.numeric(
  predict(subgroup_extension_glm_fit, newdata = test_set_sub, type = "response")
)
test_phat_subgroup <- subgroup_extension_test_phat
test_y_subgroup <- test_set_sub$SevereBinary

if (!requireNamespace("pROC", quietly = TRUE)) install.packages("pROC")

auc_subgroup <- NA_real_
auc_baseline_now <- NA_real_

if (requireNamespace("pROC", quietly = TRUE)) {
  auc_subgroup <- as.numeric(
    pROC::auc(
      pROC::roc(test_y_subgroup, subgroup_extension_test_phat,
                levels = c("Other", "Severe/Fatal"), quiet = TRUE)
    )
  )

  if (exists("test_set") && exists("test_phat")) {
    auc_baseline_now <- as.numeric(
      pROC::auc(
        pROC::roc(test_set$SevereBinary, test_phat,
                  levels = c("Other", "Severe/Fatal"), quiet = TRUE)
      )
    )
  }
}

metric_at_cutoff_local <- function(y_factor, p, cutoff) {
  y01 <- as.integer(y_factor == "Severe/Fatal")
  pred <- as.integer(p >= cutoff)

  tp <- sum(pred == 1 & y01 == 1)
  tn <- sum(pred == 0 & y01 == 0)
  fp <- sum(pred == 1 & y01 == 0)
  fn <- sum(pred == 0 & y01 == 1)

  data.frame(
    cutoff = cutoff,
    sensitivity = ifelse(tp + fn > 0, tp / (tp + fn), NA_real_),
    specificity = ifelse(tn + fp > 0, tn / (tn + fp), NA_real_),
    precision = ifelse(tp + fp > 0, tp / (tp + fp), NA_real_),
    accuracy = (tp + tn) / (tp + tn + fp + fn)
  )
}

prev_sub <- mean(train_set_sub$SevereBinary == "Severe/Fatal")

subgroup_metrics_05   <- metric_at_cutoff_local(test_y_subgroup, subgroup_extension_test_phat, 0.5)
subgroup_metrics_prev <- metric_at_cutoff_local(test_y_subgroup, subgroup_extension_test_phat, prev_sub)

subgroup_model_compare <- dplyr::bind_rows(
  data.frame(
    model = baseline_glm_name,
    test_auc = auc_baseline_now
  ),
  data.frame(
    model = subgroup_extension_glm_name,
    test_auc = auc_subgroup
  )
)

subgroup_perf_tbl <- dplyr::bind_rows(
  cbind(model = subgroup_extension_glm_name, threshold = "0.50", subgroup_metrics_05),
  cbind(model = subgroup_extension_glm_name, threshold = "Prevalence", subgroup_metrics_prev)
)

cat("\nAUC comparison:\n")
print(subgroup_model_compare)

cat("\nSubgroup extension operating-point metrics:\n")
print(subgroup_perf_tbl)

cat("\n", subgroup_extension_glm_name, "confusion matrix (cutoff=0.50):\n", sep = "")
print(table(
  Pred = factor(ifelse(subgroup_extension_test_phat >= 0.5, "Severe/Fatal", "Other"),
                levels = c("Other", "Severe/Fatal")),
  Truth = test_y_subgroup
))

cat("\n", subgroup_extension_glm_name, "confusion matrix (cutoff=prevalence=", round(prev_sub, 4), "): \n", sep = "")
print(table(
  Pred = factor(ifelse(subgroup_extension_test_phat >= prev_sub, "Severe/Fatal", "Other"),
                levels = c("Other", "Severe/Fatal")),
  Truth = test_y_subgroup
))
```

# EXECUTIVE SUMMARY (Storyboard-ready text)
```{r exec_summary, results='asis'}
cat("## Executive summary\n\n")

baseline_glm_name <- "Baseline GLM"
lasso_glmnet_name <- "LASSO GLMNET"

auc_glm_val    <- NA_real_
auc_glmnet_val <- NA_real_

if (exists("all_model_compare_tbl")) {
  auc_glm_val <- all_model_compare_tbl %>%
    dplyr::filter(model == baseline_glm_name) %>%
    dplyr::slice(1) %>%
    dplyr::pull(test_auc)

  auc_glmnet_val <- all_model_compare_tbl %>%
    dplyr::filter(model == lasso_glmnet_name) %>%
    dplyr::slice(1) %>%
    dplyr::pull(test_auc)
}

weekend_diff_pp <- NA_real_
if ("weekend" %in% names(switrs_reduced)) {
  tmp <- switrs_reduced %>%
    dplyr::filter(!is.na(SevereBinary), !is.na(weekend)) %>%
    dplyr::group_by(weekend) %>%
    dplyr::summarise(p = mean(SevereBinary == "Severe/Fatal"), n = dplyr::n(), .groups="drop")
  if (all(c("Weekend","Weekday") %in% tmp$weekend)) {
    weekend_diff_pp <- 100 * (tmp$p[tmp$weekend=="Weekend"] - tmp$p[tmp$weekend=="Weekday"])
  }
}

top_hotspot <- NA_character_
top_hotspot_n <- NA_integer_
if (exists("hotspots_by_count") && nrow(hotspots_by_count) > 0) {
  top_hotspot <- hotspots_by_count$location_key[1]
  top_hotspot_n <- hotspots_by_count$severe_n[1]
}

top_scenario <- NA_character_
top_scenario_rate <- NA_real_
if (exists("top_by_rate") && nrow(top_by_rate) > 0) {
  top_scenario <- top_by_rate$scenario[1]
  top_scenario_rate <- top_by_rate$severe_per_100[1]
}

interaction_callouts <- NULL
interaction_vars <- intersect(c("night_dui","rain_speed","ped_dark"), names(switrs_reduced))
if (length(interaction_vars) > 0) {
  interaction_callouts <- purrr::map_chr(interaction_vars, function(v) {
    df <- switrs_reduced %>%
      dplyr::filter(!is.na(SevereBinary)) %>%
      dplyr::mutate(g = as.character(.data[[v]])) %>%
      dplyr::filter(g %in% c("Yes","No")) %>%
      dplyr::group_by(g) %>%
      dplyr::summarise(rate = 100*mean(SevereBinary=="Severe/Fatal"), n = dplyr::n(), .groups="drop")
    yes_row <- df[df$g=="Yes", ]
    if (nrow(yes_row) == 0) return(NA_character_)

    label <- dplyr::recode(v,
      night_dui  = "Night × DUI proxy",
      rain_speed = "Rain × Unsafe speed proxy",
      ped_dark   = "Pedestrian × Dark",
      .default = v
    )
    paste0(label, ": ",
           round(yes_row$rate[1], 2), " per 100 (n=", yes_row$n[1], ")")
  })
  interaction_callouts <- interaction_callouts[!is.na(interaction_callouts)]
}

cat(
  "We analyzed crash-level SWITRS records (one row per CASE_ID) to identify **where** and **under what conditions** severe/fatal crashes are most concentrated.\n\n"
)

if (!is.na(top_hotspot)) {
  cat(sprintf(
    "- **Where (burden):** The highest severe/fatal *count* hotspot is **%s** (%s severe/fatal crashes).\n",
    top_hotspot, top_hotspot_n
  ))
}

if (!is.na(top_scenario)) {
  cat(sprintf(
    "- **When/conditions (risk intensity):** The highest severe/fatal *rate* scenario is **%s** (%.2f severe/fatal per 100 crashes).\n",
    top_scenario, top_scenario_rate
  ))
}

if (!is.na(weekend_diff_pp)) {
  cat(sprintf(
    "- **Weekend effect:** Severe/fatal risk is about **%.2f percentage points** higher on weekends than weekdays (difference in proportions).\n",
    weekend_diff_pp
  ))
}

if (!is.null(interaction_callouts) && length(interaction_callouts) > 0) {
  cat("\n**Compound risk signals (interaction terms):**\n")
  for (s in interaction_callouts) cat("- ", s, "\n", sep = "")
}

cat("\n")

cat(sprintf(
  "**Predictive validation:** %s achieves test AUC ≈ %s, and %s achieves test AUC ≈ %s.\n\n",
  baseline_glm_name,
  ifelse(is.na(auc_glm_val), "N/A", format(round(auc_glm_val, 3), nsmall = 3)),
  lasso_glmnet_name,
  ifelse(is.na(auc_glmnet_val), "N/A", format(round(auc_glmnet_val, 3), nsmall = 3))
))

cat("### Recommended actions\n")
cat("- **Deploy/enforce at the highest-burden hotspots** (top severe/fatal counts) for maximum near-term impact.\n")
cat("- **Prioritize the highest-risk scenarios** (top severe/fatal rates) for targeted operations (visibility, enforcement windows).\n")
cat("- **Use compound-risk conditions** (e.g., night×DUI, rain×speed proxies, pedestrian×dark) to design precision interventions.\n")
```

# model diagnostics: class imbalance and missingness
```{r}
md <- NULL
if (exists("model_df_clean2")) md <- model_df_clean2
if (is.null(md) && exists("model_df_clean")) md <- model_df_clean
if (is.null(md)) stop("No modeling frame found. Run the baseline modeling chunk first.")

cat("Class balance (modeling frame):\n")
print(table(md$SevereBinary))

cat("\nRemaining NA% by column (should be ~0 after preprocessing):\n")
remaining_na <- sort(colMeans(is.na(md)) * 100, decreasing = TRUE)
print(head(remaining_na, 20))
```

# next-step hooks (merged data expansion)
```{r}
pattern <- "PARTY_|DRIVER|VEH|SOBRIETY|DRUG|SAFETY|EQUIP|AT_FAULT|AGE|SEX|VIOL"

cands <- names(switrs_reduced)[
  grepl(pattern, names(switrs_reduced), ignore.case = TRUE)
]

leak_like <- grepl("KILLED|FATAL|INJUR|SEVER|KSI|COUNT_", cands, ignore.case = TRUE)
cands <- cands[!leak_like]

cat("Candidate merged-only predictors to consider next (non-leakage):\n")
print(head(cands, 50))
```

#################################### Start of Undersampling

# Undersampling 1-1 split from training data
```{r undersample_shared, message=FALSE, warning=FALSE}
if (!exists("train_data") || !exists("test_data")) {
  stop("train_data/test_data not found. Run shared_train_test_split first.")
}

if (!("SevereBinary" %in% names(train_data))) {
  stop("SevereBinary not found in train_data.")
}

undersample_1to1 <- function(data, outcome = "SevereBinary", seed = 544) {
  set.seed(seed)

  df <- data %>%
    dplyr::filter(!is.na(.data[[outcome]]))

  class_counts <- table(df[[outcome]])
  if (length(class_counts) != 2) {
    stop("Outcome must have exactly 2 non-missing classes for 1:1 undersampling.")
  }

  target_n <- min(class_counts)

  df_bal <- df %>%
    dplyr::group_by(.data[[outcome]]) %>%
    dplyr::slice_sample(n = target_n, replace = FALSE) %>%
    dplyr::ungroup()

  df_bal[[outcome]] <- factor(df_bal[[outcome]], levels = levels(data[[outcome]]))
  df_bal
}

align_factor_levels_to_train <- function(train_df, test_df,
                                         outcome = "SevereBinary",
                                         id_col = "CASE_ID") {

  common_cols <- intersect(names(train_df), names(test_df))
  pred_cols <- setdiff(common_cols, c(outcome, id_col))

  for (v in pred_cols) {
    is_cat <- is.factor(train_df[[v]]) || is.character(train_df[[v]]) ||
              is.factor(test_df[[v]])  || is.character(test_df[[v]])

    if (!is_cat) next

    train_chr <- as.character(train_df[[v]])
    test_chr  <- as.character(test_df[[v]])

    observed_train_levels <- unique(train_chr[!is.na(train_chr)])

    if (length(observed_train_levels) == 0) next

    fallback_level <- if ("Not Stated" %in% observed_train_levels) {
      "Not Stated"
    } else {
      observed_train_levels[1]
    }

    unseen_in_test <- !is.na(test_chr) & !(test_chr %in% observed_train_levels)
    if (any(unseen_in_test)) {
      test_chr[unseen_in_test] <- fallback_level
    }

    train_df[[v]] <- factor(train_chr, levels = observed_train_levels)
    test_df[[v]]  <- factor(test_chr,  levels = observed_train_levels)
  }

  list(train = train_df, test = test_df)
}

is_perfect_separator <- function(x, y, min_level_n = 30) {
  if (is.character(x) || is.factor(x)) {
    tab <- table(x, y, useNA = "no")
    lvl_n <- rowSums(tab)
    tab2 <- tab[lvl_n >= min_level_n, , drop = FALSE]
    if (nrow(tab2) == 0) return(FALSE)
    any(tab2[, "Other"] == 0 | tab2[, "Severe/Fatal"] == 0)
  } else if (is.numeric(x) || is.integer(x)) {
    ux <- sort(unique(x))
    ux <- ux[!is.na(ux)]
    if (length(ux) <= 1) return(FALSE)

    qs <- unique(stats::quantile(x, probs = seq(0.1, 0.9, by = 0.1), na.rm = TRUE))
    qs <- qs[!is.na(qs)]

    for (cut in qs) {
      left  <- y[x <= cut]
      right <- y[x >  cut]

      if (length(left) >= min_level_n && length(right) >= min_level_n) {
        if (all(left == "Other") || all(left == "Severe/Fatal") ||
            all(right == "Other") || all(right == "Severe/Fatal")) {
          return(TRUE)
        }
      }
    }
    FALSE
  } else {
    FALSE
  }
}

binary_metrics <- function(truth, prob, cutoff = 0.5, positive = "Severe/Fatal") {
  truth_chr <- as.character(truth)
  truth01 <- as.integer(truth_chr == positive)
  pred01  <- as.integer(prob >= cutoff)

  tp <- sum(pred01 == 1 & truth01 == 1, na.rm = TRUE)
  tn <- sum(pred01 == 0 & truth01 == 0, na.rm = TRUE)
  fp <- sum(pred01 == 1 & truth01 == 0, na.rm = TRUE)
  fn <- sum(pred01 == 0 & truth01 == 1, na.rm = TRUE)

  data.frame(
    cutoff = cutoff,
    sensitivity = ifelse(tp + fn > 0, tp / (tp + fn), NA_real_),
    specificity = ifelse(tn + fp > 0, tn / (tn + fp), NA_real_),
    precision   = ifelse(tp + fp > 0, tp / (tp + fp), NA_real_),
    accuracy    = ifelse(tp + tn + fp + fn > 0, (tp + tn) / (tp + tn + fp + fn), NA_real_)
  )
}

train_data_under <- undersample_1to1(train_data, outcome = "SevereBinary", seed = 544)

cat("Original train class balance:\n")
print(table(train_data$SevereBinary, useNA = "ifany"))

cat("\nUndersampled train class balance (1:1):\n")
print(table(train_data_under$SevereBinary, useNA = "ifany"))

cat("\nTest class balance (unchanged):\n")
print(table(test_data$SevereBinary, useNA = "ifany"))

orig_prev <- mean(train_data$SevereBinary == "Severe/Fatal", na.rm = TRUE)
cat("\nOriginal train Severe/Fatal prevalence =", round(orig_prev, 4), "\n")
```

# Baseline GLM with 1:1 undersampling
```{r baseline_glm_undersampled, message=FALSE, warning=FALSE}
set.seed(544)

baseline_glm_us_name <- "Baseline GLM (Undersampled 1:1)"

if (!exists("train_data_under") || !exists("test_data")) {
  stop("Run the undersample_shared chunk first.")
}
if (!exists("expand_predictors_with_miss")) {
  stop("expand_predictors_with_miss() not found. Run shared_imputation first.")
}
if (!exists("leakage_vars")) leakage_vars <- character(0)

candidate_predictors_us <- intersect(
  c(
    "year",
    "day_group",
    "night",
    "friday_night",
    "weather_group",
    "lighting_group",
    "ROAD_SURFACE",
    "INTERSECTION",
    "STATE_HWY_IND",
    "TYPE_OF_COLLISION",
    "PRIMARY_COLL_FACTOR",
    "school_bus_proxy",
    "pcf_dui",
    "pcf_speed",
    "ALCOHOL_INVOLVED",
    "HIT_AND_RUN",
    "is_ped",
    "BICYCLE_ACCIDENT",
    "MOTORCYCLE_ACCIDENT",
    "TRUCK_ACCIDENT",
    "night_dui",
    "rain_speed",
    "ped_dark",
    "n_parties",
    "any_party_alcohol",
    "any_party_drug",
    "any_party_at_fault",
    "mean_party_age"
  ),
  names(switrs_model_ready)
)

candidate_predictors_us <- setdiff(candidate_predictors_us, leakage_vars)

forbidden_redundant <- c(
  "DAY_OF_WEEK", "weekend", "time_bucket",
  "WEATHER_1", "is_rain",
  "LIGHTING", "is_dark",
  "PCF_VIOL_CATEGORY",
  "PEDESTRIAN_ACCIDENT"
)

if (length(intersect(candidate_predictors_us, forbidden_redundant)) > 0) {
  stop("Redundant predictors detected in candidate_predictors_us.")
}

baseline_predictors_us_final <- expand_predictors_with_miss(train_data_under, candidate_predictors_us)

train_set_us_raw <- train_data_under %>%
  dplyr::select(all_of(c("CASE_ID", "SevereBinary", baseline_predictors_us_final)))

test_set_us_raw <- test_data %>%
  dplyr::select(all_of(c("CASE_ID", "SevereBinary", baseline_predictors_us_final)))

aligned_us <- align_factor_levels_to_train(train_set_us_raw, test_set_us_raw)
train_set_us_raw <- aligned_us$train
test_set_us_raw  <- aligned_us$test

is_constant_us <- sapply(train_set_us_raw %>% dplyr::select(-CASE_ID, -SevereBinary), function(x) {
  ux <- unique(x)
  ux <- ux[!is.na(ux)]
  length(ux) <= 1
})

if (any(is_constant_us)) {
  cat("Dropping baseline undersampled train-constant predictors:\n")
  print(names(is_constant_us)[is_constant_us])

  keep_cols_us <- c("CASE_ID", "SevereBinary", names(is_constant_us)[!is_constant_us])
  train_set_us_raw <- train_set_us_raw %>% dplyr::select(all_of(keep_cols_us))
  test_set_us_raw  <- test_set_us_raw  %>% dplyr::select(all_of(keep_cols_us))
}

y_train_us <- train_set_us_raw$SevereBinary
pred_names_us <- setdiff(names(train_set_us_raw), c("CASE_ID", "SevereBinary"))

sep_flags_us <- sapply(pred_names_us, function(v) {
  is_perfect_separator(train_set_us_raw[[v]], y_train_us)
})

bad_vars_us <- names(sep_flags_us)[sep_flags_us]

cat("Baseline undersampled predictors dropped for practical separation:", length(bad_vars_us), "\n")
if (length(bad_vars_us) > 0) print(bad_vars_us)

if (length(bad_vars_us) > 0) {
  train_set_us_raw <- train_set_us_raw %>% dplyr::select(-dplyr::all_of(bad_vars_us))
  test_set_us_raw  <- test_set_us_raw  %>% dplyr::select(-dplyr::all_of(bad_vars_us))
}

train_set_us <- train_set_us_raw %>% dplyr::select(-CASE_ID)
test_set_us  <- test_set_us_raw  %>% dplyr::select(-CASE_ID)

baseline_glm_us_fit <- glm(
  SevereBinary ~ .,
  data = train_set_us,
  family = binomial(),
  control = glm.control(maxit = 50)
)

baseline_glm_us_test_phat <- as.numeric(
  predict(baseline_glm_us_fit, newdata = test_set_us, type = "response")
)

if (!requireNamespace("pROC", quietly = TRUE)) install.packages("pROC")

baseline_glm_us_auc <- NA_real_
if (requireNamespace("pROC", quietly = TRUE)) {
  baseline_glm_us_roc <- pROC::roc(
    test_set_us$SevereBinary,
    baseline_glm_us_test_phat,
    levels = c("Other", "Severe/Fatal"),
    quiet = TRUE
  )
  baseline_glm_us_auc <- as.numeric(pROC::auc(baseline_glm_us_roc))
}

baseline_glm_us_metrics_05 <- binary_metrics(test_set_us$SevereBinary, baseline_glm_us_test_phat, cutoff = 0.50)

cat("\n", baseline_glm_us_name, "converged? ", baseline_glm_us_fit$converged, "\n", sep = "")
cat(baseline_glm_us_name, "TEST AUC:", round(baseline_glm_us_auc, 4), "\n")

cat("\nMetrics at cutoff = 0.50\n")
print(baseline_glm_us_metrics_05)

cat("\nConfusion matrix (cutoff = 0.50):\n")
print(table(
  Pred = factor(ifelse(baseline_glm_us_test_phat >= 0.5, "Severe/Fatal", "Other"),
                levels = c("Other", "Severe/Fatal")),
  Truth = test_set_us$SevereBinary
))
```

# LASSO GLMNET with 1:1 undersampling
```{r glmnet_model_undersampled, message=FALSE, warning=FALSE}
lasso_glmnet_us_name <- "LASSO GLMNET (Undersampled 1:1)"

if (!exists("train_set_us") || !exists("test_set_us")) {
  stop("Run baseline_glm_undersampled first.")
}

if (!requireNamespace("glmnet", quietly = TRUE)) install.packages("glmnet")
library(glmnet)

train_set_us$.split <- "train"
test_set_us$.split  <- "test"

combo_us <- dplyr::bind_rows(train_set_us, test_set_us)
train_idx_us <- combo_us$.split == "train"
test_idx_us  <- combo_us$.split == "test"

combo_mm_us <- combo_us %>% dplyr::select(-.split)

X_all_us <- model.matrix(SevereBinary ~ ., data = combo_mm_us)[, -1, drop = FALSE]

x_train_us <- X_all_us[train_idx_us, , drop = FALSE]
x_test_us  <- X_all_us[test_idx_us,  , drop = FALSE]

y_train_us_glmnet <- as.integer(combo_us$SevereBinary[train_idx_us] == "Severe/Fatal")
y_test_us_glmnet  <- as.integer(combo_us$SevereBinary[test_idx_us]  == "Severe/Fatal")

train_set_us$.split <- NULL
test_set_us$.split  <- NULL

set.seed(544)
lasso_glmnet_us_cv_fit <- cv.glmnet(
  x = x_train_us,
  y = y_train_us_glmnet,
  family = "binomial",
  alpha = 1,
  nfolds = 5,
  type.measure = "auc"
)

lasso_glmnet_us_test_phat <- as.numeric(
  predict(lasso_glmnet_us_cv_fit, newx = x_test_us, s = "lambda.1se", type = "response")
)

lasso_glmnet_us_auc <- NA_real_
if (requireNamespace("pROC", quietly = TRUE)) {
  roc_obj_us <- pROC::roc(response = y_test_us_glmnet, predictor = lasso_glmnet_us_test_phat, quiet = TRUE)
  lasso_glmnet_us_auc <- as.numeric(pROC::auc(roc_obj_us))
}

lasso_glmnet_us_metrics_05 <- binary_metrics(test_set_us$SevereBinary, lasso_glmnet_us_test_phat, cutoff = 0.50)

cat(lasso_glmnet_us_name, "best lambda (min):", lasso_glmnet_us_cv_fit$lambda.min, "\n")
cat(lasso_glmnet_us_name, "1-SE lambda:", lasso_glmnet_us_cv_fit$lambda.1se, "\n")
cat(lasso_glmnet_us_name, "TEST AUC:", round(lasso_glmnet_us_auc, 4), "\n")

cat("\nMetrics at cutoff = 0.50\n")
print(lasso_glmnet_us_metrics_05)

cat("\nConfusion matrix (cutoff = 0.50):\n")
print(table(
  Pred = ifelse(lasso_glmnet_us_test_phat >= 0.5, 1, 0),
  Truth = y_test_us_glmnet
))

coef_mat_us <- as.matrix(coef(lasso_glmnet_us_cv_fit, s = "lambda.1se"))
coef_df_us <- data.frame(
  term = rownames(coef_mat_us),
  coef = as.numeric(coef_mat_us[, 1]),
  row.names = NULL
) %>%
  dplyr::filter(term != "(Intercept)", coef != 0) %>%
  dplyr::arrange(dplyr::desc(abs(coef)))

cat("\nTop 20 non-zero coefficients by |coef| (undersampled glmnet):\n")
print(head(coef_df_us, 20))
```

# Targeted Scenario GLM with 1:1 undersampling
```{r targeted_scenario_glm_undersampled, message=FALSE, warning=FALSE}
set.seed(544)

targeted_scenario_glm_us_name <- "Targeted Scenario GLM (Undersampled 1:1)"

scenario_predictors_us <- intersect(
  c(
    "day_group",
    "night",
    "is_ped",
    "INTERSECTION",
    "STATE_HWY_IND",
    "lighting_group",
    "pcf_dui",
    "friday_night",
    "ped_dark",
    "intersection_ped",
    "highway_night",
    "night_dui"
  ),
  names(switrs_model_ready)
)

scenario_predictors_us_final <- expand_predictors_with_miss(train_data_under, scenario_predictors_us)

train_scen_us_raw <- train_data_under %>%
  dplyr::select(all_of(c("CASE_ID", "SevereBinary", scenario_predictors_us_final)))

test_scen_us_raw <- test_data %>%
  dplyr::select(all_of(c("CASE_ID", "SevereBinary", scenario_predictors_us_final)))

aligned_scen_us <- align_factor_levels_to_train(train_scen_us_raw, test_scen_us_raw)
train_scen_us_raw <- aligned_scen_us$train
test_scen_us_raw  <- aligned_scen_us$test

is_constant_scen_us <- sapply(train_scen_us_raw %>% dplyr::select(-CASE_ID, -SevereBinary), function(x) {
  ux <- unique(x)
  ux <- ux[!is.na(ux)]
  length(ux) <= 1
})

if (any(is_constant_scen_us)) {
  cat("Dropping targeted-scenario undersampled train-constant predictors:\n")
  print(names(is_constant_scen_us)[is_constant_scen_us])

  keep_cols_scen_us <- c("CASE_ID", "SevereBinary", names(is_constant_scen_us)[!is_constant_scen_us])
  train_scen_us_raw <- train_scen_us_raw %>% dplyr::select(all_of(keep_cols_scen_us))
  test_scen_us_raw  <- test_scen_us_raw  %>% dplyr::select(all_of(keep_cols_scen_us))
}

y_train_scen_us <- train_scen_us_raw$SevereBinary
pred_names_scen_us <- setdiff(names(train_scen_us_raw), c("CASE_ID", "SevereBinary"))

sep_flags_scen_us <- sapply(pred_names_scen_us, function(v) {
  is_perfect_separator(train_scen_us_raw[[v]], y_train_scen_us)
})

bad_vars_scen_us <- names(sep_flags_scen_us)[sep_flags_scen_us]

cat("Targeted scenario undersampled predictors dropped for practical separation:", length(bad_vars_scen_us), "\n")
if (length(bad_vars_scen_us) > 0) print(bad_vars_scen_us)

if (length(bad_vars_scen_us) > 0) {
  train_scen_us_raw <- train_scen_us_raw %>% dplyr::select(-dplyr::all_of(bad_vars_scen_us))
  test_scen_us_raw  <- test_scen_us_raw  %>% dplyr::select(-dplyr::all_of(bad_vars_scen_us))
}

train_scen_us <- train_scen_us_raw %>% dplyr::select(-CASE_ID)
test_scen_us  <- test_scen_us_raw  %>% dplyr::select(-CASE_ID)

targeted_scenario_glm_us_fit <- glm(
  SevereBinary ~ .,
  data = train_scen_us,
  family = binomial(),
  control = glm.control(maxit = 50)
)

targeted_scenario_glm_us_test_phat <- as.numeric(
  predict(targeted_scenario_glm_us_fit, newdata = test_scen_us, type = "response")
)

targeted_scenario_glm_us_auc <- NA_real_
if (requireNamespace("pROC", quietly = TRUE)) {
  roc_obj_scen_us <- pROC::roc(
    test_scen_us$SevereBinary,
    targeted_scenario_glm_us_test_phat,
    levels = c("Other", "Severe/Fatal"),
    quiet = TRUE
  )
  targeted_scenario_glm_us_auc <- as.numeric(pROC::auc(roc_obj_scen_us))
}

targeted_scenario_glm_us_metrics_05 <- binary_metrics(test_scen_us$SevereBinary, targeted_scenario_glm_us_test_phat, cutoff = 0.50)

cat("\n", targeted_scenario_glm_us_name, "converged? ", targeted_scenario_glm_us_fit$converged, "\n", sep = "")
cat(targeted_scenario_glm_us_name, "TEST AUC:", round(targeted_scenario_glm_us_auc, 4), "\n")

cat("\nMetrics at cutoff = 0.50\n")
print(targeted_scenario_glm_us_metrics_05)

cat("\nConfusion matrix (cutoff = 0.50):\n")
print(table(
  Pred = factor(ifelse(targeted_scenario_glm_us_test_phat >= 0.5, "Severe/Fatal", "Other"),
                levels = c("Other", "Severe/Fatal")),
  Truth = test_scen_us$SevereBinary
))
```

# Baseline GLM + Subgroup Extension with 1:1 undersampling
```{r subgroup_extension_glm_undersampled, message=FALSE, warning=FALSE}
set.seed(544)

subgroup_extension_glm_us_name <- "Baseline GLM + Subgroup Extension (Undersampled 1:1)"

if (!exists("candidate_predictors_us")) {
  stop("Run baseline_glm_undersampled first so candidate_predictors_us is defined.")
}

subgroup_addons_us <- intersect(c("intersection_ped", "highway_night"), names(switrs_model_ready))
candidate_predictors_sub_us <- unique(c(candidate_predictors_us, subgroup_addons_us))
candidate_predictors_sub_us <- setdiff(candidate_predictors_sub_us, leakage_vars)

subgroup_predictors_us_final <- expand_predictors_with_miss(train_data_under, candidate_predictors_sub_us)

train_sub_us_raw <- train_data_under %>%
  dplyr::select(all_of(c("CASE_ID", "SevereBinary", subgroup_predictors_us_final)))

test_sub_us_raw <- test_data %>%
  dplyr::select(all_of(c("CASE_ID", "SevereBinary", subgroup_predictors_us_final)))

aligned_sub_us <- align_factor_levels_to_train(train_sub_us_raw, test_sub_us_raw)
train_sub_us_raw <- aligned_sub_us$train
test_sub_us_raw  <- aligned_sub_us$test

is_constant_sub_us <- sapply(train_sub_us_raw %>% dplyr::select(-CASE_ID, -SevereBinary), function(x) {
  ux <- unique(x)
  ux <- ux[!is.na(ux)]
  length(ux) <= 1
})

if (any(is_constant_sub_us)) {
  cat("Dropping subgroup-extension undersampled train-constant predictors:\n")
  print(names(is_constant_sub_us)[is_constant_sub_us])

  keep_cols_sub_us <- c("CASE_ID", "SevereBinary", names(is_constant_sub_us)[!is_constant_sub_us])
  train_sub_us_raw <- train_sub_us_raw %>% dplyr::select(all_of(keep_cols_sub_us))
  test_sub_us_raw  <- test_sub_us_raw  %>% dplyr::select(all_of(keep_cols_sub_us))
}

y_train_sub_us <- train_sub_us_raw$SevereBinary
pred_names_sub_us <- setdiff(names(train_sub_us_raw), c("CASE_ID", "SevereBinary"))

sep_flags_sub_us <- sapply(pred_names_sub_us, function(v) {
  is_perfect_separator(train_sub_us_raw[[v]], y_train_sub_us)
})

bad_vars_sub_us <- names(sep_flags_sub_us)[sep_flags_sub_us]

cat("Subgroup-extension undersampled predictors dropped for practical separation:", length(bad_vars_sub_us), "\n")
if (length(bad_vars_sub_us) > 0) print(bad_vars_sub_us)

if (length(bad_vars_sub_us) > 0) {
  train_sub_us_raw <- train_sub_us_raw %>% dplyr::select(-dplyr::all_of(bad_vars_sub_us))
  test_sub_us_raw  <- test_sub_us_raw  %>% dplyr::select(-dplyr::all_of(bad_vars_sub_us))
}

train_sub_us <- train_sub_us_raw %>% dplyr::select(-CASE_ID)
test_sub_us  <- test_sub_us_raw  %>% dplyr::select(-CASE_ID)

subgroup_extension_glm_us_fit <- glm(
  SevereBinary ~ .,
  data = train_sub_us,
  family = binomial(),
  control = glm.control(maxit = 50)
)

subgroup_extension_glm_us_test_phat <- as.numeric(
  predict(subgroup_extension_glm_us_fit, newdata = test_sub_us, type = "response")
)

subgroup_extension_glm_us_auc <- NA_real_
if (requireNamespace("pROC", quietly = TRUE)) {
  roc_obj_sub_us <- pROC::roc(
    test_sub_us$SevereBinary,
    subgroup_extension_glm_us_test_phat,
    levels = c("Other", "Severe/Fatal"),
    quiet = TRUE
  )
  subgroup_extension_glm_us_auc <- as.numeric(pROC::auc(roc_obj_sub_us))
}

subgroup_extension_glm_us_metrics_05 <- binary_metrics(test_sub_us$SevereBinary, subgroup_extension_glm_us_test_phat, cutoff = 0.50)

cat("\n", subgroup_extension_glm_us_name, "converged? ", subgroup_extension_glm_us_fit$converged, "\n", sep = "")
cat(subgroup_extension_glm_us_name, "TEST AUC:", round(subgroup_extension_glm_us_auc, 4), "\n")

cat("\nMetrics at cutoff = 0.50\n")
print(subgroup_extension_glm_us_metrics_05)

cat("\nConfusion matrix (cutoff = 0.50):\n")
print(table(
  Pred = factor(ifelse(subgroup_extension_glm_us_test_phat >= 0.5, "Severe/Fatal", "Other"),
                levels = c("Other", "Severe/Fatal")),
  Truth = test_sub_us$SevereBinary
))
```

# Model Comparison
```{r all_model_compare_original_and_undersampled, message=FALSE, warning=FALSE}
if (!requireNamespace("pROC", quietly = TRUE)) install.packages("pROC")

metric_at_cutoff_unified <- function(truth, prob, cutoff, positive = "Severe/Fatal") {

  truth_chr <- if (is.factor(truth) || is.character(truth)) {
    as.character(truth)
  } else {
    as.character(as.integer(truth))
  }

  truth01 <- if (all(na.omit(unique(truth_chr)) %in% c("0", "1"))) {
    as.integer(truth_chr)
  } else {
    as.integer(truth_chr == positive)
  }

  pred01 <- as.integer(prob >= cutoff)

  tp <- sum(pred01 == 1 & truth01 == 1, na.rm = TRUE)
  tn <- sum(pred01 == 0 & truth01 == 0, na.rm = TRUE)
  fp <- sum(pred01 == 1 & truth01 == 0, na.rm = TRUE)
  fn <- sum(pred01 == 0 & truth01 == 1, na.rm = TRUE)

  data.frame(
    cutoff = cutoff,
    sensitivity = ifelse(tp + fn > 0, tp / (tp + fn), NA_real_),
    specificity = ifelse(tn + fp > 0, tn / (tn + fp), NA_real_),
    precision   = ifelse(tp + fp > 0, tp / (tp + fp), NA_real_),
    accuracy    = ifelse(tp + tn + fp + fn > 0, (tp + tn) / (tp + tn + fp + fn), NA_real_),
    stringsAsFactors = FALSE
  )
}

auc_unified <- function(truth, prob, positive = "Severe/Fatal") {

  truth_chr <- if (is.factor(truth) || is.character(truth)) {
    as.character(truth)
  } else {
    as.character(as.integer(truth))
  }

  if (all(na.omit(unique(truth_chr)) %in% c("0", "1"))) {
    truth01 <- as.integer(truth_chr)
    return(as.numeric(pROC::auc(pROC::roc(response = truth01, predictor = prob, quiet = TRUE))))
  }

  truth_fac <- factor(truth_chr, levels = c("Other", positive))
  as.numeric(pROC::auc(
    pROC::roc(response = truth_fac, predictor = prob,
              levels = c("Other", positive), quiet = TRUE)
  ))
}

if (exists("train_data")) {
  original_prev_cutoff <- mean(train_data$SevereBinary == "Severe/Fatal", na.rm = TRUE)
} else if (exists("train_set")) {
  original_prev_cutoff <- mean(train_set$SevereBinary == "Severe/Fatal", na.rm = TRUE)
} else {
  stop("Could not find original training data to compute the prevalence cutoff.")
}

threshold_label_05   <- "0.50"
threshold_label_prev <- paste0("Cutoff = ", round(original_prev_cutoff, 4))

cat("Original train prevalence cutoff =", round(original_prev_cutoff, 4), "\n")

build_model_rows <- function(model_name, truth, prob,
                             include_prevalence_row = TRUE,
                             prevalence_cutoff = original_prev_cutoff,
                             prevalence_label = threshold_label_prev,
                             positive = "Severe/Fatal") {

  if (length(truth) != length(prob)) {
    stop("Length mismatch for model: ", model_name)
  }

  auc_val <- auc_unified(truth, prob, positive = positive)

  row_05 <- metric_at_cutoff_unified(truth, prob, cutoff = 0.50, positive = positive)
  row_05$model <- model_name
  row_05$threshold <- threshold_label_05

  out <- row_05

  if (isTRUE(include_prevalence_row)) {
    row_prev <- metric_at_cutoff_unified(truth, prob, cutoff = prevalence_cutoff, positive = positive)
    row_prev$model <- model_name
    row_prev$threshold <- prevalence_label
    out <- dplyr::bind_rows(out, row_prev)
  }

  out %>%
    dplyr::mutate(test_auc = auc_val) %>%
    dplyr::select(model, test_auc, threshold, sensitivity, specificity, precision, accuracy)
}

all_model_rows <- list()

if (exists("test_set") && exists("test_phat")) {
  all_model_rows[["baseline_original"]] <- build_model_rows(
    model_name = "Baseline GLM",
    truth = test_set$SevereBinary,
    prob  = as.numeric(test_phat),
    include_prevalence_row = TRUE
  )
}

if (exists("y_test") && exists("p_test")) {
  all_model_rows[["lasso_original"]] <- build_model_rows(
    model_name = "LASSO GLMNET",
    truth = y_test,
    prob  = as.numeric(p_test),
    include_prevalence_row = TRUE
  )
}

if (exists("test_y_scen") && exists("test_phat_scen")) {
  all_model_rows[["scenario_original"]] <- build_model_rows(
    model_name = "Targeted Scenario GLM",
    truth = test_y_scen,
    prob  = as.numeric(test_phat_scen),
    include_prevalence_row = TRUE
  )
}

if (exists("test_y_subgroup") && exists("test_phat_subgroup")) {
  all_model_rows[["subgroup_original"]] <- build_model_rows(
    model_name = "Baseline GLM + Subgroup Extension",
    truth = test_y_subgroup,
    prob  = as.numeric(test_phat_subgroup),
    include_prevalence_row = TRUE
  )
}

if (exists("test_set_us") && exists("baseline_glm_us_test_phat")) {
  all_model_rows[["baseline_us"]] <- build_model_rows(
    model_name = "Baseline GLM (Undersampled 1:1)",
    truth = test_set_us$SevereBinary,
    prob  = as.numeric(baseline_glm_us_test_phat),
    include_prevalence_row = FALSE
  )
}

if (exists("test_set_us") && exists("lasso_glmnet_us_test_phat")) {
  all_model_rows[["lasso_us"]] <- build_model_rows(
    model_name = "LASSO GLMNET (Undersampled 1:1)",
    truth = test_set_us$SevereBinary,
    prob  = as.numeric(lasso_glmnet_us_test_phat),
    include_prevalence_row = FALSE
  )
}

if (exists("test_scen_us") && exists("targeted_scenario_glm_us_test_phat")) {
  all_model_rows[["scenario_us"]] <- build_model_rows(
    model_name = "Targeted Scenario GLM (Undersampled 1:1)",
    truth = test_scen_us$SevereBinary,
    prob  = as.numeric(targeted_scenario_glm_us_test_phat),
    include_prevalence_row = FALSE
  )
}

if (exists("test_sub_us") && exists("subgroup_extension_glm_us_test_phat")) {
  all_model_rows[["subgroup_us"]] <- build_model_rows(
    model_name = "Baseline GLM + Subgroup Extension (Undersampled 1:1)",
    truth = test_sub_us$SevereBinary,
    prob  = as.numeric(subgroup_extension_glm_us_test_phat),
    include_prevalence_row = FALSE
  )
}

if (length(all_model_rows) == 0) {
  stop("No model predictions found. Run at least one original or undersampled model chunk first.")
}

all_model_compare_tbl <- dplyr::bind_rows(all_model_rows) %>%
  dplyr::mutate(
    model_group = dplyr::case_when(
      grepl("Undersampled 1:1", model) ~ "Undersampled 1:1",
      TRUE ~ "Original"
    )
  ) %>%
  dplyr::arrange(model_group, dplyr::desc(test_auc), threshold)

cat("\nUnified model comparison table:\n")
print(all_model_compare_tbl)

auc_only_tbl <- all_model_compare_tbl %>%
  dplyr::distinct(model, model_group, test_auc) %>%
  dplyr::arrange(model_group, dplyr::desc(test_auc))

cat("\nAUC-only ranking:\n")
print(auc_only_tbl)
```

AFTER A NEW MODEL:
random forest and check the sensitivity and accuracy again
check if the days are included with weekend and weekday 
undersampling: random sampling frm not severe and then do more testing
after getting better sensitivity use odds ratios
priority based on odds ratio (for policy)

CURRENT:
we need to make more specific and more intensive models and decide on where we're going to allocate resources
need more specific hypothesis testing
  focus on specific days, locations, conditions, freeways, etc. (aka focus on subgroups)
  do the conditions make huge differences
  focus on micro actions (small but impactful)
  what's happening in intersections?
  which groups and subgroups are interesting
  
Midterm presentation feedback:
(1) A binary target is well-suited for model implementation.
(2) For the research questions, careful consideration of causality and correlation may be required.
Please consider using non-binary (multinomial) features directly or merging certain categories. For example, within the Lighting feature, you might merge some dark-related categories because categories lead to too many binary features after one-hot encoding.