---
title: "R Notebook"
output: html_notebook
---
# Load packages used for merging
```{r}
# What: Load data-wrangling packages used for merge + diagnostics.
# Why: Ensures joins/aggregations are consistent and readable.

library(dplyr)
library(stringr)
```

# Load in data
```{r}
na_tokens <- c("NA","N/A","NULL","null", "", " ", "  ", "   ")
crashes  <- read.csv("Crashes.csv",  na.strings = na_tokens, stringsAsFactors = FALSE)
parties  <- read.csv("Parties.csv",  na.strings = na_tokens, stringsAsFactors = FALSE)
victims  <- read.csv("Victims.csv",  na.strings = na_tokens, stringsAsFactors = FALSE)
```

# Confrim keys exist and basic structure
```{r}
# What: Confirm CASE_ID exists and check basic sizes.
# Why: CASE_ID is the primary join key; we need it in all three tables.

stopifnot("CASE_ID" %in% names(crashes))
stopifnot("CASE_ID" %in% names(parties))
stopifnot("CASE_ID" %in% names(victims))

cat("Rows:\n",
    "crashes =", nrow(crashes), "\n",
    "parties =", nrow(parties), "\n",
    "victims =", nrow(victims), "\n")

cat("Unique CASE_ID:\n",
    "crashes =", n_distinct(crashes$CASE_ID), "\n",
    "parties =", n_distinct(parties$CASE_ID), "\n",
    "victims =", n_distinct(victims$CASE_ID), "\n")
```

# Chech crash table grain (should be 1 row per crash)
```{r}
# What: Verify crashes table is already crash-grain (1 row per CASE_ID).
# Why: If crashes has duplicates, all downstream “one row per crash” logic breaks.

dup_crashes <- crashes %>%
  count(CASE_ID, name = "n") %>%
  filter(n > 1)

cat("Number of duplicated CASE_ID in crashes:", nrow(dup_crashes), "\n")

# If you want to inspect duplicates:
head(dup_crashes, 10)
```

# Standardize CASE_ID type across all tables
```{r}
# What: Standardize CASE_ID as character in all tables.
# Why: Joins can silently fail or misbehave if CASE_ID types differ (int vs char).

crashes <- crashes %>% mutate(CASE_ID = as.character(CASE_ID))
parties <- parties %>% mutate(CASE_ID = as.character(CASE_ID))
victims <- victims %>% mutate(CASE_ID = as.character(CASE_ID))
```

# Quick missing-join diagnostics (who's missing from what)
```{r}
# What: Identify CASE_IDs present in parties/victims but missing from crashes (and vice versa).
# Why: This tells you if joins will drop information due to key mismatches.

cases_in_parties_not_crashes <- setdiff(unique(parties$CASE_ID), unique(crashes$CASE_ID))
cases_in_victims_not_crashes <- setdiff(unique(victims$CASE_ID), unique(crashes$CASE_ID))

cat("CASE_IDs in parties but not in crashes:", length(cases_in_parties_not_crashes), "\n")
cat("CASE_IDs in victims but not in crashes:", length(cases_in_victims_not_crashes), "\n")

cases_in_crashes_not_parties <- setdiff(unique(crashes$CASE_ID), unique(parties$CASE_ID))
cases_in_crashes_not_victims <- setdiff(unique(crashes$CASE_ID), unique(victims$CASE_ID))

cat("Crashes with no party rows:", length(cases_in_crashes_not_parties), "\n")
cat("Crashes with no victim rows:", length(cases_in_crashes_not_victims), "\n")
```

# Make party-level aggregates (crash-level features)
```{r}
# What: Aggregate parties to crash-level (1 row per CASE_ID).
# Why: Prevents 1-to-many join explosion; produces interpretable crash-level predictors.

# Identify party id column if it exists; if not, fall back to counting rows.
safe_any_code <- function(x, yes_code, notstated_codes = c(NA, "", "-", "Not Stated")) {
  x <- trimws(as.character(x))
  if (any(x == yes_code, na.rm = TRUE)) return("Y")
  if (all(x %in% notstated_codes)) return("Not Stated")
  "N"
}
party_id_col <- if ("PARTY_NUMBER" %in% names(parties)) "PARTY_NUMBER" else NA_character_

parties_agg <- parties %>%
  group_by(CASE_ID) %>%
  summarise(
    n_parties = if (!is.na(party_id_col)) n_distinct(.data[[party_id_col]], na.rm = TRUE) else n(),

    any_party_alcohol = if ("PARTY_SOBRIETY" %in% names(parties))
      safe_any_code(PARTY_SOBRIETY, "B")
    else NA_character_,

    any_party_drug = if ("PARTY_DRUG_PHYSICAL" %in% names(parties))
      safe_any_code(PARTY_DRUG_PHYSICAL, "E")
    else NA_character_,

    any_party_at_fault = if ("AT_FAULT" %in% names(parties))
      safe_any_code(AT_FAULT, "Y")
    else NA_character_,

    mean_party_age = if ("PARTY_AGE" %in% names(parties)) {
      age <- suppressWarnings(as.numeric(PARTY_AGE))
      age[age %in% c(998, 999)] <- NA_real_
      if (all(is.na(age))) NA_real_ else mean(age, na.rm = TRUE)
    } else NA_real_,

    .groups = "drop"
  ) %>%
  mutate(
    any_party_alcohol  = factor(any_party_alcohol,  levels = c("N","Y","Not Stated")),
    any_party_drug     = factor(any_party_drug,     levels = c("N","Y","Not Stated")),
    any_party_at_fault = factor(any_party_at_fault, levels = c("N","Y","Not Stated"))
  )

dim(parties_agg)
head(parties_agg)
```

# Make victim-level aggregates (crash-level features)
```{r}
# What: Aggregate victims to crash-level (1 row per CASE_ID).
# Why: Victim table is also 1-to-many; we need crash-level summaries for modeling/EDA.

victim_id_col <- if ("VICTIM_ID" %in% names(victims)) "VICTIM_ID" else NA_character_

victims_agg <- victims %>%
  group_by(CASE_ID) %>%
  summarise(
    n_victims = if (!is.na(victim_id_col)) n_distinct(.data[[victim_id_col]], na.rm = TRUE) else n(),

    n_victims_killed = if ("VICTIM_DEGREE_OF_INJURY" %in% names(victims)) {
      sum(VICTIM_DEGREE_OF_INJURY == "1", na.rm = TRUE)
    } else NA_integer_,

    any_victim_ksi = if ("VICTIM_DEGREE_OF_INJURY" %in% names(victims)) {
  any(VICTIM_DEGREE_OF_INJURY %in% c("1","2","5"), na.rm = TRUE)
} else NA,

    min_victim_age = if ("VICTIM_AGE" %in% names(victims)) {
      age <- suppressWarnings(as.numeric(VICTIM_AGE))
      age[age %in% c(998, 999)] <- NA_real_
      if (all(is.na(age))) NA_real_ else min(age, na.rm = TRUE)
    } else NA_real_,

    .groups = "drop"
  )

dim(victims_agg)
head(victims_agg)
```

# Merge to crashes
```{r}
# What: Left-join crash-level party/victim aggregates onto crashes.
# Why: Preserves crash grain (same number of rows as crashes) while adding richer features.

merged_fixed <- crashes %>%
  left_join(parties_agg, by = "CASE_ID") %>%
  left_join(victims_agg, by = "CASE_ID")

# Must still be crash-grain
stopifnot(nrow(merged_fixed) == nrow(crashes))
stopifnot(n_distinct(merged_fixed$CASE_ID) == nrow(crashes))

dim(merged_fixed)
```

# Post-merge QA checks 
```{r}
# What: Sanity-check that merge didn’t create duplicates and that aggregates look reasonable.
# Why: Prevents subtle grain bugs that break EDA/modeling.

cat("Any duplicated CASE_ID after merge? ", anyDuplicated(merged_fixed$CASE_ID) > 0, "\n")

# Compare crash-reported killed vs victim-derived killed (not necessarily identical, but should be correlated)
if (all(c("NUMBER_KILLED","n_victims_killed") %in% names(merged_fixed))) {
  cat("Correlation between crash NUMBER_KILLED and victim n_victims_killed:\n")
  print(cor(merged_fixed$NUMBER_KILLED, merged_fixed$n_victims_killed, use = "complete.obs"))
}

# Quick summaries
summary(select(merged_fixed, n_parties, n_victims))

if ("PARTY_COUNT" %in% names(merged_fixed)) {
  cat("PARTY_COUNT (crashes) minus n_parties (parties_agg):\n")
  print(table(merged_fixed$PARTY_COUNT - merged_fixed$n_parties, useNA = "ifany"))
}
```

```{r}
table(merged_fixed$ROAD_SURFACE, useNA="ifany")
# --- Validate ROAD_SURFACE against codebook ---

valid_road_surface <- c("A","B","C","D","-")

merged_fixed <- merged_fixed %>%
  mutate(
    ROAD_SURFACE = trimws(ROAD_SURFACE),
    ROAD_SURFACE = ifelse(is.na(ROAD_SURFACE) | ROAD_SURFACE == "", "-", ROAD_SURFACE),
    ROAD_SURFACE_invalid = ifelse(
      is.na(ROAD_SURFACE) | ROAD_SURFACE %in% valid_road_surface,
      0L, 1L
    ),

    ROAD_SURFACE_raw_invalid = ifelse(
      ROAD_SURFACE_invalid == 1L,
      ROAD_SURFACE,
      NA_character_
    ),

    ROAD_SURFACE = ifelse(
      ROAD_SURFACE_invalid == 1L,
      "-",
      ROAD_SURFACE
    )
  )

cat("Invalid ROAD_SURFACE values found:\n")
print(table(merged_fixed$ROAD_SURFACE_raw_invalid, useNA="ifany"))

check_codes <- function(data, var, valid_codes) {
  invalid <- setdiff(unique(data[[var]]), valid_codes)
  invalid[!is.na(invalid)]
}
table(merged_fixed$ROAD_SURFACE, useNA="ifany")
table(merged_fixed$ROAD_SURFACE_raw_invalid, useNA="ifany")

merged_fixed <- merged_fixed %>%
  mutate(INTERSECTION = ifelse(is.na(INTERSECTION) | INTERSECTION == "", "-", INTERSECTION))

check_codes(merged_fixed, "WEATHER_1", c("A","B","C","D","E","F","G","-"))
check_codes(merged_fixed, "LIGHTING",  c("A","B","C","D","E","-"))
check_codes(merged_fixed, "INTERSECTION", c("Y","N", "-"))
```

```{r}
check_codes(merged_fixed, "ROAD_COND_1", c("A","B","C","D","E","F","G","H","-",""))
check_codes(merged_fixed, "ROAD_COND_2", c("A","B","C","D","E","F","G","H","-",""))
```

```{r}
all_equal_killed <- all(merged_fixed$NUMBER_KILLED == merged_fixed$n_victims_killed, na.rm = TRUE)
all_equal_killed

# If FALSE, inspect differences
table(merged_fixed$NUMBER_KILLED - merged_fixed$n_victims_killed, useNA = "ifany")
```

# Save the merged dataset
```{r}
# What: Save crash-grain merged dataset.
# Why: Makes analysis reproducible and avoids repeating expensive work.

write.csv(merged_fixed, "merged_fixed_switrs.csv", row.names = FALSE)
```
