---
title: "R Notebook"
output: html_notebook
---

# Load in data and view summary of each variable
```{r}
getwd()
crashes <- read.csv(
  "~/Desktop/STAT544/Crashes.csv",
  na.strings = c("", " ", "  ", "   ", "-", "NA", "N/A", "NULL", "null", "Unknown", "UNKNOWN", "UNK")
)
```

#view summary and the dimension of the data (168,332 rows and 80 variables)
```{r}
dim(crashes)

summary(crashes)
```

#check for duplicates: all are unique
```{r}
duplicates_num = sum(duplicated(crashes)) 
duplicates_num
```

#Finding the percentage of missing info from each variable.
# For the variables missing 30% or more of the data, we will delete. (we deleted 5 variables)
```{r}
# percentage missing for each column
missing_pct <- colMeans(is.na(crashes)) * 100

# look at them, sorted high → low
sort(round(missing_pct, 2), decreasing = TRUE)

# choose a threshold for "too much" missingness
high_miss_threshold <- 30  # percent

# names of variables with > 30% missing
high_miss_vars <- names(missing_pct)[missing_pct > high_miss_threshold]

high_miss_vars   # just to see which ones they are

crashes_red <- crashes[, !(names(crashes) %in% high_miss_vars)]
```

#let's look at frequency tables/bar charts

#correlation
```{r}

```


#skewness

#outliers

#Define severe accidents
```{r}
crashes_red$severe_binary <- ifelse(
  crashes_red$COLLISION_SEVERITY %in% 
    c("FATAL INJURY", "SUSPECTED SERIOUS INJURY OR SEVERE INJURY"),
  1, 0
)
```


#find SIGNIFICANT FEATURES USING T-TESTS or hypothesis tests
