---
title: "STAT 544 - SWITRS Crash Severity Notebook"
output: html_notebook
---
# Package Chunk
```{r setup, message=FALSE, warning=FALSE}
# What: Global setup (packages, options, theme)
# Why: Reproducibility and consistent output formatting across the whole notebook.

options(stringsAsFactors = FALSE)
set.seed(544)

required_pkgs <- c(
  "dplyr", "tidyr", "stringr", "ggplot2", "scales", "purrr",
  "readr", "forcats", "corrplot", "broom", "rlang"
)

installed <- rownames(installed.packages())
to_install <- setdiff(required_pkgs, installed)
if (length(to_install) > 0) install.packages(to_install)

library(rlang)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(scales)
library(purrr)
library(readr)
library(forcats)
library(corrplot)
library(broom)

theme_set(theme_minimal(base_size = 12))
```

# Load data
```{r}
# What: Load the crash dataset with MINIMAL NA conversion.
# Why: In SWITRS-style coded data, "-" often means "Not Stated" (meaningful category),
# and blanks can be structural (e.g., Y/blank flags). If we convert these at import,
# we lose the ability to correctly distinguish missing vs not stated vs structural blank.

# Only treat explicit NA tokens as NA at import (do NOT include "" or "-" here)
na_tokens <- c("NA", "N/A", "NULL", "null")

crashes_raw <- read.csv("Crashes.csv", na.strings = na_tokens, stringsAsFactors = FALSE)

dim(crashes_raw)
glimpse(crashes_raw)
```

# Quick Data Audit
```{r}
# What: Basic audit (duplicates, missingness profile).
# Why: Catch data integrity issues early; document missingness before transformations.

duplicates_num <- sum(duplicated(crashes_raw))
duplicates_num

table(crashes_raw$COLLISION_SEVERITY, useNA = "ifany")

missing_pct_raw <- sort(colMeans(is.na(crashes_raw)) * 100, decreasing = TRUE)
head(round(missing_pct_raw, 2), 20)
```

# Standardize Strings + "Not Stated" Handling
```{r}
# What: Standardize strings, preserve "-" as "Not Stated", then define remaining blanks as NA.
# Why: This is the key ordering fix:
#      1) Trim whitespace
#      2) Convert dash-coded "-" to "Not Stated" (explicit category)
#      3) Handle structural blanks (Y/blank flags) separately
#      4) Only then treat remaining "" as true missing NA

crashes_clean <- crashes_raw %>%
  mutate(across(where(is.character), ~ str_trim(.x))) %>%
  mutate(across(where(is.character), ~ ifelse(.x == "-", "Not Stated", .x)))

# Sanity check: severity distribution after "-" -> "Not Stated"
if ("COLLISION_SEVERITY" %in% names(crashes_clean)) {
  table(crashes_clean$COLLISION_SEVERITY, useNA = "ifany")
}

# Helper: turn NA into explicit "Not Stated" for selected categorical variables
na_to_not_stated <- function(x) {
  x <- as.character(x)
  ifelse(is.na(x), "Not Stated", x)
}
```

#Treat "UNKNOWN" road names as missing
```{r}
# What: Convert literal "UNKNOWN" placeholders to missing (NA), especially for roadway name fields.
# Why: The raw CSV often uses "UNKNOWN" as a filler for missing road names; treating it as a real category
#      inflates "valid" categories and underestimates missingness.

road_name_vars <- intersect(c("PRIMARY_RD", "SECONDARY_RD"), names(crashes_clean))

crashes_clean <- crashes_clean %>%
  mutate(across(all_of(road_name_vars), ~ {
    x <- as.character(.x)
    x <- str_trim(x)
    x[toupper(x) == "UNKNOWN"] <- NA_character_
    x
  }))
```

# Convert Y/N/Y-Only Flags Carefully
```{r}
# What: Convert binary flags to 0/1 with consistent rules.
# Why: Some columns are Y/blank (blank => 0), others are Y/N/blank (blank => NA).
#      We must do this BEFORE converting "" to NA globally, because blanks are meaningful in Y/blank fields.

# A) Columns that are Y/blank meaning "Y else 0"
yn_blank_vars <- c("PEDESTRIAN_ACCIDENT",
                   "BICYCLE_ACCIDENT",
                   "MOTORCYCLE_ACCIDENT",
                   "TRUCK_ACCIDENT",
                   "NOT_PRIVATE_PROPERTY",
                   "ALCOHOL_INVOLVED")

yn_blank_vars <- yn_blank_vars[yn_blank_vars %in% names(crashes_clean)]

crashes_clean[yn_blank_vars] <- lapply(crashes_clean[yn_blank_vars], function(x) {
  x <- toupper(str_trim(as.character(x)))
  # Y -> 1, blank/NA/other -> 0 (structural No)
  as.integer(!is.na(x) & x == "Y")
})

# B) Detect columns that are strictly Y/N (possibly with NAs) and map Y->1, N->0, blank=>NA
yn_cols <- names(crashes_clean)[
  sapply(crashes_clean, function(x) {
    if (is.character(x) || is.factor(x)) {
      vals <- unique(na.omit(toupper(str_trim(as.character(x)))))
      length(vals) > 0 && all(vals %in% c("Y","N"))
    } else FALSE
  })
]

# Avoid double-handling Y/blank columns
yn_cols <- setdiff(yn_cols, yn_blank_vars)

crashes_clean[yn_cols] <- lapply(crashes_clean[yn_cols], function(x) {
  x <- toupper(str_trim(as.character(x)))
  # For true Y/N fields, blank should be NA (missing), not 0
  x[x == ""] <- NA
  ifelse(is.na(x), NA_integer_,
         ifelse(x == "Y", 1L,
                ifelse(x == "N", 0L, NA_integer_)))
})

list(
  yn_blank_vars = yn_blank_vars,
  yn_cols = yn_cols
)
```

# Define remaining blanks as NA (after Y/blank handled)
```{r}
# What: Convert remaining empty strings "" to NA for all character fields.
# Why: Now that dash-coded not stated is preserved and Y/blank fields are already converted to 0/1,
#      remaining blanks are best treated as true missing.

crashes_clean <- crashes_clean %>%
  mutate(across(where(is.character), ~ na_if(.x, "")))
```

# Collision time: remove unknown codes (2500-2559)
```{r}
# What: Remove unknown time codes (2500–2559) and then derive hour safely.
# Why: SWITRS encodes unknown collision time as 2500–2559. Keeping these corrupts time analyses.

if ("COLLISION_TIME" %in% names(crashes_clean)) {
  crashes_clean <- crashes_clean %>%
  mutate(
    COLLISION_TIME = suppressWarnings(as.integer(as.character(COLLISION_TIME))),
    COLLISION_TIME = ifelse(!is.na(COLLISION_TIME) & COLLISION_TIME >= 2500 & COLLISION_TIME <= 2559,
                            NA_integer_, COLLISION_TIME)
  )
}

# Helper to parse hour from HHMM after unknowns removed
parse_hour <- function(x) {
  x <- suppressWarnings(as.integer(as.character(x)))
  hh <- ifelse(is.na(x), NA_integer_, x %/% 100)
  hh[hh < 0 | hh > 23] <- NA_integer_
  hh
}

if ("COLLISION_TIME" %in% names(crashes_clean)) {
  crashes_clean <- crashes_clean %>%
    mutate(
      hour = parse_hour(COLLISION_TIME),
      time_bucket = case_when(
        is.na(hour) ~ "Not Stated",
        hour %in% 0:5 ~ "Late Night (12–5)",
        hour %in% 6:9 ~ "AM Commute (6–9)",
        hour %in% 10:15 ~ "Midday (10–3)",
        hour %in% 16:19 ~ "PM Commute (4–7)",
        hour %in% 20:23 ~ "Evening (8–11)",
        TRUE ~ "Not Stated"
      )
    )
}
```

# Missingness diagnostics aligned with class notes
```{r missingness-diagnostics}
# What: Create a missingness audit that separates TRUE missing (NA) from "Not Stated".
# Why: SWITRS often encodes nonresponse as "-" (which we convert to "Not Stated") rather than NA.
#      Class notes emphasize identifying missingness mechanisms (MCAR/MAR/MNAR) before dropping variables. 

missing_audit <- dplyr::tibble(
  var = names(crashes_clean),
  pct_na = sapply(crashes_clean, function(x) mean(is.na(x)) * 100),
  pct_not_stated = sapply(crashes_clean, function(x) {
    if (is.character(x) || is.factor(x)) mean(as.character(x) == "Not Stated", na.rm = TRUE) * 100 else NA_real_
  }),
  n_unique = sapply(crashes_clean, function(x) length(unique(na.omit(x))))
) %>%
  arrange(desc(pct_na))

head(missing_audit, 25)
```

# Remove High-Missing Variables
```{r}
## NOTE: high_miss_threshold applies to TRUE NA (colMeans(is.na(...))),
# not to "Not Stated". We preserve "Not Stated" as an informative level.
# What: Drop columns with >30% missingness (rule-based).
# Why: Matches your current approach; applied consistently after correct missingness handling.

high_miss_threshold <- 30
missing_pct <- colMeans(is.na(crashes_clean)) * 100
high_miss_vars <- names(missing_pct)[missing_pct > high_miss_threshold]

high_miss_vars
length(high_miss_vars)

crashes_reduced <- crashes_clean %>%
  select(-all_of(high_miss_vars))

# Preserve raw-coded outcome before any label recodes (codebook alignment)
# COLLISION_SEVERITY is coded: 0,1,2,3,4 and sometimes "-" (we converted "-" to "Not Stated" earlier)
if ("COLLISION_SEVERITY" %in% names(crashes_reduced)) {
  crashes_reduced <- crashes_reduced %>%
    mutate(
      COLLISION_SEVERITY_CODE = as.character(COLLISION_SEVERITY),
      COLLISION_SEVERITY_CODE = str_trim(COLLISION_SEVERITY_CODE),
      COLLISION_SEVERITY_CODE = ifelse(COLLISION_SEVERITY_CODE %in% c("", "Not Stated"), NA, COLLISION_SEVERITY_CODE)
    )
}

dim(crashes_reduced)
```

# Drop Deprecated / Non-Analytic Variables
```{r}
# What: Drop IDs or non-analytic variables you previously agreed to remove.
# Why: Reduce leakage and clutter; keep CASE_ID only for potential merges.
#
# IMPORTANT FIX:
# - Do NOT drop CHP_SHIFT / BEAT_TYPE / CHP_BEAT_CLASS before evaluating signal.
#   We keep them for now and screen them later.

# Derive year for base-rate plots BEFORE dropping ACCIDENT_YEAR
if (!("year" %in% names(crashes_reduced)) && "ACCIDENT_YEAR" %in% names(crashes_reduced)) {
  crashes_reduced <- crashes_reduced %>%
    mutate(year = suppressWarnings(as.integer(as.character(ACCIDENT_YEAR))))
}

vars_to_remove <- c(
  "ACCIDENT_YEAR",
  "PROC_DATE",
  "OFFICER_ID",
  "ROUTE_SUFFIX",
  "POSTMILE_PREFIX",
  "REPORTING_DISTRICT",
  # "CHP_SHIFT",         # kept for screening
  "POPULATION",
  "CALTRANS_COUNTY",
  "CALTRANS_DISTRICT",
  "LATITUDE",
  "LONGITUDE",
  "POINT_X",
  "POINT_Y",
  "COUNTY"
)

vars_to_remove <- intersect(vars_to_remove, names(crashes_reduced))

crashes_reduced <- crashes_reduced %>%
  select(-all_of(vars_to_remove))

dim(crashes_reduced)
```

# Recode SWITRS Coded Fields (Safe Mappings)
```{r}
# What: Apply SWITRS codebook label mappings for key coded fields.
# Why: Improves interpretability and supports client-ready EDA.
#
# IMPORTANT FIX:
# - Keep original classes if you want; here we map to readable labels.
# - We preserve "Not Stated" explicitly via na_to_not_stated() afterwards.

recode_maps <- list(
  DAY_OF_WEEK = c(
    "1" = "Monday","2" = "Tuesday","3" = "Wednesday","4" = "Thursday",
    "5" = "Friday","6" = "Saturday","7" = "Sunday"
  ),
  DIRECTION = c("N" = "North","E" = "East","S" = "South","W" = "West"),
  INTERSECTION = c("Y" = "Intersection","N" = "Not Intersection"),
  STATE_HWY_IND = c("Y" = "State Highway","N" = "Not State Highway"),

  WEATHER_1 = c("A" = "Clear","B" = "Cloudy","C" = "Raining","D" = "Snowing","E" = "Fog","F" = "Other","G" = "Wind"),
  WEATHER_2 = c("A" = "Clear","B" = "Cloudy","C" = "Raining","D" = "Snowing","E" = "Fog","F" = "Other","G" = "Wind"),

  LOCATION_TYPE = c("H" = "Highway","I" = "Intersection","R" = "Ramp (or Collector)"),

  TOW_AWAY = c("Y" = "Yes","N" = "No"),

  COLLISION_SEVERITY = c(
    "1" = "Fatal Injury",
    "2" = "Suspected Serious Injury / Severe Injury",
    "3" = "Suspected Minor Injury / Visible Injury",
    "4" = "Possible Injury / Complaint of Pain",
    "0" = "No Injury (PDO)"
  ),

  PRIMARY_COLL_FACTOR = c(
    "A" = "Code Violation",
    "B" = "Other Improper Driving",
    "C" = "Other Than Driver",
    "D" = "Unknown",
    "E" = "Fell Asleep"
  ),

  PCF_VIOL_CATEGORY = c(
    "01"="DUI (Alcohol/Drug)", "1"="DUI (Alcohol/Drug)",
    "02"="Impeding Traffic",   "2"="Impeding Traffic",
    "03"="Unsafe Speed",       "3"="Unsafe Speed",
    "04"="Following Too Closely","4"="Following Too Closely",
    "05"="Wrong Side of Road", "5"="Wrong Side of Road",
    "06"="Improper Passing",   "6"="Improper Passing",
    "07"="Unsafe Lane Change", "7"="Unsafe Lane Change",
    "08"="Improper Turning",   "8"="Improper Turning",
    "09"="Auto Right-of-Way",  "9"="Auto Right-of-Way",
    "10"="Pedestrian Right-of-Way",
    "11"="Pedestrian Violation",
    "12"="Traffic Signals & Signs",
    "13"="Hazardous Parking",
    "14"="Lights",
    "15"="Brakes",
    "16"="Other Equipment",
    "17"="Other Hazardous Violation",
    "18"="Other Than Driver (or Pedestrian)",
    "21"="Unsafe Starting/Backing",
    "22"="Other Improper Driving",
    "23"="Pedestrian/Other DUI (Alcohol/Drug)",
    "24"="Fell Asleep",
    "00"="Unknown", "0"="Unknown"
  ),

  HIT_AND_RUN = c("F"="Felony","M"="Misdemeanor","N"="Not Hit and Run"),

  TYPE_OF_COLLISION = c(
    "A" = "Head-On","B" = "Sideswipe","C" = "Rear End","D" = "Broadside",
    "E" = "Hit Object","F" = "Overturned","G" = "Vehicle/Pedestrian","H" = "Other"
  ),

  ROAD_SURFACE = c("A" = "Dry","B" = "Wet","C" = "Snowy or Icy","D" = "Slippery (Muddy/Oily/etc.)"),

  LIGHTING = c(
    "A" = "Daylight",
    "B" = "Dusk/Dawn",
    "C" = "Dark - Street Lights",
    "D" = "Dark - No Street Lights",
    "E" = "Dark - Street Lights Not Functioning"
  ),

  CHP_SHIFT = c(
    "1" = "0600–1359",
    "2" = "1400–2159",
    "3" = "2200–0559",
    "4" = "CHP Not Stated",
    "5" = "Not CHP"
  ),
  
  BEAT_TYPE = c(
    "1" = "CHP State Highway",
    "2" = "CHP County Road Line",
    "3" = "CHP County Road Area",
    "4" = "Schoolbus on City Roadway (Adm Beat 901)",
    "5" = "Schoolbus not on Public Roadway (Adm Beat 903)",
    "6" = "Offroad (Unimproved) (Adm Beat 906/907)",
    "7" = "Vista Point/Rest Area or Scales/Inspection Facility",
    "8" = "Other Public Access (Improved)",
    "0" = "Not CHP"
  ),
  
  CHP_BEAT_CLASS = c(
    "1" = "CHP Primary",
    "2" = "CHP Other",
    "0" = "Not CHP"
  )
)

# Do NOT overwrite the coded COLLISION_SEVERITY; create a labeled copy instead
if ("COLLISION_SEVERITY" %in% names(recode_maps)) {
  recode_maps$COLLISION_SEVERITY_LAB <- recode_maps$COLLISION_SEVERITY
  recode_maps$COLLISION_SEVERITY <- NULL
}

# Ensure the labeled severity column exists prior to applying recodes
if ("COLLISION_SEVERITY_CODE" %in% names(crashes_reduced)) {
  crashes_reduced <- crashes_reduced %>%
    mutate(COLLISION_SEVERITY_LAB = COLLISION_SEVERITY_CODE)
}

apply_recode <- function(df, var, map) {
  if (!var %in% names(df)) return(df)
  df %>%
    mutate(!!rlang::sym(var) := dplyr::recode(
      as.character(.data[[var]]),
      !!!map,
      .default = as.character(.data[[var]])
    ))
}

crashes_labeled <- crashes_reduced
for (v in names(recode_maps)) {
  crashes_labeled <- apply_recode(crashes_labeled, v, recode_maps[[v]])
}

# For key coded categoricals, keep "Not Stated" explicitly
key_cat_vars <- intersect(c(names(recode_maps), "COLLISION_SEVERITY_LAB"), names(crashes_labeled))

crashes_labeled <- crashes_labeled %>%
  mutate(across(all_of(key_cat_vars), ~ na_to_not_stated(.x)))

# Prevent coded categoricals from being treated as numeric in models
if ("DAY_OF_WEEK" %in% names(crashes_labeled)) {
  crashes_labeled$DAY_OF_WEEK <- factor(crashes_labeled$DAY_OF_WEEK,
                                       levels = c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday","Not Stated"))
}
```

# Define Target (Severe/Fatal)
```{r}
# What: Create the binary target variable.
# Why: Your proposal defines severe/fatal as COLLISION_SEVERITY in {1,2}.

# Define Target (Severe/Fatal) using codebook-coded values (1–2)
crashes_model <- crashes_labeled %>%
  mutate(
    COLLISION_SEVERITY_CODE = as.character(COLLISION_SEVERITY_CODE),

    SevereBinary = case_when(
      COLLISION_SEVERITY_CODE %in% c("1", "2") ~ "Severe/Fatal",
      COLLISION_SEVERITY_CODE %in% c("0", "3", "4") ~ "Other",
      TRUE ~ NA_character_   # includes "-" Not Stated / blanks / anything unexpected
    ),
    SevereBinary = factor(SevereBinary, levels = c("Other", "Severe/Fatal"))
  )

# Sanity checks (these catch “numeric(0)” / drift bugs fast)
table(crashes_model$COLLISION_SEVERITY_CODE, useNA = "ifany")
table(crashes_model$COLLISION_SEVERITY_LAB, useNA = "ifany")
table(crashes_model$SevereBinary, useNA = "ifany")
```

# Missingness diagnostics
```{r}
# Missingness diagnostics (overall + by outcome)
missing_overall <- tibble(
  variable = names(crashes_model),
  missing_pct = colMeans(is.na(crashes_model)) * 100
) %>%
  arrange(desc(missing_pct))

head(missing_overall, 20)

# Missingness differs by outcome? (MAR warning sign)
# For each variable, compare missing % between Severe/Fatal vs Other
missing_by_outcome <- map_dfr(names(crashes_model), function(v) {
  x <- crashes_model[[v]]
  tibble(
    variable = v,
    miss_other = mean(is.na(x[crashes_model$SevereBinary == "Other"]), na.rm = TRUE),
    miss_severe = mean(is.na(x[crashes_model$SevereBinary == "Severe/Fatal"]), na.rm = TRUE)
  )
}) %>%
  mutate(
    diff = miss_severe - miss_other
  ) %>%
  arrange(desc(abs(diff)))

head(missing_by_outcome, 20)
```

# Client ready base-rate plot
```{r}
# Base rate check (severity share by year)
if ("year" %in% names(crashes_model)) {
  crashes_model %>%
    filter(!is.na(SevereBinary)) %>%
    group_by(year) %>%
    summarize(severe_rate = mean(SevereBinary == "Severe/Fatal"), n = n(), .groups = "drop") %>%
    ggplot(aes(x = year, y = severe_rate)) +
    geom_line() +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(
      title = "Severe/Fatal Crash Share by Year",
      x = "Year",
      y = "Severe/Fatal Share"
    )
}
```


# Feature Engineering (Time + Interactions)
```{r}
# What: Create derived time features and interaction flags.
# Why: The proposal emphasizes time-of-day, weekend, and interactions like rain×dark, alcohol×night, ped×dark.

# What: Carry forward hour/time_bucket created in crashes_clean rather than re-parsing collision time.
# Why: Ensures time features reflect the same cleaning rule (including dropping unknown 2500–2559).

# If hour/time_bucket exist, they should already be in crashes_labeled -> crashes_model through the pipeline.
# This chunk is a no-op guard to ensure they exist.
if (!("hour" %in% names(crashes_model)) && "COLLISION_TIME" %in% names(crashes_model)) {
  crashes_model <- crashes_model %>%
    mutate(
      hour = parse_hour(COLLISION_TIME),
      time_bucket = case_when(
        is.na(hour) ~ "Not Stated",
        hour %in% 0:5 ~ "Late Night (12–5)",
        hour %in% 6:9 ~ "AM Commute (6–9)",
        hour %in% 10:15 ~ "Midday (10–3)",
        hour %in% 16:19 ~ "PM Commute (4–7)",
        hour %in% 20:23 ~ "Evening (8–11)",
        TRUE ~ "Not Stated"
      )
    )
}

if ("DAY_OF_WEEK" %in% names(crashes_model)) {
  crashes_model <- crashes_model %>%
    mutate(
      weekend = ifelse(DAY_OF_WEEK %in% c("Saturday","Sunday"), 1,
                       ifelse(DAY_OF_WEEK == "Not Stated", NA, 0))
    )
}

# Night indicator from LIGHTING (preferred if present)
if ("LIGHTING" %in% names(crashes_model)) {
  crashes_model <- crashes_model %>%
    mutate(
      night = ifelse(LIGHTING %in% c("Dark - Street Lights",
                                    "Dark - No Street Lights",
                                    "Dark - Street Lights Not Functioning"), 1,
                     ifelse(LIGHTING == "Not Stated", NA, 0))
    )
}

# Weather rain indicator (uses WEATHER_1 if present)
if ("WEATHER_1" %in% names(crashes_model)) {
  crashes_model <- crashes_model %>%
    mutate(
      raining = ifelse(WEATHER_1 == "Raining", 1,
                       ifelse(WEATHER_1 == "Not Stated", NA, 0))
    )
}

# Pedestrian indicator: if you created PEDESTRIAN_ACCIDENT earlier, it is already 0/1
if ("PEDESTRIAN_ACCIDENT" %in% names(crashes_model)) {
  crashes_model <- crashes_model %>%
    mutate(ped = PEDESTRIAN_ACCIDENT)
}

# Alcohol indicator: if you created ALCOHOL_INVOLVED earlier, it is already 0/1
if ("ALCOHOL_INVOLVED" %in% names(crashes_model)) {
  crashes_model <- crashes_model %>%
    mutate(alcohol = ALCOHOL_INVOLVED)
}

# Interaction flags
crashes_model <- crashes_model %>%
  mutate(
    rain_x_dark = ifelse(!is.na(raining) & !is.na(night), as.integer(raining == 1 & night == 1), NA),
    alcohol_x_night = ifelse(!is.na(alcohol) & !is.na(night), as.integer(alcohol == 1 & night == 1), NA),
    ped_x_dark = ifelse(!is.na(ped) & !is.na(night), as.integer(ped == 1 & night == 1), NA)
  )

glimpse(crashes_model)
```

# EDA: Rate Calculations (not just counts)
```{r}
# What: Compute severe/fatal rates overall and by key dimensions.
# Why: Rates (e.g., severe per 100 crashes) are the right comparison metric for decision-making.

overall_rates <- crashes_model %>%
  summarise(
    crashes_n = n(),
    severe_n = sum(SevereBinary == "Severe/Fatal", na.rm = TRUE),
    severe_rate = severe_n / crashes_n,
    severe_per_100 = 100 * severe_rate
  )

overall_rates

# By time bucket, if present
if ("time_bucket" %in% names(crashes_model)) {
  rate_by_time_bucket <- crashes_model %>%
    filter(!is.na(time_bucket)) %>%
    group_by(time_bucket) %>%
    summarise(
      crashes_n = n(),
      severe_n = sum(SevereBinary == "Severe/Fatal", na.rm = TRUE),
      severe_per_100 = 100 * severe_n / crashes_n,
      .groups = "drop"
    ) %>%
    arrange(desc(crashes_n))
  
  rate_by_time_bucket
}
```

# Screen CHP variables before deciding to drop
```{r}
# What: Evaluate CHP_SHIFT / BEAT_TYPE / CHP_BEAT_CLASS by severe rate and sample size.
# Why: Prevent dropping potentially informative predictors prematurely.

screen_rates <- function(df, var) {
  if (!var %in% names(df)) return(NULL)
  df %>%
    mutate(v = as.character(.data[[var]]),
           v = ifelse(is.na(v), "Not Stated", v)) %>%
    group_by(v) %>%
    summarise(
      n = n(),
      severe_n = sum(SevereBinary == "Severe/Fatal", na.rm = TRUE),
      severe_per_100 = 100 * severe_n / n,
      .groups = "drop"
    ) %>%
    arrange(desc(n))
}

candidate_vars <- intersect(c("CHP_SHIFT","BEAT_TYPE","CHP_BEAT_CLASS"), names(crashes_model))
rates_by_candidate <- lapply(candidate_vars, function(v) screen_rates(crashes_model, v))
names(rates_by_candidate) <- candidate_vars

rates_by_candidate
```

# EDA: Decision-Oriented Severity Rate Plots (Ordered, Horizontal)
```{r}
# What: Plot Severe/Fatal share by categorical variables (proportions).
# Why: Client-ready: compare risk via proportions rather than raw counts.

library(dplyr)
library(ggplot2)
library(scales)
library(purrr)
library(tidyr)

exclude_vars <- intersect(c("CASE_ID", "SevereBinary", "COLLISION_SEVERITY"), names(crashes_model))

multi_class_vars <- names(crashes_model)[
  sapply(crashes_model, function(x) {
    (is.character(x) || is.factor(x)) && length(unique(na.omit(x))) > 2
  })
]
multi_class_vars <- setdiff(multi_class_vars, exclude_vars)

crashes_model[multi_class_vars] <- lapply(crashes_model[multi_class_vars], as.factor)

plot_severity_by_variable <- function(var_name, data = crashes_model, fill_var = "SevereBinary") {

  sev_levels <- c("Other", "Severe/Fatal")

  df <- data %>%
    mutate(
      .x = .data[[var_name]],
      .fill = factor(.data[[fill_var]], levels = sev_levels)
    ) %>%
    filter(!is.na(.x), !is.na(.fill)) %>%
    count(.x, .fill, name = "n") %>%
    group_by(.x) %>%
    mutate(
      n_total = sum(n),
      prop = n / n_total
    ) %>%
    ungroup()

  totals <- df %>% distinct(.x, n_total)

  ggplot(df, aes(x = .x, y = prop, fill = .fill)) +
    geom_col(width = 0.8, position = "stack") +
    coord_flip(clip = "off") +

    # % labels centered in each stacked segment (ggplot computes the midpoint)
    geom_text(
      aes(label = percent(prop, accuracy = 1)),
      position = position_stack(vjust = 0.5),
      size = 3,
      color = "white"
    ) +

    # n labels on the RIGHT side of each bar
    geom_text(
      data = totals,
      aes(x = .x, y = 1.02, label = paste0("n=", comma(n_total))),
      inherit.aes = FALSE,
      hjust = 0,
      size = 3.2
    ) +

    scale_y_continuous(
      labels = NULL,
      breaks = NULL,
      expand = expansion(mult = c(0, 0.12))
    ) +

    scale_fill_manual(
      breaks = sev_levels,
      values = c("Other" = "#F8766D", "Severe/Fatal" = "#00BFC4")
    ) +

    labs(
      title = paste("Severity Distribution by", gsub("_", " ", var_name)),
      x = gsub("_", " ", var_name),
      y = NULL,
      fill = "Severity"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      plot.margin = margin(5.5, 40, 5.5, 5.5)
    )
}

plots <- setNames(
  map(multi_class_vars, ~ plot_severity_by_variable(.x, data = crashes_model)),
  multi_class_vars
)

vars_to_show <- intersect(
  c("DAY_OF_WEEK","time_bucket","WEATHER_1","LIGHTING","TYPE_OF_COLLISION",
    "PRIMARY_COLL_FACTOR","PCF_VIOL_CATEGORY"),
  names(plots)
)

walk(vars_to_show, ~ print(plots[[.x]]))
```

# Correlation Check 
```{r}
# What: Correlation scan for numeric predictors (excluding IDs and zero-variance cols).
# Why: Identify redundancy safely without crashing on NA correlations.

numeric_data <- crashes_model %>%
  select(where(is.numeric)) %>%
  select(-any_of(c("CASE_ID")))

numeric_data <- numeric_data %>%
  select(where(~ !all(is.na(.x)))) %>%
  select(where(~ sd(.x, na.rm = TRUE) > 0))

if (ncol(numeric_data) >= 2) {
  cor_matrix <- cor(numeric_data, use = "pairwise.complete.obs")
  corrplot(cor_matrix, tl.cex = 0.5)

  high_corr <- c()
  moderate_corr <- c()

  for (i in 1:(ncol(cor_matrix) - 1)) {
    for (j in (i + 1):ncol(cor_matrix)) {
      r <- cor_matrix[i, j]
      if (!is.na(r)) {
        pair <- paste(colnames(cor_matrix)[i],
                      colnames(cor_matrix)[j],
                      "r =", round(r, 3))
        if (abs(r) >= 0.8) {
          high_corr <- c(high_corr, pair)
        } else if (abs(r) >= 0.5) {
          moderate_corr <- c(moderate_corr, pair)
        }
      }
    }
  }

  cat("High Correlations (|r| ≥ 0.8):\n")
  high_corr

  cat("\nModerate Correlations (0.5 ≤ |r| < 0.8):\n")
  moderate_corr

} else {
  cat("Not enough numeric columns for correlation analysis.\n")
}
```

# Baseline Model: Logistic Regression (Level-Safe)
```{r}
# What: Fit baseline logistic regression with factor levels handled safely.
# Why: Prevent predict() errors from unseen factor levels and ensure "Not Stated" is explicit.

candidate_predictors <- c(
  "weekend", "night", "raining",
  "alcohol", "ped",
  "rain_x_dark", "alcohol_x_night", "ped_x_dark",
  "DAY_OF_WEEK", "time_bucket", "WEATHER_1", "LIGHTING",
  "TYPE_OF_COLLISION", "PRIMARY_COLL_FACTOR",
  # optionally include CHP vars after screening:
  "CHP_SHIFT", "BEAT_TYPE", "CHP_BEAT_CLASS"
)

candidate_predictors <- intersect(candidate_predictors, names(crashes_model))

model_df <- crashes_model %>%
  select(SevereBinary, all_of(candidate_predictors)) %>%
  filter(!is.na(SevereBinary)) %>%
  mutate(across(where(is.character), as.factor)) %>%
  mutate(across(where(is.factor), ~ forcats::fct_explicit_na(.x, na_level = "Not Stated"))) %>%
  mutate(across(where(is.factor), ~ factor(.x, levels = union(levels(.x), "Not Stated"))))

set.seed(544)
idx <- sample.int(nrow(model_df), size = floor(0.8 * nrow(model_df)))
train_df <- model_df[idx, ]
test_df  <- model_df[-idx, ]

form <- as.formula(paste("SevereBinary ~", paste(candidate_predictors, collapse = " + ")))
glm_fit <- glm(form, data = train_df, family = binomial())

summary(glm_fit)
```

# Baseline Evaluation (Accuracy + Confusion Matrix)
```{r}
# What: Predict on test set safely with aligned factor levels.
# Why: predict.glm requires factor levels in newdata to match training levels.

prop.table(table(crashes_model$SevereBinary))

for (nm in names(glm_fit$xlevels)) {
  if (nm %in% names(test_df)) {
    test_df[[nm]] <- factor(as.character(test_df[[nm]]), levels = glm_fit$xlevels[[nm]])
  }
}

cc <- complete.cases(test_df[, all.vars(delete.response(terms(glm_fit)))])
test_df_cc <- test_df[cc, ]

test_probs <- predict(glm_fit, newdata = test_df_cc, type = "response")
test_pred <- ifelse(test_probs >= 0.5, "Severe/Fatal", "Other") %>%
  factor(levels = levels(test_df_cc$SevereBinary))

conf_mat <- table(Predicted = test_pred, Actual = test_df_cc$SevereBinary)
conf_mat

safe_div <- function(a, b) ifelse((a + b) == 0, NA_real_, a / (a + b))

tp <- conf_mat["Severe/Fatal","Severe/Fatal"]
fp <- ifelse(!is.na(conf_mat["Severe/Fatal","Other"]), conf_mat["Severe/Fatal","Other"], 0)
fn <- ifelse(!is.na(conf_mat["Other","Severe/Fatal"]), conf_mat["Other","Severe/Fatal"], 0)

precision <- safe_div(tp, tp + fp)
recall    <- safe_div(tp, tp + fn)
precision
recall

accuracy <- mean(test_pred == test_df_cc$SevereBinary)
accuracy

dropped <- nrow(test_df) - nrow(test_df_cc)
dropped
```
# Model Diagnostics
```{r model-diagnostics}
# What: Report class imbalance and complete-case loss for the baseline model evaluation.
# Why: Class notes emphasize diagnostics (imbalance, feasibility, missingness) before trusting model outputs. 

# Outcome prevalence
prev <- prop.table(table(crashes_model$SevereBinary))
prev

# Complete-case rate for test scoring
cc_rate <- mean(cc)
cc_rate

# Quick check: is complete-case loss differential by outcome?
loss_by_outcome <- data.frame(
  SevereBinary = test_df$SevereBinary,
  complete_case = cc
) %>%
  group_by(SevereBinary) %>%
  summarise(
    n = n(),
    pct_complete = mean(complete_case) * 100,
    .groups = "drop"
  )

loss_by_outcome
```