---
title: "STAT 544 - SWITRS Crash Severity Notebook"
output: html_notebook
---
# Package Chunk
```{r setup, message=FALSE, warning=FALSE}
# What: Global setup (packages, options, theme)
# Why: Reproducibility and consistent output formatting across the whole notebook.

options(stringsAsFactors = FALSE)
set.seed(544)

required_pkgs <- c(
  "dplyr", "tidyr", "stringr", "ggplot2", "scales", "purrr",
  "readr", "forcats", "corrplot", "broom"
)

installed <- rownames(installed.packages())
to_install <- setdiff(required_pkgs, installed)
if (length(to_install) > 0) install.packages(to_install)

library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(scales)
library(purrr)
library(readr)
library(forcats)
library(corrplot)
library(broom)

theme_set(theme_minimal(base_size = 12))
```

# Load data
```{r}
# What: Load the crash dataset and standardize common NA tokens at import.
# Why: Avoid treating blanks/dashes/Unknown inconsistently across different chunks.

na_tokens <- c("", " ", "  ", "   ",
               "-", " - ", "- ", " -", "-  ",
               "NA", "N/A", "NULL", "null",
               "Unknown", "UNKNOWN", "UNK")

crashes_raw <- read.csv("Crashes.csv", na.strings = na_tokens)

dim(crashes_raw)
glimpse(crashes_raw)
```

# Quick Data Audit
```{r}
# What: Basic audit (duplicates, missingness profile).
# Why: Catch data integrity issues early; document missingness before changes.

duplicates_num <- sum(duplicated(crashes_raw))
duplicates_num

missing_pct_raw <- sort(colMeans(is.na(crashes_raw)) * 100, decreasing = TRUE)
head(round(missing_pct_raw, 2), 20)
```

# Standardize Strings + "Not Stated" Handling
```{r}
# What: Clean string formatting and standardize "Not Stated" markers.
# Why: SWITRS often encodes nonresponse as blanks/dashes; we keep explicit "Not Stated"
#      as its own level for key categorical fields rather than collapsing into NA.

crashes_clean <- crashes_raw %>%
  mutate(across(where(is.character), ~ str_trim(.x))) %>%
  mutate(across(where(is.character), ~ na_if(.x, ""))) %>%
  mutate(across(where(is.character), ~ ifelse(.x %in% c("-", " - ", " -", "- "), NA, .x)))

# Helper: turn NA into explicit "Not Stated" for selected categorical variables
na_to_not_stated <- function(x) {
  x <- as.character(x)
  ifelse(is.na(x), "Not Stated", x)
}
```

# Convert Y/N/Y-Only Flags Carefully
```{r}
# What: Convert binary flags to 0/1 with consistent rules.
# Why: Some columns are Y/blank (treat blank as 0), others are Y/N/blank (blank => NA).
#      This preserves meaning and reduces accidental misclassification.

# A) Columns that are Y/blank meaning "Y else 0"
yn_blank_vars <- c("PEDESTRIAN_ACCIDENT",
                   "BICYCLE_ACCIDENT",
                   "MOTORCYCLE_ACCIDENT",
                   "TRUCK_ACCIDENT",
                   "NOT_PRIVATE_PROPERTY",
                   "ALCOHOL_INVOLVED")

yn_blank_vars <- yn_blank_vars[yn_blank_vars %in% names(crashes_clean)]

crashes_clean[yn_blank_vars] <- lapply(crashes_clean[yn_blank_vars], function(x) {
  x <- toupper(str_trim(as.character(x)))
  as.integer(!is.na(x) & x == "Y")
})

# B) Detect columns that are strictly Y/N (possibly with NAs) and map Y->1, N->0, NA stays NA
yn_cols <- names(crashes_clean)[
  sapply(crashes_clean, function(x) {
    if (is.character(x) || is.factor(x)) {
      vals <- unique(na.omit(str_trim(as.character(x))))
      length(vals) > 0 && all(vals %in% c("Y","N"))
    } else FALSE
  })
]

crashes_clean[yn_cols] <- lapply(crashes_clean[yn_cols], function(x) {
  x <- toupper(str_trim(as.character(x)))
  ifelse(is.na(x), NA,
         ifelse(x == "Y", 1,
                ifelse(x == "N", 0, NA)))
})

list(
  yn_blank_vars = yn_blank_vars,
  yn_cols = yn_cols
)
```

# Remove High-Missing Variables
```{r}
# What: Drop columns with >30% missingness (rule-based).
# Why: Matches your current approach; we document it and apply consistently.

high_miss_threshold <- 30
missing_pct <- colMeans(is.na(crashes_clean)) * 100
high_miss_vars <- names(missing_pct)[missing_pct > high_miss_threshold]

high_miss_vars
length(high_miss_vars)

crashes_reduced <- crashes_clean %>%
  select(-all_of(high_miss_vars))

dim(crashes_reduced)
```

# Drop Deprecated / Non-Analytic Variables
```{r}
# What: Drop IDs or non-analytic variables you previously agreed to remove.
# Why: Reduce leakage and clutter; keep CASE_ID only for potential merges.

vars_to_remove <- c(
  "ACCIDENT_YEAR",
  "PROC_DATE",
  "OFFICER_ID",
  "ROUTE_SUFFIX",
  "POSTMILE_PREFIX",
  "REPORTING_DISTRICT",
  "CHP_SHIFT",
  "POPULATION",
  "CALTRANS_COUNTY",
  "CALTRANS_DISTRICT",
  "LATITUDE",
  "LONGITUDE",
  "POINT_X",
  "POINT_Y",
  "COUNTY"
)

vars_to_remove <- intersect(vars_to_remove, names(crashes_reduced))

crashes_reduced <- crashes_reduced %>%
  select(-all_of(vars_to_remove))

dim(crashes_reduced)
```

# Recode SWITRS Coded Fields (Safe Mappings)
```{r}
# What: Apply SWITRS codebook label mappings for key coded fields.
# Why: Improves interpretability and supports "client-ready" EDA.

recode_maps <- list(
  DAY_OF_WEEK = c(
    "1" = "Monday","2" = "Tuesday","3" = "Wednesday","4" = "Thursday",
    "5" = "Friday","6" = "Saturday","7" = "Sunday"
  ),
  DIRECTION = c("N" = "North","E" = "East","S" = "South","W" = "West"),
  INTERSECTION = c("Y" = "Intersection","N" = "Not Intersection"),
  STATE_HWY_IND = c("Y" = "State Highway","N" = "Not State Highway"),

  WEATHER_1 = c("A" = "Clear","B" = "Cloudy","C" = "Raining","D" = "Snowing","E" = "Fog","F" = "Other","G" = "Wind"),
  WEATHER_2 = c("A" = "Clear","B" = "Cloudy","C" = "Raining","D" = "Snowing","E" = "Fog","F" = "Other","G" = "Wind"),

  LOCATION_TYPE = c("H" = "Highway","I" = "Intersection","R" = "Ramp (or Collector)"),

  TOW_AWAY = c("Y" = "Yes","N" = "No"),

  COLLISION_SEVERITY = c(
    "1" = "Fatal Injury",
    "2" = "Suspected Serious Injury / Severe Injury",
    "3" = "Suspected Minor Injury / Visible Injury",
    "4" = "Possible Injury / Complaint of Pain",
    "0" = "No Injury (PDO)"
  ),

  PRIMARY_COLL_FACTOR = c(
    "A" = "Code Violation",
    "B" = "Other Improper Driving",
    "C" = "Other Than Driver",
    "D" = "Unknown",
    "E" = "Fell Asleep"
  ),

  PCF_VIOL_CATEGORY = c(
  "01"="DUI (Alcohol/Drug)", "1"="DUI (Alcohol/Drug)",
  "02"="Impeding Traffic",   "2"="Impeding Traffic",
  "03"="Unsafe Speed",       "3"="Unsafe Speed",
  "04"="Following Too Closely","4"="Following Too Closely",
  "05"="Wrong Side of Road", "5"="Wrong Side of Road",
  "06"="Improper Passing",   "6"="Improper Passing",
  "07"="Unsafe Lane Change", "7"="Unsafe Lane Change",
  "08"="Improper Turning",   "8"="Improper Turning",
  "09"="Auto Right-of-Way",  "9"="Auto Right-of-Way",
  "10"="Pedestrian Right-of-Way",
  "11"="Pedestrian Violation",
  "12"="Traffic Signals & Signs",
  "13"="Hazardous Parking",
  "14"="Lights",
  "15"="Brakes",
  "16"="Other Equipment",
  "17"="Other Hazardous Violation",
  "18"="Other Than Driver (or Pedestrian)",
  "21"="Unsafe Starting/Backing",
  "22"="Other Improper Driving",
  "23"="Pedestrian/Other DUI (Alcohol/Drug)",
  "24"="Fell Asleep",
  "00"="Unknown", "0"="Unknown"
),

  HIT_AND_RUN = c("F"="Felony","M"="Misdemeanor","N"="Not Hit and Run"),

  TYPE_OF_COLLISION = c(
    "A" = "Head-On","B" = "Sideswipe","C" = "Rear End","D" = "Broadside",
    "E" = "Hit Object","F" = "Overturned","G" = "Vehicle/Pedestrian","H" = "Other"
  ),

  ROAD_SURFACE = c("A" = "Dry","B" = "Wet","C" = "Snowy or Icy","D" = "Slippery (Muddy/Oily/etc.)"),

  LIGHTING = c(
    "A" = "Daylight",
    "B" = "Dusk/Dawn",
    "C" = "Dark - Street Lights",
    "D" = "Dark - No Street Lights",
    "E" = "Dark - Street Lights Not Functioning"
  )
)

apply_recode <- function(df, var, map) {
  if (!var %in% names(df)) return(df)
  df %>%
    mutate("{var}" := dplyr::recode(as.character(.data[[var]]),
                                   !!!map,
                                   .default = as.character(.data[[var]])))
}

crashes_labeled <- crashes_reduced
for (v in names(recode_maps)) {
  crashes_labeled <- apply_recode(crashes_labeled, v, recode_maps[[v]])
}

# For key coded categoricals, keep "Not Stated" as explicit level
key_cat_vars <- intersect(names(recode_maps), names(crashes_labeled))
crashes_labeled <- crashes_labeled %>%
  mutate(across(all_of(key_cat_vars), ~ na_to_not_stated(.x)))
```

# Define Target (Severe/Fatal)
```{r}
# What: Create the binary target variable.
# Why: Your proposal defines severe/fatal as COLLISION_SEVERITY in {1,2}.

crashes_model <- crashes_labeled %>%
  mutate(
    COLLISION_SEVERITY = as.character(COLLISION_SEVERITY),
    SevereBinary = ifelse(COLLISION_SEVERITY %in% c("Fatal Injury",
                                                   "Suspected Serious Injury / Severe Injury"),
                          "Severe/Fatal", "Other"),
    SevereBinary = factor(SevereBinary, levels = c("Other", "Severe/Fatal"))
  )

table(crashes_model$SevereBinary, useNA = "ifany")
```

# Feature Engineering (Time + Interactions)
```{r}
# What: Create derived time features and interaction flags.
# Why: The proposal emphasizes time-of-day, weekend, and interactions like rain×dark, alcohol×night, ped×dark.

# Helper to parse COLLISION_TIME if it's HHMM numeric/character (e.g., 2359)
parse_hour <- function(x) {
  if (all(is.na(x))) return(rep(NA_integer_, length(x)))
  x <- as.character(x)
  x <- str_replace_all(x, "\\.0$", "")
  x <- str_pad(x, width = 4, side = "left", pad = "0")
  hh <- suppressWarnings(as.integer(str_sub(x, 1, 2)))
  hh[hh < 0 | hh > 23] <- NA_integer_
  hh
}

if ("COLLISION_TIME" %in% names(crashes_model)) {
  crashes_model <- crashes_model %>%
    mutate(
      hour = parse_hour(COLLISION_TIME),
      time_bucket = case_when(
        is.na(hour) ~ "Not Stated",
        hour %in% 0:5 ~ "Late Night (12–5)",
        hour %in% 6:9 ~ "AM Commute (6–9)",
        hour %in% 10:15 ~ "Midday (10–3)",
        hour %in% 16:19 ~ "PM Commute (4–7)",
        hour %in% 20:23 ~ "Evening (8–11)",
        TRUE ~ "Not Stated"
      )
    )
}

if ("DAY_OF_WEEK" %in% names(crashes_model)) {
  crashes_model <- crashes_model %>%
    mutate(
      weekend = ifelse(DAY_OF_WEEK %in% c("Saturday","Sunday"), 1,
                       ifelse(DAY_OF_WEEK == "Not Stated", NA, 0))
    )
}

# Night indicator from LIGHTING (preferred if present)
if ("LIGHTING" %in% names(crashes_model)) {
  crashes_model <- crashes_model %>%
    mutate(
      night = ifelse(LIGHTING %in% c("Dark - Street Lights",
                                    "Dark - No Street Lights",
                                    "Dark - Street Lights Not Functioning"), 1,
                     ifelse(LIGHTING == "Not Stated", NA, 0))
    )
}

# Weather rain indicator (uses WEATHER_1 if present)
if ("WEATHER_1" %in% names(crashes_model)) {
  crashes_model <- crashes_model %>%
    mutate(
      raining = ifelse(WEATHER_1 == "Raining", 1,
                       ifelse(WEATHER_1 == "Not Stated", NA, 0))
    )
}

# Pedestrian indicator: if you created PEDESTRIAN_ACCIDENT earlier, it is already 0/1
if ("PEDESTRIAN_ACCIDENT" %in% names(crashes_model)) {
  crashes_model <- crashes_model %>%
    mutate(ped = PEDESTRIAN_ACCIDENT)
}

# Alcohol indicator: if you created ALCOHOL_INVOLVED earlier, it is already 0/1
if ("ALCOHOL_INVOLVED" %in% names(crashes_model)) {
  crashes_model <- crashes_model %>%
    mutate(alcohol = ALCOHOL_INVOLVED)
}

# Interaction flags
crashes_model <- crashes_model %>%
  mutate(
    rain_x_dark = ifelse(!is.na(raining) & !is.na(night), as.integer(raining == 1 & night == 1), NA),
    alcohol_x_night = ifelse(!is.na(alcohol) & !is.na(night), as.integer(alcohol == 1 & night == 1), NA),
    ped_x_dark = ifelse(!is.na(ped) & !is.na(night), as.integer(ped == 1 & night == 1), NA)
  )

glimpse(crashes_model)
```

# EDA: Numeric Summaries + Histograms (Selective)
```{r}
# What: Summarize numeric variables; show histograms for continuous-like fields.
# Why: Quick sanity checks; not meant to be the "client story" (that comes next).

num_vars <- names(crashes_model)[sapply(crashes_model, is.numeric)]

# Frequency tables for discrete numeric variables (<= 20 unique)
for (v in num_vars) {
  u <- length(unique(na.omit(crashes_model[[v]])))
  if (u <= 20) {
    cat("\n==============================\n")
    cat("Variable:", v, "\n")
    print(table(crashes_model[[v]], useNA = "ifany"))
  }
}

# Histograms for more continuous numeric variables
for (v in num_vars) {
  u <- length(unique(na.omit(crashes_model[[v]])))
  if (u > 20) {
    p <- ggplot(crashes_model, aes(x = .data[[v]])) +
      geom_histogram(bins = 25, color = "white") +
      labs(title = paste("Histogram:", v), x = v, y = "Count")
    print(p)
  }
}
```

# EDA: Decision-Oriented Severity Rate Plots (Ordered, Horizontal)
```{r}
# What: Plot Severe/Fatal proportion by categorical variables.
# Why: Client-ready view of *rates* (proportions), ordered by Severe/Fatal share for interpretability.

# Identify multi-class categorical variables (exclude IDs and target)
exclude_vars <- intersect(c("CASE_ID","SevereBinary","COLLISION_SEVERITY"), names(crashes_model))

multi_class_vars <- names(crashes_model)[
  sapply(crashes_model, function(x) {
    (is.character(x) || is.factor(x)) && length(unique(na.omit(x))) > 2
  })
]
multi_class_vars <- setdiff(multi_class_vars, exclude_vars)

# Convert to factor (preserve labels as-is)
crashes_model[multi_class_vars] <- lapply(crashes_model[multi_class_vars], as.factor)

plot_severity_by_variable <- function(df, var_name, min_n = 200) {
  # Compute severe share by category
  tab <- df %>%
    filter(!is.na(.data[[var_name]]), !is.na(SevereBinary)) %>%
    group_by(.data[[var_name]]) %>%
    summarise(
      n = n(),
      severe_rate = mean(SevereBinary == "Severe/Fatal"),
      .groups = "drop"
    ) %>%
    mutate(keep = n >= min_n)

  # Keep all categories but visually warn small-n via ordering + optional filtering toggle
  # Here we keep all, but you can change to filter(keep) if you want.
  levels_order <- tab %>% arrange(severe_rate) %>% pull(1)

  df2 <- df %>%
    mutate("{var_name}" := factor(.data[[var_name]], levels = levels_order))

  ggplot(df2, aes(y = .data[[var_name]], fill = SevereBinary)) +
    geom_bar(position = "fill") +
    scale_x_continuous(labels = percent_format()) +
    labs(
      title = paste("Severe/Fatal Share by", str_replace_all(var_name, "_", " ")),
      x = "Proportion of collisions",
      y = str_replace_all(var_name, "_", " "),
      fill = "Severity"
    ) +
    theme(
      axis.text.y = element_text(size = 9),
      plot.title = element_text(face = "bold")
    )
}

# Generate plots (may be many; printing all can be long)
plots <- map(multi_class_vars, ~ plot_severity_by_variable(crashes_model, .x, min_n = 200))
names(plots) <- multi_class_vars

# Print a subset you care about most (edit this list)
vars_to_show <- intersect(
  c("DAY_OF_WEEK","time_bucket","WEATHER_1","LIGHTING","TYPE_OF_COLLISION","PRIMARY_COLL_FACTOR","PCF_VIOL_CATEGORY"),
  names(plots)
)

plots[vars_to_show]
```

# Correlation Check 
```{r}
# What: Correlation scan for numeric predictors (excluding IDs and zero-variance cols).
# Why: Identify redundancy safely without crashing on NA correlations.

numeric_data <- crashes_model %>%
  select(where(is.numeric)) %>%
  select(-any_of(c("CASE_ID")))

# Remove columns with zero variance or all NA
numeric_data <- numeric_data %>%
  select(where(~ !all(is.na(.x)))) %>%
  select(where(~ sd(.x, na.rm = TRUE) > 0))

if (ncol(numeric_data) >= 2) {
  
  cor_matrix <- cor(numeric_data, use = "pairwise.complete.obs")
  
  corrplot(cor_matrix, tl.cex = 0.7)
  
  high_corr <- c()
  moderate_corr <- c()
  
  for (i in 1:(ncol(cor_matrix) - 1)) {
    for (j in (i + 1):ncol(cor_matrix)) {
      
      r <- cor_matrix[i, j]
      
      # Skip NA correlations safely
      if (!is.na(r)) {
        
        pair <- paste(colnames(cor_matrix)[i],
                      colnames(cor_matrix)[j],
                      "r =", round(r, 3))
        
        if (abs(r) >= 0.8) {
          high_corr <- c(high_corr, pair)
        } else if (abs(r) >= 0.5) {
          moderate_corr <- c(moderate_corr, pair)
        }
      }
    }
  }
  
  cat("High Correlations (|r| ≥ 0.8):\n")
  high_corr
  
  cat("\nModerate Correlations (0.5 ≤ |r| < 0.8):\n")
  moderate_corr
  
} else {
  cat("Not enough numeric columns for correlation analysis.\n")
}
```

# Baseline Model: Logistic Regression (Level-Safe)
```{r}
# What: Fit baseline logistic regression with factor levels guaranteed (including "Not Stated").
# Why: Prevent predict() errors from unseen factor levels.

candidate_predictors <- c(
  "weekend", "night", "raining",
  "alcohol", "ped",
  "rain_x_dark", "alcohol_x_night", "ped_x_dark",
  "DAY_OF_WEEK", "time_bucket", "WEATHER_1", "LIGHTING",
  "TYPE_OF_COLLISION", "PRIMARY_COLL_FACTOR"
)

candidate_predictors <- intersect(candidate_predictors, names(crashes_model))

model_df <- crashes_model %>%
  select(SevereBinary, all_of(candidate_predictors)) %>%
  filter(!is.na(SevereBinary)) %>%
  mutate(across(where(is.character), as.factor)) %>%
  mutate(across(where(is.factor), ~ forcats::fct_explicit_na(.x, na_level = "Not Stated")))

# Force "Not Stated" to exist in the *levels attribute* for every factor (even if absent in the data)
model_df <- model_df %>%
  mutate(across(where(is.factor), ~ factor(.x, levels = union(levels(.x), "Not Stated"))))

# Split
set.seed(544)
idx <- sample.int(nrow(model_df), size = floor(0.8 * nrow(model_df)))
train_df <- model_df[idx, ]
test_df  <- model_df[-idx, ]

# Fit
form <- as.formula(paste("SevereBinary ~", paste(candidate_predictors, collapse = " + ")))
glm_fit <- glm(form, data = train_df, family = binomial())

# Quick check (should include "Not Stated" now if WEATHER_1 exists)
if ("WEATHER_1" %in% names(glm_fit$xlevels)) glm_fit$xlevels$WEATHER_1
```

# Baseline Evaluation (Accuracy + Confusion Matrix)
```{r}
# What: Predict on test set safely.
# Why: predict.glm requires that factor levels in newdata match training xlevels exactly.

# Align factor levels in test_df to what the model was trained on
for (nm in names(glm_fit$xlevels)) {
  if (nm %in% names(test_df)) {
    test_df[[nm]] <- factor(as.character(test_df[[nm]]), levels = glm_fit$xlevels[[nm]])
  }
}

# If anything was truly unseen (not even "Not Stated"), it becomes NA after alignment.
# predict() will fail if NA remains in predictors; so we score only complete cases.
cc <- complete.cases(test_df[, all.vars(delete.response(terms(glm_fit)))])
test_df_cc <- test_df[cc, ]

test_probs <- predict(glm_fit, newdata = test_df_cc, type = "response")
test_pred <- ifelse(test_probs >= 0.5, "Severe/Fatal", "Other") %>%
  factor(levels = levels(test_df_cc$SevereBinary))

conf_mat <- table(Predicted = test_pred, Actual = test_df_cc$SevereBinary)
conf_mat

accuracy <- mean(test_pred == test_df_cc$SevereBinary)
accuracy

# How many rows were dropped due to remaining NA predictors after alignment?
dropped <- nrow(test_df) - nrow(test_df_cc)
dropped
```
